---
title: "Heart failure data analysis in R"
author: "Jyotishka Datta"
date: "April 9, 2019"
output: 
  html_document:
      toc: true
      number_sections: true
fontsize: 12pt
geometry: margin=1in
urlcolor: blue
---

```{r setup, include=FALSE}
library(knitr)

opts_chunk$set(comment = NA,
               prompt = FALSE,
               cache = TRUE,
               warning = FALSE,
               message = FALSE,
               fig.asp = 0.5)
library(summarytools)
st_options(plain.ascii = FALSE,      
           style = "rmarkdown",      # This too
           footnote = NA,             # Avoids footnotes which would clutter the results
           subtitle.emphasis = FALSE  # This is a setting to experiment with - according to 
           )                                     # the theme used, it might improve the headings'
                                      # layout
```

# Background 

## Data reading 

```{r}
rm(list=ls())
setwd("C:/Users/jd033/OneDrive/Documents/R/Nursing")
heart <- read.csv("Heart-failure-project-data.csv",sep=",",header=T)
str(heart)
```

Before we analyze the data in R, we need to convert the data from a long format to a wide format. We also change the name of the first column from `patient ID` to `subject`.

This is done as follows: 

```{r}
library(tidyverse)
library(reshape2)
colnames(heart)[1] <- "subject"
heart.new <- heart[,c(1:3,5)]
heart.wide <-  dcast(heart.new, subject + Group ~ Question, sum, value.var="Score")
dim(heart.wide)
```

# Basic Descriptive Statistics

## Descriptive Statistics for the "Pre" group

```{r, results = 'asis'}
library(summarytools)
st_options(dfSummary.na.col = FALSE)

attach(heart.wide)
heart.wide.pre = heart.wide[Group == "Pre",]
dfSummary(heart.wide.pre[,-1], plain.ascii = FALSE, style = "grid",           graph.magnif = 0.75, valid.col = FALSE, tmp.img.dir = "/tmp")
```

## Descriptive Statistics for the "Post" group

```{r, results = 'asis'}
attach(heart.wide)

heart.wide.post = heart.wide[Group == "Post",]
dfSummary(heart.wide.post[,-1], plain.ascii = FALSE, style = "grid",           graph.magnif = 0.75, valid.col = FALSE, tmp.img.dir = "/tmp")
```


# Analysis of Variance 

## Calculating normalized score 

One of the variable `Forget Meds` is coded in reverse order. We recode that here using the recode function in R. 

```{r}
cat("Before re-coding \n")
(heart.wide$`Forget Meds`)
heart.wide$`Forget Meds`<- recode(heart.wide$`Forget Meds`, `1` = 4, `2` = 3, `3`= 2, `4` = 1, `0` = 0)
cat("After re-coding \n")
(heart.wide$`Forget Meds`)
heart.wide[is.na(heart.wide)] <- 0
```

**Then we apply the following formuale from page 22 of your paper:**

-  Maintenance (reverse code #8) :( sum of section A items -10) x 3.333
-  Management (score only if patient has answered yes to trouble breathing or ankle swelling): (sum of section B items -4) x 5
- Confidence: (sum of section C items -6) x 5.56

### Pre-post comparison for Section A

The next step is calculating the normalized score for all section A items after adding them. We will follow the formulae given above and then test whether the normalized scores were different between the **pre** and **post** groups using a one-way ANOVA. The R codes and the ANOVA tables are given below:


```{R}
heart.wide$sec.A.score = with(heart.wide, Weigh+Swelling+Sick+Activity+Appointment+`Low Salt Diet`+
       Exercise+`Forget Meds`+`Ask for  low Salt`+`Pill box`)
heart.wide$maintenance = (heart.wide$sec.A.score - 10)*3.33

ggplot(heart.wide, aes(Group, maintenance))+ geom_boxplot() + geom_point() + labs(title = "Maintenance")

fit <- lm(maintenance ~ Group, data = heart.wide)
summary(fit)

# Max.sec.A.score = 4*10
# Min.sec.A.score = 0
# (heart.wide$normalized.sec.A.score = (heart.wide$sec.A.score - Min.sec.A.score)/(Max.sec.A.score - Min.sec.A.score)*100)
```

**The P-value for the test of difference in mean normalized section A score between pre and post groups is 0.000491, which implies that there was a significant improvement of the test score in the post group.**

### Pre-post comparison for Section B 

```{R}
heart.wide$sec.B.score = with(heart.wide, Symptoms + Recognize + `Reduce Na`+`Reduce H2O`
                              +`Inc Med`+`Contact NP`+ `Remedy Eval`)
heart.wide$management = (heart.wide$sec.B.score-4)*5

p2 <- ggplot(heart.wide, aes(Group, management))
p2 + geom_boxplot() + geom_point()+ labs(title = "Management")

fit <- lm(management ~ Group, data = heart.wide)
summary(fit)

## alternative normalization 
# Max.sec.B.score = max(heart.wide$sec.B.score)
# Min.sec.B.score = min(heart.wide$sec.B.score)
# (heart.wide$normalized.sec.B.score = (heart.wide$sec.B.score - Min.sec.B.score)/(Max.sec.B.score - Min.sec.B.score))
```

**The P-value for the test of difference in mean normalized section B score between pre and post groups is $6.3\times 10^{-9}$, which implies that there was a significant improvement of the test score in the post group.**

### Pre-post comparison for Section C 

```{R}
heart.wide$sec.C.score = with(heart.wide, `Free fromSx` + `Follow Tx`+ `Eval Sx` + `Rec change`+`Act on Sx`+`Eval Tx`)
heart.wide$confidence = (heart.wide$sec.C.score - 6)*5.56

p3 <- ggplot(heart.wide, aes(Group, confidence))
p3 + geom_boxplot() + geom_point()+ labs(title = "Confidence")

fit <- lm(confidence ~ Group, data = heart.wide)
summary(fit)
```

**The P-value for the test of difference in mean normalized section C score (confidence) between pre and post groups is $1.92\times 10^{-5}$, which implies that there was a significant improvement of the test score in the post group.**




