---
title: "Deviance Explained"
subtitle: "Logistic Regression"
author: "Jyotishka Datta"
institute: "Virginia Tech"
date: "2016/12/12 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = T, warning = F, message = F)

```

```{r include, include=FALSE, cache = TRUE}
set.seed(12345)
library(ggplot2)
library(dplyr)
library(ISLR)
```

## Example 

```{r, eval = F}
options(contrasts = c("contr.treatment", "contr.poly"))
ldose <- rep(0:5, 2)
numdead <- c(1, 4, 9, 13, 18, 20, 0, 2, 6, 10, 12, 16)
sex <- factor(rep(c("M", "F"), c(6, 6)))
SF <- cbind(numdead, numalive = 20 - numdead)

budworm.lg <- glm(SF ~ sex*ldose, family = binomial)
summary(budworm.lg, cor = F)
```

---
## R output

```{r, echo = F}
options(contrasts = c("contr.treatment", "contr.poly"))
ldose <- rep(0:5, 2)
numdead <- c(1, 4, 9, 13, 18, 20, 0, 2, 6, 10, 12, 16)
sex <- factor(rep(c("M", "F"), c(6, 6)))
SF <- cbind(numdead, numalive = 20 - numdead)

budworm.lg <- glm(SF ~ sex*ldose, family = binomial)
summary(budworm.lg, cor = F)
```


---
## Deviances are stored in `summary` object

```{r}
summaryLog <- summary(budworm.lg)
# 'Residual deviance' is the deviance; 
# 'Null deviance' is the null deviance

(summaryLog$deviance)
(summaryLog$null.deviance)
```

-  Recall: Intuitively, if the null deviance is much larger than the fitted deviance, the predictors are important, if they are similar the predictors are noise. 


---
## Fit the null model to check

```{r}
# Null model - only intercept
budworm.null <- glm(SF ~ 1, family = binomial)
summary(budworm.null, cor = F)
```

---
## Calculate $R^2$

Recall:

$$
		R^2 = 1 - \frac{D}{D_0} = 1 - \frac{\text{deviance of fitted model to saturated}}{\text{deviance of null model to saturated}}.
$$
-  It is a quantity between $0$ and $1$. 

-  If the fit is perfect, then $D=0$ and $R^2 = 1$, but if the predictors have no information about $Y$, then $D = D_0$, and $R^2 = 0$. 
-  For the example at hand, 

```{r}
(R_sq = 1 - summaryLog$deviance / summaryLog$null.deviance)
```
-  Is the high value of $R^2$ surprising or not? 
---
## Smarket 

```{r, echo = T}
library(ISLR)
names(Smarket)
dim(Smarket)
```

---
## Smarket data 

```{r}
glm.fits=glm(Direction~Lag1+Lag2+Lag3+Lag4+Lag5+Volume,data=Smarket,family=binomial)
summary(glm.fits)
```


---
## Smarket data 

- Calculate the $R^2$ for this fitted model. 

```{r, echo = T}
summarySmarket <- summary(glm.fits)
(R_sq = 1 - (summarySmarket$deviance / summarySmarket$null.deviance))
```
-  Now, what does this say about the quality of fit? 

-  How do you connect this with the $p$-values? 

