---
title: "Kernel Density Estimation using R"
author: "Jyotishka Datta (<jyotishka@vt.edu>) <br> Department of Statistics, Virginia Tech"
output: 
  ioslides_presentation:
    smaller: yes
logo: ../vt.png
transition: faster
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Histograms 

-  Histograms provide an idea about the data distribution. Defined as:

$$
\hat{f}(x) = \frac{ \text{No. of observations within } h \text{ of } x }{n \times 2h} \\
$$
or, 

$$
	\hat{f}(x) = \frac{1}{2nh} \sum_{i=1}^{n} I(|{X_i - x}| \le h).
$$

-  Narrow intervals (smaller h): histogram has a choppy appearance, large variance.
 - Large intervals (larger h): histogram may lose the local feature, large bias.

## R `hist()` {.smaller}

- In base R's `hist()` method, you can use the `break` argument:

-  `breaks` can be one of:

1. a vector giving the breakpoints between histogram cells,
2. a function to compute the vector of breakpoints,
3. a single number giving the number of cells for the histogram,
4. a character string naming an algorithm to compute the number of cells (see ‘Details’),
5. a function to compute the number of cells.


## Sampling distribution 

- By default, number of breaks is calculated by using Sturges' formula. 

```{r,echo = TRUE}
data(mtcars)
hist(mtcars$mpg)
```

## Colored Histogram with Different Number of Bins

```{r,echo = TRUE}
hist(mtcars$mpg, breaks=12, col="red")
```

## Different h 

```{r, echo = T, eval = F}
par(mfrow = c(2,2))
hist(mtcars$mpg, breaks=2, col="red")
hist(mtcars$mpg, breaks=8, col="red")
hist(mtcars$mpg, breaks=12, col="red")
hist(mtcars$mpg, breaks=20, col="red")
par(mfrow = c(1,1))

```

## Different h 

```{r, echo = F, eval = T}
par(mfrow = c(2,2))
hist(mtcars$mpg, breaks=2, col="red")
hist(mtcars$mpg, breaks=8, col="red")
hist(mtcars$mpg, breaks=12, col="red")
hist(mtcars$mpg, breaks=20, col="red")
par(mfrow = c(1,1))

```

## Kernel Density Estimation

-  Drawback of histogram: density estimation is piecewise constant and thus unsmooth.
-  Main idea of kernel density estimation: instead of counting number of observations within $h$ of $x$, we take a certain weighted average of data points near $x$ to estimate $f(x)$.
  
-  Kernel Density Estimate:

$$
	\hat{f}(x) = \frac{1}{2nh} \sum_{i=1}^{n} K(|{X_i - x}| \le h).
$$
where $K(\cdot)$ is a Kernel function. 

- Choices: $h$ bandwidth and $K()$ the kernel.
	

## Kernel Density Plot

```{r,echo = TRUE}
d <- density(mtcars$mpg) # returns the density data 
plot(d) # plots the results
```

## Kernel Density Plot with 1/3rd bandwith 

```{r,echo = TRUE}
d <- density(mtcars$mpg,adjust = 1/3) # returns the density data 
plot(d) # plots the results
```

## Kernel Density Plot with 2x bandwith 

```{r,echo = TRUE}
d <- density(mtcars$mpg,adjust = 2) # returns the density data 
plot(d) # plots the results
```

## Kernel Density Plot with a rectangular kernel

```{r,echo = TRUE}
d <- density(mtcars$mpg, kernel = "rectangular") # returns the density data 
plot(d) # plots the results
```

## Kernel Density Plot with the Epanechnikov kernel

```{r,echo = TRUE}
d <- density(mtcars$mpg, kernel = "epanechnikov") # returns the density data 
plot(d) # plots the results
```

## Faithful Geyser 

```{r, echo = T}
#take a look at the data set (the first 5 observations)
faithful[1:5,]
duration = faithful$eruptions
n=length(duration)
#estimate the density using histogram and kernel estimation
```

## Different Kernels 

```{r, fig.asp =0.8}
par(mfrow=c(2,2))
#histogram
out=hist(duration, nclass=15, prob=TRUE, main="Histogram, h=0.2")
z = (out$breaks[-1] + out$breaks[-19])/2
lines(out$density~z, col="blue")
#kernel density estimation
d1 = density(duration, kernel="gaussian", bw=0.2)
plot(d1, main="Gaussian kernel, h=0.2")
d2=density(duration, kernel="gaussian", bw=0.05)
plot(d2, main="Gaussian kernel, h=0.05")
d3=density(duration, kernel="triangular", bw=0.2)
plot(d3,main="Triangular kernel, h=0.2")
par(mfrow=c(1,1))
```


## We can also use ggplot

```{r, echo = T, fig.asp =0.4}
library(ggplot2)
ggplot(data = faithful, aes(x = eruptions)) + geom_density()
```

## Different bandwidth {.smaller}

```{r, echo = T, fig.asp =0.4}
library(patchwork)
p1 <- ggplot(data = faithful, aes(x = eruptions)) + geom_density(adjust = 1/10) +
  labs(title = "adjust = 1/10")+theme_bw()
p2 <- ggplot(data = faithful, aes(x = eruptions)) + geom_density(adjust = 1/2) +
  labs(title = "adjust = 1/2")+theme_bw()
p1+p2
```


## Different Kernels {.smaller}

```{r, echo = T, fig.asp =0.4}
p1 <- ggplot(data = faithful, aes(x = eruptions)) + geom_density(kernel = "gaussian") +
  labs(title = "Gaussian")+theme_bw()
p2 <- ggplot(data = faithful, aes(x = eruptions)) + geom_density(kernel = "rectangular") +
  labs(title = "Rectangular")+theme_bw()
p1+p2
```
