---
title: "Cumulative Sum"
author: "Jyotishka Datta"
date: "October 12, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Cumulative Sum

We generate $n = 1000$ random numbers 
$X_1, \ldots, X_n \sim N(\mu, \sigma)$ 
from a Normal distribution with mean $\mu = 0.033$ and standard deviation $\sigma = 5.19$ (These are values from some simulation experiments done before). 

We want to see how the cumulative sum $S_k = \sum_{i=1}^{k} X_i$ behaves. Theoretically, $S_k \sim N(k\mu, \sqrt{k}\sigma)$. 
We also want to normalize the cumulative summation data to make it look more normal. We can do the following transformation:
$$
\frac{S_k - k\mu}{\sqrt{k}\sigma} \sim N(\mu, \sigma), \; k = 1, \ldots, n. 
$$

The outputs are below. As expected, the normalization makes the series staionary but reduces the variance almost to zero as $k$ grows big. 

```{r, fig.align='center'}
set.seed(222)
mu = 0.033
sigma = 5.19
x = rnorm(1000, mu, sigma)
par(mfrow=c(1,1))
plot(cumsum(x),type="l",col="red")
lines(x,col="blue")
z = cumsum(x)
z_std = (cumsum(x)-mu*seq(1,length(x)))/(sqrt(seq(1,length(x)))*sigma)
lines(z_std, lwd= 2, col="magenta")
legend("topleft",c("N(mu,sigma)", "Cum-sum", "Normalized Cum-sum"), col=c("red","blue","magenta"),lty = c(1,1,1))
```

