---
title: Data Visualization
date: "`r Sys.Date()`"
output: 
  html_document:
      css: css/lesson.css
      toc: true
      number_sections: true
      toc_float: true
  pdf_document:
      toc: true
      number_sections: false
fontsize: 10pt
geometry: margin=1in
urlcolor: blue
---

> ## Learning Objectives {.objectives}
>
> * Grammar of graphics concepts (geoms, aesthetics)
> * Advanced plots (scales, facets, themes)
> * Writing images (and other things) to file

# Basic plots in R

When we are working with large sets of numbers it can be useful to display that information graphically. R has a number of built-in tools for basic graph types such as hisotgrams, scatter plots, bar charts, boxplots and much [more](http://www.statmethods.net/graphs/). We'll test a few of these out here on the `cats` data.


```{r import data}
cats <- read.csv("data/herding-cats.csv")
```

## Scatterplot
Let's start with a **scatterplot**. A scatter plot provides a graphical view of the relationship between two sets of numbers.

Let's make a scatterplot of birth weight by mother's age.

```{r scatter-plot1, fig.align='center'}
plot(x = cats$age, y = cats$weight)
```

Each point represents a row in our dataset. The value on the x-axis is the mother's age and the values on the y-axis correspond to the birth weight for the infant. For any plot you can customize many features of your graphs (fonts, colors, axes, titles) through [graphic options](http://www.statmethods.net/advgraphs/parameters.html)

# Advanced figures (`ggplot2`)

More recently, R users have moved away from base graphic options and towards a plotting package called `ggplot2` that adds a lot of functionality to the basic plots seen above. The syntax is different but itb