---
title: "Datasets: Sparse Regression or Classification"
author: "JD"
date: "`r Sys.Date()`"
output: 
  html_document:
      toc: true
      number_sections: false
      toc_float: true
fontsize: 11pt
geometry: margin=1in
urlcolor: blue
theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Install HDRM package (source of data)

**Warning: the installation of packages and libraries will take some time when you run this for the first time ever. Then you can safely change the flag so that it does not install the package every time you run the R codes.**

```{r, echo = T, cache = T, message=FALSE, warning=FALSE}

installed_flag = T

if(installed_flag == F){
  if(!require(remotes)){
  install.packages("remotes", dependencies = TRUE, repos = 'http://cran.rstudio.com')
 }
remotes::install_github("pbreheny/hdrm")
}
```


## Sparse Regression 

### Comparing penalized methods for High-dimensional Regression

We would like to compare the performance of several modern regression methods including penalized regression methods such as LASSO and Elastic Net regression with 10-fold cross-validation. 

### Mammalian Eye Data

We will illustrate the methods using a high-dimensional data-set that consists of gene expression in the mammalian eye from (Scheetz, 2006). This data-set has 120 observations for 18,976 variables (predictors): a typical $p \gg n$ data-set where the usual regression methods do not work because the $X^TX$ matrix is not invertible anymore. 

Also, in such cases, it is natural to ask: which of these many variables ($X_j$'s) are important for predicting $y$?

For the original analysis, microarrays were used to measure levels of RNA expression in the isolated eye tissues extracted from 120 laboratory inbred rats from F2 generation. 18,976 probes were detected to be considered as mammalian eye. In the data, we treat gene Trim32 as the outcome(y) since it is known to be linked with a genetic disorder called BBS and remaining gene expression measurements are x.

```{r, echo = T, cache = T, message=FALSE, warning=FALSE}
library(hdrm)
downloadData(Scheetz2006)
attachData(Scheetz2006)

n = nrow(X)
# Split data into train and test sets
train_rows <- sample(1:n, n/2)
X.train <- X[train_rows, ]
X.test <- X[-train_rows, ]

y.train <- y[train_rows]
y.test <- y[-train_rows]

dim(X.train)
dim(X.test)

length(y.train)

length(y.test)
```


## Sparse Classification 

### Comparing penalized methods for High-dimensional Classification

We would like to compare the performance of several modern regression methods including penalized regression methods such as LASSO and Elastic Net regression with 10-fold cross-validation. 

### Golub 1999 Data 

This data set consists of 47 patients with acute lymphoblastic leukemia (ALL) and 25 patients with acute myeloid leukemia (AML). Each of the 72 patients had bone marrow samples obtained at the time of diagnosis. Gene expression measurements for the samples were taken using Affymetrix Hgu6800 chips, resulting in 7129 measurements per patient.

Of the two diseases, AML has a considerably worse prognosis: only 26% survive at least 5 years following diagnosis, compared to 68% for ALL.



```{r, echo = T, cache = T, message=FALSE, warning=FALSE}

library(hdrm)
downloadData(Golub1999)
attachData(Golub1999)

n = nrow(X)
# Split data into train and test sets
train_rows <- sample(1:n, n/2)
X.train <- X[train_rows, ]
X.test <- X[-train_rows, ]

y.train <- y[train_rows]
y.test <- y[-train_rows]

dim(X.train)
dim(X.test)

length(y.train)

length(y.test)
```


