---
title: "Sampling Lab"
author: "Jyotishka Datta"
date: "February 24, 2017"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Source 

Both the examples were taken from Jason Goldstick's R Lab Notes, available at [http://dept.stat.lsa.umich.edu/~jasoneg/Stat406/index.html](http://dept.stat.lsa.umich.edu/~jasoneg/Stat406/index.html). 


## Importance Sampling 

Suppose you want to evaluate the integral: 

$$
\int_a^b h(x) dx = \int_{0}^{10} \exp(-2 | x- 5| ) dx 
$$
The simplest way is to generate $U_i \sim \text{Unif}(0,10)$ and look at the sample mean $(b-a) \times h(U_i)$. 

## R codes 

```{r, echo = T}
set.seed(123)
U <-  runif(1e4, 0, 10)
Y <- 10* exp(-2*abs(U - 5))
(I <- mean(Y))
(Error <- var(Y))
```

```{r, echo = TRUE}
h <- function(u) exp(-2*abs(u - 5))
integrate(h, 0,10)$val
```

## Improving the accuracy 

1.  This method has a high variance, or, estimation error. 
2.  The function h in this case is peaked at 5, and decays quickly elsewhere. 
3.  Under the $U[0,10]$ density, many of the points are contributing very little to this  expectation.
4. Something more like a gaussian function with a peak at 5 and small variance, say, 1, would provide greater precision.

## Importance Sampling 

Write the same integral as an expectation of a function under Normal distribution: 

$$
I = 10 \int_{0}^{10} 10 \exp(-2 | x- 5| ) \frac{\frac{1}{10}}{\phi(x-5)} \phi(x-5)dx \\
\text{where } \phi(x-5) = \frac{1}{\sqrt{2\pi}} e^{-(x-5)^2/2} \\
= E(h(X)w(X)), \text{ for } X \sim N(5,1)
$$
Q. What are $h(x)$ and $w(x)$ in the above example? 

## Importance Sampling {.smaller}

$$
\begin{align}
I & = 10 \int_{0}^{10} 10 \exp(-2 | x- 5| ) \frac{\frac{1}{10}}{\phi(x-5)} \phi(x-5)dx \\
& \text{ where }\phi(x-5) = \frac{1}{\sqrt{2\pi}} e^{-(x-5)^2/2} \; (\text{Normal density}) 
\end{align}
$$
Hence, we can write the integral as 

$$
\begin{align}
I & = E(h(x)w(x)), \text{ for } X \sim N(5,1) \\
& \text{ Here, } h(x) = 10 \exp(-2 | x- 5| ) \\
& \text{ and } w(x) = \frac{\frac{1}{10}}{\phi(x-5)} = \frac{\texttt{dunif}(x, 0, 10)}{\texttt{dnorm}(x, mean=5, sd=1)}
\end{align}
$$ 

## R code 

```{r, echo = TRUE}
w <- function(x) dunif(x, 0, 10)/dnorm(x, mean=5, sd=1)
f <- function(x) 10*exp(-2*abs(x-5))
X=rnorm(1e5,mean=5,sd=1)
Y=w(X)*f(X)

(I <- mean(Y))
(Error <- var(Y))
```

Same integral but the new error is 1/10th of the previous error. 

## Lab Exercise 

The standard double exponential or Laplace density is given by: 

$$
p(x) = \frac{1}{2} e^{-\lvert x \rvert}
$$

Calculate the second moment $E(X^2)$ for Laplace using the Normal(0,4) density as the proposal. In other words, calculate the integral: 

$$
\int_{-\infty}^{\infty}  \frac{1}{2} x^2 e^{-\lvert x \rvert} dx 
$$

Report both your estimate and the variance of your estimate. 

Could you also calculate the true value? 

## Solution 

```{r, echo = TRUE}
X <- rnorm(1e5, sd=4)
Y <- (X^2) * .5 * exp(-abs(X))/dnorm(X, sd=4)
(I <- mean(Y))
(Error <- var(Y))
```
 
Think: What happens if we use N(0,1) instead? Or a Cauchy instead?

## True Value
The true value of the integral is 2. You can find that by integrating by hand, or, evaluating it in R. 

```{r, echo = TRUE}
h <- function(x) 0.5*x^2*exp(-abs(x))
integrate(h,-Inf,Inf)$val
```

