---
title: "Importance Sampling"
subtitle: "Examples"
author: "Jyotishka Datta"
date: "February 8, 2019"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Source 

Both the examples were taken from Jason Goldstick's R Lab Notes, available at [http://dept.stat.lsa.umich.edu/~jasoneg/Stat406/index.html](http://dept.stat.lsa.umich.edu/~jasoneg/Stat406/index.html). 

## Importance Sampling : Example 1

Suppose you want to evaluate the integral: 

$$
\int_a^b h(x) dx = \int_{0}^{2\pi} \sin(x \cos(x)) dx 
$$
The simplest way is to generate $U_i \sim \text{Unif}(0,2\pi)$ and look at the sample mean $(b-a) \times h(U_i)$. 

## Importance Sampling 

$$
\begin{align*}
\int_a^b h(x) dx & = \int_{0}^{2\pi} \sin(x \cos(x)) dx \\
& = \int_{0}^{2\pi} 2\pi \sin(x \cos(x)) \times \frac{1}{2\pi} dx \\
& = E \left[ 2\pi\sin(Ucos(U)) \right]; \; U \sim \text{Unif}(0,2\pi)
\end{align*}
$$

## R codes 

```{r, echo = T}
set.seed(123)
U <-  runif(1e4, 0, 2*pi)
Y <- 2*pi*sin(U*cos(U))
(I <- mean(Y))
```

```{r, echo = TRUE}
h <- function(u) sin(u*cos(u))
integrate(h, 0,2*pi)$val
```
- They are very close ! 

## Importance Sampling : Example 2

Suppose you want to evaluate the integral: 

$$
\int_a^b h(x) dx = \int_{0}^{10} \exp(-2 | x- 5| ) dx 
$$
The simplest way is to generate $U_i \sim \text{Unif}(0,10)$ and look at the sample mean $(b-a) \times h(U_i)$. 

## Importance Sampling 

$$
\begin{align*}
\int_a^b h(x) dx & = \int_{0}^{10} \exp(-2 | x- 5| ) dx \\
& = \int_{0}^{10} 10 \exp(-2 |x- 5| ) \frac{1}{10} dx \\
& = E (10\exp(-2 \lvert U - 5 \rvert )); \; U \sim \text{Unif}(0,10)
\end{align*}
$$

## R codes 

```{r, echo = T}
set.seed(123)
U <-  runif(1e4, 0, 10)
Y <- 10* exp(-2*abs(U - 5))
(I <- mean(Y))
```

```{r, echo = TRUE}
h <- function(u) exp(-2*abs(u - 5))
integrate(h, 0,10)$val
```
- They are close ! 

## Error 

-  How do you estimate the error ? 
-  Monte Carlo Error: use the variance of generated samples. 

```{r, echo = T}
(Error <- var(Y))
```

-  Now this is quite large ! 


## Improving the accuracy {.build}

1.  This method has a high variance, or, estimation error. 
2.  The function h in this case is peaked at 5, and decays quickly elsewhere. 
3.  Under the $U[0,10]$ density, many of the points are contributing very little to this  expectation.
4. Something more like a gaussian function with a peak at 5 and small variance, (e.g. $\sigma^2 = 1$) would provide greater precision.

## Better 

```{R, echo = F}
par(mfrow=c(1,2))
hist(runif(1000,0,10), breaks = 30, col=rgb(1,0,0,0.5), freq=F, main = "Uniform Proposal")
curve(expr=1/10*exp(-2*abs(x-5)),from = 0, to = 10, lwd=2, add= T)

hist(rnorm(1000,5,1), breaks = 30, col=rgb(1,0,0,0.5), freq=F, main = "Normal Proposal")
curve(expr=dnorm(x,5,1)*exp(-2*abs(x-5)),from = 0, to = 10, lwd=2, add= T)
# curve(10*dunif(x,0,10),col="blue",add=T)
# curve(dnorm(x,5,1),add=T)
```


## Importance Sampling 

Write the same integral as an expectation of a function under Normal distribution: 

$$
\begin{align*}
I & = \int_{0}^{10} 10 \exp(-2 | x- 5| ) \frac{\frac{1}{10}}{\phi(x-5)} \phi(x-5) dx \\
& = \int_{\mathbb{R}} 10 \exp(-2 | x- 5| ) \frac{\frac{1}{10} I(0 \le x \le 10)}{\phi(x-5)} \phi(x-5) dx \\
& \text{where } \phi(x-5) = \frac{1}{\sqrt{2\pi}} e^{-(x-5)^2/2} \\
& = E \{ h(X)w(X) \}, \text{ for } X \sim N(5,1)
\end{align*}
$$
Q. What are $h(x)$ and $w(x)$ in the above example? 

## Importance Sampling {.smaller}


$$
\begin{align*}
I  & = \int_{\mathbb{R}} 10 \exp(-2 | x- 5| ) \frac{\frac{1}{10} I(0 \le x \le 10)}{\phi(x-5)} \phi(x-5) dx \\
& = E \{ h(X)w(X) \}, \text{ for } X \sim N(5,1)
\end{align*}
$$

-  We can compare sides and find $h(x)$ and $w(x)$: 

$$
\begin{align*}
& \Rightarrow h(x) =  10 \exp(-2 | x- 5| ) \\
& \Rightarrow w(x) = \frac{\text{density function of } U(0,10)}{\text{density function of } N(5,1)} \\
& \text{in R-language, }w(x) = \frac{\text{dunif(x, 0, 10)}}{\text{dnorm(x, 5, 1)}}
\end{align*}
$$

<!-- ## Importance Sampling {.smaller} -->

<!-- $$ -->
<!-- \begin{align} -->
<!-- I & = \int_{0}^{10} 10 \exp(-2 | x- 5| ) \frac{\frac{1}{10}}{\phi(x-5)} \phi(x-5)dx \\ -->
<!-- & \text{ where }\phi(x-5) = \frac{1}{\sqrt{2\pi}} e^{-(x-5)^2/2} \; (\text{Normal density})  -->
<!-- \end{align} -->
<!-- $$ -->
<!-- Hence, we can write the integral as  -->

<!-- $$ -->
<!-- \begin{align} -->
<!-- I & = E(h(x)w(x)), \text{ for } X \sim N(5,1) \\ -->
<!-- & \text{ Here, } h(x) = 10 \exp(-2 | x- 5| ) \\ -->
<!-- & \text{ and } w(x) = \frac{\frac{1}{10}}{\phi(x-5)} = \frac{\texttt{dunif}(x, 0, 10)}{\texttt{dnorm}(x, mean=5, sd=1)} -->
<!-- \end{align} -->
<!-- $$  -->

## R code 

```{r, echo = TRUE}
w <- function(x) dunif(x, 0, 10)/dnorm(x, mean=5, sd=1)
f <- function(x) 10*exp(-2*abs(x-5))
X=rnorm(1e5,mean=5,sd=1)
Y=w(X)*f(X)

(I <- mean(Y))
(Error <- var(Y))
```

Same integral but the new error is 1/10th of the previous error. 

## Example 3

The standard double exponential or Laplace density is given by: 

$$
p(x) = \frac{1}{2} e^{-\lvert x \rvert}
$$

Calculate the second moment $E(X^2)$ for Laplace using the Normal(0,4) density as the proposal. In other words, calculate the integral: 

$$
\int_{-\infty}^{\infty}  \frac{1}{2} x^2 e^{-\lvert x \rvert} dx 
$$

Report both your estimate and the variance of your estimate. 

Could you also calculate the true value? 

## Solution 

```{r, echo = TRUE}
X <- rnorm(1e5, sd=4)
Y <- (X^2) * .5 * exp(-abs(X))/dnorm(X, sd=4)
(I <- mean(Y))
(Error <- var(Y))
```
 
**Think**: What happens if we use $N(0,1)$ instead? Or a Cauchy instead?

## True Value

The true value of the integral is 2. You can find that by integrating by hand, or, evaluating it in R. 

```{r, echo = TRUE}
h <- function(x) 0.5*x^2*exp(-abs(x))
integrate(h,-Inf,Inf)$val
```


