---
title: "Introduction to R"
subtitle: " Efficiency Considerations"
author: "Jyotishka Datta"
date: "February 3, 2020"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## apply, lapply, sapply, tapply 

- You have some structured blob of data and you want to perform some operations by dimensions. 
- e.g. calculate row-wise means or column-wise means 

```{r, echo = TRUE}
m <- matrix(data=cbind(rnorm(30, 0), rnorm(30, 2), rnorm(30, 5)), 
            nrow=30, ncol=3)
dim(m)
apply(m, 2, mean)
```

## apply (contd.)

- We can also calculate row-wise mean. 

```{r, echo = TRUE}
apply(m, 1, mean)
```
- Or, our own function: (how many negative numbers in each column?)
```{r, echo = TRUE}
apply(m, 2, function(x) length(x[x<0]))
```

## lapply and sapply 

- traversing over a set of data like a list or vector, and calling the specified function for each item. 

- Usage: lapply(alist, afunction) - applies afunction to all the elements of a list or vector, returns a list of
the results. Example:

```{r, echo = TRUE}
cap.state.name <- lapply(state.name, toupper)
head(cap.state.name, n = 3)
```

## sapply 

- Similar to lapply, but returns a simpler data structure: either a vector or array.

```{r, echo = TRUE}
sapply(1:3, function(x) x^2)
```

- Remember: lapply is a list apply, sapply is simple lapply. 
- Why do we care? apply, sapply, lapply are easier to read and sometimes faster than traditional for loops.


## Functions 

- Functions are blocks of code that allows R to be modular and facilitate code reuse.
- An R programmer can define their own functions as follows:
```{r, echo = TRUE, eval = FALSE}
function_name <- function([arg1], [arg2], ...){
#function code body
}
```
- The function arguments are optional. Function arguments are the variables passed
to the function, used by the function's code to perform calculations. A function can take no arguments.
- A function can also return any R primitive or object using the return(object)
statement.

## Functions 

-  Examples 
```{r, echo = TRUE, eval = FALSE}
# computes the mean of a vector of numbers
mean <- function(a_vector) {
s <- sum(a_vector)
x <- s/length(a_vector)
}
```

```{r, echo = TRUE, eval = FALSE}
# checks to see if a string s starts with letter x
startsWith <- function(x, s){
if(x == substr(s, 0, 1){
return(TRUE)
}
return(FALSE)
}
```

## Functions 

-  **Default arguments**
-  Function arguments can be assigned default values as follows:
```{r, echo = TRUE, eval = FALSE}
sort_vector <- function(a_vector, ascending=TRUE){
# sorting algorithm
}
```
- When calling this function, the arguments given default values do not need to be specified. In this case, the defaults are used.
- Example:
```{r, echo = TRUE, eval = FALSE}
sort_vector(a_vector) # returns a_vector in ascending order
sort_vector(a_vector, FALSE) # returns a vector in descending order
```

## Variable Scoping

-  Variables that are bound to an R object outside a function are called global variables, and are accessible everywhere in an R program.
-  Example:
```{r, echo = TRUE}
x <- 5
test <- function() {
cat(x + 5)
}
test(); 
```

## Functions 

-  Variables bound inside a function are only accessible within that function. These are called local variables. Example:
```{r, echo = TRUE}
x <- 10
test <- function() {
x <- 5
cat(x + 20)
}
test() #prints 25
cat(x + 20) #prints 30
```
-  **Note that the local variable assignment takes precedence inside the function test over the global assignment**.

## Functions 

-  R functions have no side effects-- they cannot change the value of global variables. Example:
```{r, echo =  TRUE}
x <- 10
test <- function(z) {
z <- z + 10
cat(z)
}
test(x) #prints 20
cat(x) #prints 10
```
-  **Note that x is still bound to 10 outside of test()**.


## CPU Times

```{r, echo = TRUE, eval = FALSE}
system.time(expr)
```

-  Returns CPU times that "expr" used. 
-  A numeric vector of length 5 containing the user cpu, system cpu, elapsed, subproc1, subproc2 times. 
-  The subproc times are the user and system cpu time used by child processes (and so are usually zero).

```{R, echo = TRUE}
system.time(for(i in 1:50) x<-mean(rnorm(1000)))
```


## `proc.time` function


- The other way to get system time is using proc.time() function.
- `proc.time' determines how much time (in seconds) the currently running R process already consumed. For example:

```{r, echo = TRUE}
now<-proc.time()
for(i in 1:50) x<-mean(rnorm(1000))
proc.time()-now
```

## Calculating mean 

```{r, echo = TRUE}
x<-c(1:10000)
sum<-0
now<-proc.time()
for(i in 1:10000){sum<-sum+x[i]}
meanx<-sum/10000
proc.time()-now
```

## Avoid for loops

- Use vectorized arithmetic instead of loops:
```{r, echo = TRUE}
system.time(sum(x)/length(x))
system.time(mean(x))
```

## Finer control 

```{r, echo = TRUE}
#install.packages("microbenchmark")
library(microbenchmark)
microbenchmark(sqrt(x),x^(0.5))
```

