---
title: "Lab Exercise Sampling 2"
author: "Jyotishka Datta"
date: "February 20, 2018"
output: 
  html_document:
         toc: true
         theme: flatly
         highlight: zenburn
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Rules:

1.  Email your answers to [stat5443.fall@gmail.com](stat5443.fall@gmail.com).
2.  No late submissions. 


## Problem 1 

-  We'll see that change of variable often helps with difficult sampling problems. Suppose you have a random variable $X$ with the following density function with an unknown normalizing constant: 
\[
f(x) = \exp\{-x^6/10\} \; \text{ where } x \in (0, \infty) 
\]
How do we sample from this density? 
Let's look at a general density of this class with parameters $k$ and $n$: 
\[
f(x) \propto \exp\{-k x^{n}\}, \; \text{ where } x \in (0, \infty), k > 0, n > 0.
\]
Show that the transformed variable $Y = X^{n}$ is a Gamma random variable with shape = $1/n$ and rate = $k$, following the steps outlined below: 

(1)  Show that under the transformation $x = y^{1/n}$, the transformed density will be: 
\[
f(y) \propto y^{\frac{1}{n}-1} \exp\{-k y\} \; \text{ where } y \in (0, \infty)
\]

(2)  Now, compare the density of $y$ with the Gamma density function with shape parameter $\alpha$ and rate parameter $\lambda$: 
\[
f(u) = \frac{\lambda^{\alpha} e^{-\lambda u} u^{\alpha - 1}}{\Gamma(\alpha)}, \; u \ge 0
\]
What are your parameters $\alpha$ and $\lambda$? 

(3)  In R, the function for generating $\text{Gamma}(\alpha, \lambda)$ random samples is `rgamma(n, shape = alpha, rate = lambda)`. Generate $10^5$ samples $Y$ from Gamma distribution with $\lambda = 1/10$ and $\alpha =1/6$. This will give you samples from the distribution of $X$.

Plot the histogram of $X = Y^{\alpha}$ along with the function $\exp\{-x^6/10\}$ to visually verify your sampler. You can use the following code for plotting (your plot should look like the Fig. 1): 

```{r, echo =F, cache = T}
lambda = 1/10
alpha = 1/6
n = 1e5
y = rgamma(n,shape = alpha, rate = lambda)
z = y^alpha
```

```{r, echo = T, fig.align = 'center', fig.asp=0.6}
hist(z, breaks = 30, freq = F, col = rgb(0.75,0.4,0.1,0.5)) # z is your sample
lambda = 1/10; alpha = 1/6
target <- function(x){exp(-lambda*x^(1/alpha))/
integrate(function(x) exp(-lambda*x^(1/alpha)),0,Inf)$value}
curve(target,lwd=2,add=T)
```


## Problem 2 

Suppose you have a distribution with density: 
\[
p(x) = \alpha \frac{x_m}{x^{\alpha+1}} \quad \text{where } x \ge x_m, \; x_m \ge 0, \alpha > 0. 
\]

(1)  Write an **one-line function** in `R` based on the inversion method to generate $n=100$ samples from the distribution where $\alpha$ and $x_m$ are the inputs for the function.

(2)  How would you estimate $\text{Median}(X)$ if $X$ has the distribution described above with $\lambda = 3$ using your samples from part (1)? 

(3)  The theoretical median of this distribution is given by:
\[
\text{Median}(X) = x_m 2^{1/\alpha}.
\]
How close is your estimate?

(3) The mean of this distribution is given as follows: 
\[
\mathbb{E}(X) = \begin{cases} 
\infty \text{ for } \alpha \le 1 \\
\frac{x_m \alpha}{(\alpha-1)} \text{ for } \alpha > 1.
\end{cases}
\]
Calculate the mean for $\alpha = 3$, and report your Monte Carlo mean versus the true mean. 

(4) The true mean is infinite for $\alpha = 0.75$, what happens to the Monte Carlo mean if you keep increasing the number of sample size $n = 10, \ldots, 100$ and calculate the Monte Carlo mean for $\alpha = 0.75$ for each $n$? What pattern do you observe? [You can plot the Monte Carlo mean vs. $n$ for different $\alpha$ and compare, like the one given below.]

```{r, cache = T}
paretomean <- function(n,x_m,lambda){return(mean(x_m/runif(n)^(1/lambda)))}
```

```{r, echo = T, fig.asp = 0.9, fig.align='center'}

lambda1 = 3
lambda2 = 0.75
nset = 10*{c(1:1e2)} # or any sequence you like

paretomeans.1 = paretomeans.2 = NULL # initialize

for(i in 1:length(nset)){
  n = nset[i]
  ## paretomean(n,x_m,lambda) returns the Monte Carlo mean
  
  paretomeans.1 = c(paretomeans.1,paretomean(n, x_m = 1, lambda = lambda1))
    paretomeans.2 = c(paretomeans.2,paretomean(n, x_m = 1, lambda = lambda2))
}

pareto.data = rbind(data.frame(values = paretomeans.1, 
                               type= "Lambda = 3"),
                    data.frame(values = paretomeans.2, 
                               type= "Lambda = 0.75"))
pareto.data = cbind(pareto.data, nset)

library(ggplot2) 
ggplot(pareto.data, aes(x = nset, y = values, group = type))+
  geom_line()+facet_grid(type~., scales = "free_y")
```








