---
title: "Lab Solutions"
author: "Jyotishka Datta"
date: "February 7, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Problem 1

A ``truncated" distribution is a distribution where the support of the random variable is restricted to some interval $[a,b]$. If we know the pdf $f(x)$ and cdf $F(x)$ of the original random variable, then the pdf and cdf of the truncated random variable can be easily derived as: 
$$
f_{Tr}(x) = \begin{cases}
\frac{f(x)}{F(b) - F(a)} \; \forall a < x \le b \\
0 \; \text{otherwise}
\end{cases}
$$
Consider the exponential random variable $X \sim \text{Exp}(\lambda)$, with pdf 
$$
f(x) = \lambda e^{-\lambda x}, \; x \ge 0, \lambda > 0 
$$

Answer the following: 

1.  How would you generate from an exponential truncated to the interval $[0,b]$ using accept-reject scheme? Try the R code for $\lambda = 2$, $b = 5$.

2.  Can you use the inverse CDF transformation here? 


## Solution 

Part 1 is easy. All you have to do is sample from $Exp(\lambda)$ and discard all samples that fall outside the truncation boundary. 
Also, to generate from exponential we can use the inverse CDF method with 
$$
F^{-1}(u) = - \frac{1}{\lambda} \log(1-u)
$$

```{r, echo = TRUE}
set.seed(123)
n = 1000
u = runif(n)
lambda = 2 
a = 0
b = 5

x = (-1/lambda)*log(1-u*(1-exp(-lambda*b)))
## Standard Exponential then truncate 
x = -1/lambda*log(u) #u ~ U[0,1] implies 1-u ~ U[0,1]
x = x[x<=b]

## Plotting
library(ggplot2)
ggplot(data.frame(x))+geom_histogram(aes(x=x),bins = 50)
```

Part 2 involves a little math! The pdf of the truncated distribution is given above. We can then calculate the CDF of the truncated exponential: 

First the area between $[a,b]$ would be 
$$
\int_{a}^{b} \lambda e^{-\lambda x} d x = F(b)-F(a) = e^{-\lambda a} - e^{-\lambda b}
$$

The CDF for the truncated distribution can be written as: 
$$
F_{tr}(x) = \frac{F(x)-F(a)}{F(b)-F(a)}, a \le x \le b 
$$

Since $F(0) = 0$ for exponential, the inverse can be derived by solving the following equation for $x$ 
$$
\text{Solve } u = \frac{F(x)-F(a)}{F(b)-F(a)} = \frac{1- \exp(-\lambda x)}{1- exp(-\lambda b)} \\
1 - exp(-\lambda x) = u (1 - exp(-\lambda b)) \\
x = (-1/\lambda)*\log(1-u (1 - exp(-\lambda b)))
$$
The R code can be written as follows: 

```{r, echo = TRUE}
n = 1000
lambda = 2
b = 5
u = runif(n)
x = (-1/lambda)*log(1-u*(1-exp(-lambda*b)))

## Plotting
library(ggplot2)
ggplot(data.frame(x))+geom_histogram(aes(x=x),bins = 50)
```



