---
title: "Lab Exercise: Sampling"
author: "Jyotishka Datta"
date: "February 17, 2020"
output: 
  html_document:
         toc: true
         theme: flatly
         highlight: zenburn
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Rules:

1.  Email your answers to [stat5443.fall@gmail.com](stat5443.fall@gmail.com).
2.  No late submissions. 


## Problem 1 

-  We'll see that change of variable often helps with difficult sampling problems. Suppose you have a random variable $X$ with the following density function with an unknown normalizing constant: 
\[
f(x) = \exp\{-x^6/10\} \; \text{ where } x \in (0, \infty) 
\]
How do we sample from this density? 

-  Let's look at a general density of this class with parameters $k$ and $n$: 
\[
f(x) \propto \exp\{-k x^{n}\}, \; \text{ where } x \in (0, \infty), k > 0, n > 0.
\]
Show that the transformed variable $Y = X^{n}$ is a Gamma random variable with shape = $1/n$ and rate = $k$, following the steps outlined below: 

(1)  Show that under the transformation $x = y^{1/n}$, the transformed density will be: 
\[
f(y) \propto y^{\frac{1}{n}-1} \exp\{-k y\} \; \text{ where } y \in (0, \infty)
\]

(2)  Now, compare the density of $y$ with the Gamma density function with shape parameter $\alpha$ and rate parameter $\lambda$: 
\[
f(u) = \frac{\lambda^{\alpha} e^{-\lambda u} u^{\alpha - 1}}{\Gamma(\alpha)}, \; u \ge 0
\]
What are your parameters $\alpha$ and $\lambda$? 

(3)  In R, the function for generating $\text{Gamma}(\alpha, \lambda)$ random samples is `rgamma(n, shape = alpha, rate = lambda)`. Generate $10^5$ samples $Y$ from Gamma distribution with $\lambda = 1/10$ and $\alpha =1/6$. This will give you samples from the distribution of $Y$. Then you can get $X$ samples by $X = Y^{1/n}$. 

Plot the histogram of $X = Y^{\alpha}$ along with the function $\exp\{-x^6/10\}$ to visually verify your sampler. You can use the following code for plotting (your plot should look like the Fig. 1): 

```{r, echo =F, cache = T}
lambda = 1/10
alpha = 1/6
n = 1e5
y = rgamma(n,shape = alpha, rate = lambda)
z = y^alpha
```

```{r, echo = T, fig.align = 'center', fig.asp=0.6}
hist(z, breaks = 30, freq = F, col = rgb(0.75,0.4,0.1,0.5)) # z is your sample
lambda = 1/10; alpha = 1/6
target <- function(x){exp(-lambda*x^(1/alpha))/
integrate(function(x) exp(-lambda*x^(1/alpha)),0,Inf)$value}
curve(target,lwd=2,add=T)
```


## Problem 2 (Truncated distribution)

A ``truncated" distribution is a distribution where the support of the random variable is restricted to some interval $[a,b]$. If we know the pdf $f(x)$ and cdf $F(x)$ of the original random variable, then the pdf and cdf of the truncated random variable can be easily derived as: 
$$
f_{Tr}(x) = \begin{cases}
\frac{f(x)}{F(b) - F(a)} \; \forall a < x \le b \\
0 \; \text{otherwise}
\end{cases}
$$
Consider the exponential random variable $X \sim \text{Exp}(\lambda)$, with pdf 
$$
f(x) = \lambda e^{-\lambda x}, \; x \ge 0, \lambda > 0 
$$

Answer the following: 

1.  How would you generate from an exponential truncated to the interval $[0,b]$ using accept-reject scheme? Try the R code for $\lambda = 2$, $b = 5$.

2.  Can you use the inverse CDF transformation here? 

## Solution (Problem 2)

Part 1 is easy. All you have to do is sample from $Exp(\lambda)$ and discard all samples that fall outside the truncation boundary. 
Also, to generate from exponential we can use the inverse CDF method with 
$$
F^{-1}(u) = - \frac{1}{\lambda} \log(1-u)
$$

```{r, echo = TRUE}
set.seed(123)
n = 1000
u = runif(n)
lambda = 2 
a = 0
b = 5

x = (-1/lambda)*log(1-u*(1-exp(-lambda*b)))
## Standard Exponential then truncate 
x = -1/lambda*log(u) #u ~ U[0,1] implies 1-u ~ U[0,1]
x = x[x<=b]

## Plotting
library(ggplot2)
ggplot(data.frame(x))+geom_histogram(aes(x=x),bins = 50)
```

Part 2 involves a little math! The pdf of the truncated distribution is given above. We can then calculate the CDF of the truncated exponential: 

First the area between $[a,b]$ would be 
$$
\int_{a}^{b} \lambda e^{-\lambda x} d x = F(b)-F(a) = e^{-\lambda a} - e^{-\lambda b}
$$

The CDF for the truncated distribution can be written as: 
$$
F_{tr}(x) = \frac{F(x)-F(a)}{F(b)-F(a)}, a \le x \le b 
$$

Since $F(0) = 0$ for exponential, the inverse can be derived by solving the following equation for $x$ 
$$
\text{Solve } u = \frac{F(x)-F(a)}{F(b)-F(a)} = \frac{1- \exp(-\lambda x)}{1- \exp(-\lambda b)} \\
1 - \exp(-\lambda x) = u (1 - \exp(-\lambda b)) \\
x = (-1/\lambda)*\log(1-u (1 - \exp(-\lambda b)))
$$
The R code can be written as follows: 

```{r, echo = TRUE}
n = 1000
lambda = 2
b = 5
u = runif(n)
x = (-1/lambda)*log(1-u*(1-exp(-lambda*b)))

## Plotting
library(ggplot2)
ggplot(data.frame(x))+geom_histogram(aes(x=x),bins = 50)
```
