---
title: "One-way Analysis of Variance"
subtitle: "Examples in R"
author: 
- name: Jyotishka Datta
  affiliation: University of Arkansas, Fayetteville. 
date: "`r format(Sys.time(), '%d %B, %Y')`"
tags: [ANOVA, examples, R]
output:
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both **content** as well as the output of **any embedded R code chunks** within the document.

I will show here the R codes for ANOVA. The two examples are taken from the following websites.

  -  [Spray Insects Example](http://homepages.inf.ed.ac.uk/bwebb/statistics/ANOVA_in_R.pdf)
  -  [Migraine Example (Martin Lindquist)](http://www.stat.columbia.edu/~martin/W2024/R3.pdf)


## First Example

We first read the InsectSprays dataset that has data on pain levels for three types of drugs. We create a data frame in R to store the data. 

```{r, echo=TRUE, eval=TRUE}
pain = c(4, 5, 4, 3, 2, 4, 3, 4, 4, 6, 8, 4, 5, 4, 6, 5, 
         8, 6, 6, 7, 6, 6, 7, 5, 6, 5, 5)
drug = c(rep("A",9), rep("B",9), rep("C",9))
migraine = data.frame(pain,drug)
str(migraine)
```

### Descriptive statistics
-  Mean, variance, number of elements in each cell
-  Visualise the data – boxplot; look at distribution, look for outliers
-  We’ll use the tapply() function which is a helpful shortcut in processing data. 
-  tapply() basically allows you to specify a response variable, a factor (or factors) and a function that should be applied to each subset of the response variable defined by each level of the factor. 

Instead of doing 
```{r, echo = TRUE, eval = TRUE}
mean(pain[drug=="A"])
mean(pain[drug=="B"])
```
We can simply call 
```{r,echo = TRUE, eval = TRUE}
tapply(pain, drug, mean)
```

-  We need the sample variances, i.e. $s_j^2$ for $j = 1, 2, 3$.
```{r,echo = TRUE, eval = TRUE}
tapply(pain, drug, var)
```
-  We also need the sample sizes for each group. (Note: it will not always be equal or obvious!)
```{r,echo = TRUE, eval = TRUE}
tapply(pain, drug, length)
```

## Visual comparison 

We can look at the boxplot for the three groups: 

```{r, echo = TRUE}
plot(pain ~ drug, data=migraine)
```

### More Visual comparisons
R package 'mosaic' and 'gmodel' can create better looking plots and summary:

```{r, echo = TRUE, warning=FALSE, message=FALSE}
require(mosaic)
require(gmodels)
options(digits=3)
favstats(pain ~ drug, data=migraine)
```

### Histogram for the three groups

```{r, echo = TRUE, warning=FALSE, message=FALSE}
trellis.par.set(theme=col.mosaic()) # better color scheme for lattice
histogram(~ pain| drug, fit="normal", layout=c(1, 3), data=migraine)
```

### Density Plots for the three groups

```{r, echo = TRUE, warning=FALSE, message=FALSE}
trellis.par.set(theme=col.mosaic()) # better color scheme for lattice
densityplot(~ pain, groups=drug, auto.key=TRUE, data=migraine)
```

### Box and Whisker Plots for the three groups

```{r, echo = TRUE, warning=FALSE, message=FALSE}
trellis.par.set(theme=col.mosaic()) # better color scheme for lattice
bwplot(pain ~ drug, data=migraine)
```


### ANOVA Sum of squares 
-  We need to calculate two sum of squares for the F-statistic. 
-  The numertaor is between groups variation or $\sum_{j=1}^{K} n_j (\bar{x}_j - \bar{X})^2$. ($n_j$: sample size of the $j^{th}$ group).
- We have calculated the $n_j$'s and the $\bar{x}_j$'s before using tapply().

```{r, echo = TRUE}
n_j = tapply(pain,drug,length)
SST_j = n_j * (tapply(pain,drug,mean)-mean(pain))^2
SST = sum(SST_j)
cat("SST = ", SST, '\n')
```

### SSE
Variation within the treatments = Sum of squares for Error (SSE):
$$
SSE = (n_1 -1)s_1^2 + (n_2 - 1)s_2^2 + \cdots + (n_K -1)s_K^2 
$$

```{r, echo = TRUE}
SSE_j = (n_j-1)*tapply(pain,drug,var)
SSE = sum(SSE_j)
cat("SSE =", SSE, "\n")
```

### F-statistic 

-  Mean square for treatments = MST = $\frac{SST}{k-1} = \frac{28.2}{2} = 14.11$
-  Mean square for errors = MSE = $\frac{SSE}{n-k} = \frac{28.4}{27-3} = 1.19$.
-  The ratio of the variability among the treatment means to that within the treatment means is an $F$-statistic:
$$
				F_{k-1,n-k} = \frac{MST}{MSE} = \frac{14.11}{1.19} = 11.9
$$
with $k-1 = 2$ numerator and $n-k = 24$ denominator degrees of freedom.
-  P-value 
```{r, echo = TRUE}
p.value = 1-pf(11.9,2,24);cat("P value =", p.value, "\n")
```
Reject the null.

### ANOVA in R 

Of course, we can skip the step-by-step calculation and use a pre-canned package in R: 

```{r, echo = TRUE}
results <- aov(pain~drug, data=migraine)
summary(results)
```

## Another Example

We are going to see another data-set InsectSprays. 6 different insect sprays (1 Categorical Variable with 6 levels) were tested to see if there was a difference in the number of insects found in the field after each spraying (Dependent Variable).
```{r, echo=TRUE, message= FALSE, warning = FALSE}
attach(InsectSprays)
data(InsectSprays)
str(InsectSprays)
```

### Descriptive Statistics 
```{r, echo=TRUE, message= FALSE, warning = FALSE}
tapply(count, spray, mean)
tapply(count, spray, var)
tapply(count, spray, length)
```

### More Visual comparisons
R package 'mosaic' and 'gmodel' can create better looking plots and summary:

```{r, echo = TRUE, warning=FALSE, message=FALSE}
require(mosaic)
require(gmodels)
options(digits=3)
favstats(count ~ spray, data=InsectSprays)
```

### Histogram for the three groups

```{r, echo = TRUE, warning=FALSE, message=FALSE}
trellis.par.set(theme=col.mosaic()) # better color scheme for lattice
histogram(~ count| spray, fit="normal", layout=c(2, 3), data=InsectSprays)
```

### Density Plots for the three groups

```{r, echo = TRUE, warning=FALSE, message=FALSE}
trellis.par.set(theme=col.mosaic()) # better color scheme for lattice
densityplot(~ count, groups=spray, auto.key=TRUE, data=InsectSprays)
```

### Box and Whisker Plots for the three groups

```{r, echo = TRUE, warning=FALSE, message=FALSE}
trellis.par.set(theme=col.mosaic()) # better color scheme for lattice
bwplot(count ~ spray, data=InsectSprays)
```


### ANOVA Sum of squares 
-  We need to calculate two sum of squares for the F-statistic. 
-  The numertaor is between groups variation or $\sum_{j=1}^{K} n_j (\bar{x}_j - \bar{X})^2$. ($n_j$: sample size of the $j^{th}$ group).
- We have calculated the $n_j$'s and the $\bar{x}_j$'s before using tapply().

```{r, echo = TRUE}
n_j = tapply(count,spray,length)
SST_j = n_j * (tapply(count,spray,mean)-mean(count))^2
SST = sum(SST_j)
cat("SST = ", SST, '\n')
```

### SSE
Variation within the treatments = Sum of squares for Error (SSE):
$$
SSE = (n_1 -1)s_1^2 + (n_2 - 1)s_2^2 + \cdots + (n_K -1)s_K^2 
$$

```{r, echo = TRUE}
SSE_j = (n_j-1)*tapply(count,spray,var)
SSE = sum(SSE_j)
cat("SSE =", SSE, "\n")
```

### F-statistic 

-  Mean square for treatments = MST = $\frac{SST}{k-1} = \frac{2669}{5} = 533.8$
-  Mean square for errors = MSE = $\frac{SSE}{n-k} = \frac{1015}{72-6} = 15.378$.
-  The ratio of the variability among the treatment means to that within the treatment means is an $F$-statistic:
$$
				F_{k-1,n-k} = \frac{MST}{MSE} = \frac{533.8}{15.378} = 34.7
$$
with $k-1 = 5$ numerator and $n-k = 66$ denominator degrees of freedom.
-  P-value 
```{r, echo = TRUE}
p.value = 1-pf(34.7,5,66);cat("P value =", p.value, "\n")
```
Reject the null.

### ANOVA in R 

Of course, we can skip the step-by-step calculation and use a pre-canned package in R: 

```{r, echo = TRUE}
results <- aov(count~spray, data=InsectSprays)
summary(results)
```