---
title: "Introduction to R - II"
author: "Jyotishka Datta"
date: "January 30, 2017"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = T, warning = F, message = F)
```

## for loop

```{R, echo = T}
x = 36 
for (x in seq(10)){
  cat(x, "\t") # x is the iterator
}
```

## for loop 

```{r, echo = T}
for (x in seq(10)){
  x = 3.9*x*(1-x)
  cat(x, "\t")
}
```

## while loop 

```{r, echo = T}
a = 0 
while(a<10){
  a = a+1 ## unlike C, a++, a+=1 won't work in R
  cat(a, "\t")
}
```


## apply, lapply, sapply, tapply 

- You have some structured blob of data and you want to perform some operations by dimensions. 
- e.g. calculate row-wise means or column-wise means 

```{r, echo = TRUE}
m <- matrix(data=cbind(rnorm(30, 0), rnorm(30, 2), rnorm(30, 5)), nrow=30, ncol=3)
dim(m)
apply(m, 2, mean)
```

## apply (contd.)

- We can also calculate row-wise mean. 

```{r, echo = TRUE}
apply(m, 1, mean)
```
- Or, our own function: (how many negative numbers in each column?)
```{r, echo = TRUE}
apply(m, 2, function(x) length(x[x<0]))
```

## lapply and sapply 

- traversing over a set of data like a list or vector, and calling the specified function for each item. 

- Usage: lapply(alist, afunction) - applies afunction to all the elements of a list or vector, returns a list of
the results. Example:

```{r, echo = TRUE}
cap.state.name <- lapply(state.name, toupper)
head(cap.state.name, n = 3)
```

## sapply 

- Similar to lapply, but returns a simpler data structure: either a vector or array.

```{r, echo = TRUE}
sapply(1:3, function(x) x^2)
```

- Remember: lapply is a list apply, sapply is simple lapply. 
- Why do we care? apply, sapply, lapply are easier to read and **sometimes** faster than traditional for loops.


## Functions 

- Functions are blocks of code that allows R to be modular and facilitate code reuse.
- An R programmer can define their own functions as follows:
```{r, echo = TRUE, eval = FALSE}
function_name <- function([arg1], [arg2], ...){
#function code body
}
```
- The function arguments are optional. Function arguments are the variables passed to the function, used by the function's code to perform calculations. A function can take no arguments.
- A function can also return any R primitive or object using the return(object) statement.

## For loop vs apply 

```{R, echo = T}
N = 1e5
d <- as.data.frame(cbind(rnorm(N),runif(N)))

system.time(for (i in 1:N) {
    d$mean2[i] <- mean(c(d[i, 1], d[i, 2]))
})

system.time(d$mean1 <- apply(d, 1, mean))
```

## Functions 

-  Examples 
```{r, echo = TRUE, eval = FALSE}
# computes the mean of a vector of numbers
mean <- function(a_vector) {
s <- sum(a_vector)
x <- s/length(a_vector)
}
```

```{r, echo = TRUE, eval = FALSE}
# checks to see if a string s starts with letter x
startsWith <- function(x, s){
if(x == substr(s, 0, 1){
return(TRUE)
}
return(FALSE)
}
```

## Functions 

-  **Default arguments**
-  Function arguments can be assigned default values as follows:
```{r, echo = TRUE, eval = FALSE}
sort_vector <- function(a_vector, ascending=TRUE){
# sorting algorithm
}
```
- When calling this function, the arguments given default values do not need to be specified. In this case, the defaults are used.
- Example:
```{r, echo = TRUE, eval = FALSE}
sort_vector(a_vector) # returns a_vector in ascending order
sort_vector(a_vector, FALSE) # returns a vector in descending order
```

## Functions 

-  Variable Scoping
-  Variables that are bound to an R primitive or object outside a function are called global variables, and are accessible everywhere in an R program.
-  Example:
```{r, echo = TRUE}
x <- 5
test <- function() {
cat(x + 5)
}
test(); 
```

## Functions 

-  Variables bound inside a function are only accessible within that function. These are called local variables. 
```{r, echo = TRUE}
x <- 10
test <- function() {
x <- 5; cat(x + 20)
}
test() #prints 25
cat(x + 20) #prints 30
```
**local variable assignment takes precedence inside the function over the global assignment**.

## Functions 

-  R functions have no side effects-- they cannot change the value of global variables. Example:
```{r, echo =  TRUE}
x <- 10
test <- function(z) {
z <- z + 10
cat(z)
}
test(x) #prints 20
cat(x) #prints 10
```
-  **Note that x is still bound to 10 outside of test()**.

## System time in R
```{r, echo = TRUE, eval = FALSE}
system.time(expr)
```

Return CPU times that "expr" used. A numeric vector of length 5 containing 
the user cpu, system cpu, elapsed, subproc1, subproc2 times. The subproc 
times are the user and system cpu time used by child processes (and so 
are usually zero).

## System time in R

```{R, echo = TRUE}
system.time(for(i in 1:5000) x<-mean(rnorm(1000)))
```

- The other way to get system time is using proc.time() function.
- `proc.time' determines how much time (in seconds) the currently running R process already consumed. For example:

```{r, echo = TRUE}
now<-proc.time()
for(i in 1:5000) x<-mean(rnorm(1000))
proc.time()-now
```


## Vectorized arithmetic > loops

```{r, echo = TRUE}
x<-rnorm(1e7)
sum<-0
now<-proc.time()
for(i in 1:1e7){sum<-sum+x[i]}
meanx<-sum/1e7
proc.time()-now
```

- Use vectorized arithmetic instead of loops:
```{r, echo = TRUE}
system.time(mean(x))
```

## Finer control

```{r, echo = TRUE}
x = seq(1:1e4)
#install.packages("microbenchmark")
library(microbenchmark)
microbenchmark(sqrt(x),x^(0.5))
```

## Excercise 1

Here are two other ways to compute the square root of a vector. Which do you think will be fastest? Which will be slowest? Use microbenchmarking to test your answers.

```{r, eval = F, echo = T}
x ^ (1 / 2)
exp(log(x) / 2)
```

# More advanced topics in effcient R programming

## Time comparison: Ordinary Least Squares in Regression

- Compare execution times for the different methods of solving the Least Squares problem when the number of observations is large.

-  Let us start by making up our own random data: $n \times p$ design matrix $X$ and an $n \times 1$ response vector $y$. 
-  Take $n = 5 \times 10^5$. 

```{r, echo = TRUE, cache = TRUE}
set.seed(123)
n <- 5e5; p <- 20
X <- matrix(rnorm(n * p, mean = 1 : p, sd = 10), nr = n, nc = p, byrow = TRUE)
y <- rowSums(X) + rnorm(n)
```

## OLS Solution 

- The naive approach to solve the LS problem is to compute: 
$\hat{\beta} = (X^T X)^{-1}X^T y$

```{r, echo = TRUE, cache = TRUE}
system.time(bHat1 <- solve(t(X) %*% X) %*% t(X) %*% y)
```

-  Not so naive: use `crossprod`.

```{r, echo = TRUE, cache = TRUE}
system.time(bHat2 <- solve(crossprod(X), crossprod(X, y)))
```

## Knowledge of Linear Algebra helps

- Inverting a matrix using Cholesky needs less time. 
- A symmetric positive definite matrix can be factorized as: 
$$
A = L L^T
$$

```{r, echo = TRUE, cache = TRUE}
p = 1000; A = array(rnorm(p*p), c(p,p))
A = crossprod(A)
ptm <- proc.time(); B = solve(A); proc.time()-ptm
ptm <- proc.time(); B = chol2inv(A); proc.time()-ptm
```

## More advanced : Parallel

- If you have a task that has to be completed many times, but each repeat is independent, you should use parallel computing. 
- For instance, if you are repreating a test, and in each case are using a new data-set, that is an easily parallelizable task.
- A good rule of thumb is that if you can wrap your task in an apply function or one of its variants, it's a good option for parallelization. 
- Parallel processing breaks up your task, splits it among **multiple** processors, and puts the components back together. [https://hpc.uark.edu/](https://hpc.uark.edu/)

## Parallelism in R 

- You can use R packages "doParallel" or "snow" and most parellelizing tasks are easy to implement. 
- We will come back to Parallel computing when we talk about a more advanced nonparametric method: "Bootstrap"

# Permutation test

## Permutation test (Recap)

R function `combn` can be used to produce the indices corresponding to all $\binom{6}{3}$ ways of choosing 3 elements out of 6 elements. 

```{R, echo = T}
(idx = combn(x = 6, m=3))
```


## Permutation test 

```{r, echo = T}
x=c(37,55,57)
y=c(23,31,70)
xy = c(x,y) # the combined data set

permut = NULL # the permuted data set (a 20*6 matrix)

for(i in 1:ncol(idx)){
  permut = rbind(permut, c(xy[idx[,i]], xy[-idx[,i]]))
}
permut.x = permut[, 1:3] # the permuted X matrix (20*3)
permut.y = permut[, 4:6] # the permuted Y matrix (20*3)
head(permut.x, n = 3)
```

## Permutation test 

```{r, echo = T, fig.asp = 0.5}
delta1 = apply(permut.x, 1, mean) - apply(permut.y, 1, mean)
delta1.obs = mean(x)-mean(y)
```

P-value for permutation of sample mean

```{r, echo = T, fig.asp = 0.5}
(pval1.upper = mean(delta1 >= delta1.obs)) #upper-tailed
(pval1.2sided = mean(abs(delta1) >= abs(delta1.obs))) #two-tailed
```

## Histogram 

```{r, echo = T, fig.asp = 0.5}
hist(delta1,breaks=10);abline(v = delta1.obs,col="red")
```


## Permutation test code {.smaller}

```{r, eval=F, echo = T}
idx = combn(x = 6, m=3)
x=c(37,55,57)
y=c(23,31,70)
xy = c(x,y) # the combined data set
permut = NULL # the permuted data set (a 20*6 matrix)

for(i in 1:20){
  permut = rbind(permut, c(xy[idx[,i]], xy[-idx[,i]]))
}
permut.x = permut[, 1:3] # the permuted X matrix (20*3)
permut.y = permut[, 4:6] # the permuted Y matrix (20*3)

delta1 = apply(permut.x, 1, mean) - apply(permut.y, 1, mean)

delta1.obs = mean(x)-mean(y)

#pvalue for permutation of sample mean
(pval1.upper = mean(delta1 >= delta1.obs)) #upper-tailed
(pval1.2sided = mean(abs(delta1) >= abs(delta1.obs))) #two-tailed
hist(delta1,breaks=10);abline(v = delta1.obs,col="red")
```


## Exercise 2 {.smaller}

Carter and Hubert (1985) give data for percentage variation
in blood sugar over 1-hour periods for rabbits given two different
dose levels of a drug, Is there evidence of a response difference between
levels? Apply a Permutation test. 
```{r, echo = T, fig.asp = 0.5}
dose.1 = c(0.21,16.20,10.10,8.67,11.13,1.96,10.19,15.87,12.81)
dose.2 = c(1.59,2.66,6.27,2.32,10.87,7.23,3.76,3.02,15.01)
boxplot(dose.1,dose.2)
```

