---
title: "Wilcoxon and Quantiles"
author: "Jyotishka Datta"
date: "September 7, 2018"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)
```

## Wilcoxon Test with Ties 

-  Will show you warning message ! 

```{r}
wilcox.test(c(-2, -2, 2, 2, 5, 5, 5), mu = 3)
```


## Suppress Warnings !

-  We can suppress this warning message, but it's not recommended in practice

```{r}
suppressWarnings(wilcox.test(c(-2,-2,2,2,5,5,5),mu = 3))
```


## Another Example 

```{r, fig.height = 3}
x2 <- c(5.6, 6.1, 6.3, 6.4, 6.5, 6.6, 7.0, 7.5, 7.9, 8.0,
        8.0, 8.1, 8.1, 8.2, 8.4, 8.5, 8.7, 9.4, 14.3, 26.0)
## Plot the density
plot(density(x2), main = "Water Content")
```

## Wilcoxon Signed Rank Test 

```{R}
suppressWarnings(wilcox.test(x2, mu=9, conf.int=TRUE))
```


## Sign Test 

```{r}
binom.test(sum(x2>9),length(x2),alternative = "two.sided")
```

## Parametric t-test 

```{R}
t.test(x2,alternative = "two.sided",mu=9)
```

## Make a nice looking plot 

```{R, message = FALSE, Warnings = FALSE}
library(ggplot2)
qplot(y=x2, x= 1, geom = "violin")+geom_abline(slope=0,intercept = 9)
```

# Quantiles 


## The ecdf() function in R {.smaller}

```{r, echo = FALSE, results = "asis"}
static_help <- function(pkg, topic, out, links = tools::findHTMLlinks()) {
  pkgRdDB = tools:::fetchRdDB(file.path(find.package(pkg), 'help', pkg))
  force(links)
  tools::Rd2HTML(pkgRdDB[[topic]], out, package = pkg,
                 Links = links, no_links = is.null(links))
}
tmp <- tempfile()
static_help("stats", "ecdf", tmp)
out <- readLines(tmp)
headfoot <- grep("body", out)
cat(out[(headfoot[1] + 5):(headfoot[2] - 1)], sep = "\n")
```

## Convergence

[https://jdatta.shinyapps.io/eCDFdemo/](https://jdatta.shinyapps.io/eCDFdemo/)


## Empirical CDF {.smaller}

```{r}
set.seed(123)
emp1 <- ecdf(rnorm(20)) # E-CDF of 20 samples from N(0,1)
emp2 <- ecdf(rnorm(50)) # E-CDF of 20 samples from N(0,1)
```

```{r, echo = F}
par(mfrow=c(1,2)) # Two panels
x <- seq(-3,3,length.out = 100) # grid of values from -3 to 3
plot(emp1,verticals = TRUE,lwd = 2,main="n = 20")
lines(x,pnorm(x))
plot(emp2,verticals = TRUE,lwd = 2,main="n = 50")
lines(x,pnorm(x))
```

## Q-Q Plots (One-sample)

```{r}
set.seed(12) # Reproducibility
y <- rnorm(100)
qqnorm(y, ylim=c(-3,3), main = "Normal Q-Q Plot",
       xlab = "Theoretical Quantiles", ylab = "Sample Quantiles",
       plot.it = TRUE)
qqline(y, distribution = qnorm)
```

## Two samples same 

```{r}
z <- rnorm(100)
qqplot(y,z,main = "Two-sample Q-Q Plot",
       xlab = "Sample 1 Quantiles", ylab = "Sample 2 Quantiles")
abline(0, 1)
```


## Two samples different

```{R, fig.height = 3.5}
x = rnorm(10000,0,2); y = rnorm(10000,0,4)
qqplot(x,y,main = "Two-sample Q-Q Plot",
       xlab = "Sample 1 Quantiles", ylab = "Sample 2 Quantiles")
abline(0, 1)
```

## A Common Mistake 

```{R, fig.height = 3.5}
x = rnorm(10000,0,2); y = rnorm(10000,0,4)
qqplot(x,y,main = "Two-sample Q-Q Plot",
       xlab = "Sample 1 Quantiles", ylab = "Sample 2 Quantiles")
```

## Histograms 

```{R}
hist(x,breaks=50,freq=F,col=rgb(1,0,0,0.5),xlim=c(-15,15))
hist(y,breaks=50,freq=F,col=rgb(0,0,1,0.5),add=T)
box()
```

## Or, Boxplots 

```{R}
boxplot(y,z)
```

## Even better, test a hypothesis 

```{r}
ks.test(rnorm(20),pnorm)
```

## In our case 

```{r}
ks.test(x,y)
```

## Next Time 

-  We will dig deeper into these tests ! 
-  Kolmogorov-Smirnov, Lilliefors, Chi-square GoF etc. 
-  These are important as they tell you whether your data distribution is Normal, i.e. whether we should use nonparametric methods that do not assume Normality or parametric methods that assume Normality. 


