---
title: "Order Statistics, Ranks & Goodness-of-Fit Tests"
subtitle: "From Empirical CDFs to the Kolmogorov–Smirnov Test"
author: "Jyotishka Datta"
institute: "Virginia Tech"
date: today
format:
  revealjs:
    slide-number: true
    progress: true
    chalkboard: true
    code-fold: false
    code-line-numbers: true
    highlight-style: github
    fig-align: center
    smaller: true
    scrollable: false
    incremental: false
    html-math-method: mathjax
    transition: slide
    background-transition: fade
    footer: "STAT — Virginia Tech | Jyotishka Datta"
    logo: "vt.png"
execute:
  echo: true
  warning: false
  message: false
  fig-width: 7
  fig-height: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center")
```

# Part 1: Order Statistics {background-color="#861F41"}

## What Are Order Statistics?

Given a random sample $X_1, X_2, \ldots, X_n$, the **order statistics** are the sample values arranged in non-decreasing order:

$$X_{(1)} \leq X_{(2)} \leq \cdots \leq X_{(n)}$$

- $X_{(1)}$ = **sample minimum**
- $X_{(n)}$ = **sample maximum**
- $X_{(i)}$ = $i^{\text{th}}$ smallest value

::: {.callout-important}
Order statistics are **random quantities** — their distribution depends on the parent distribution $F$.
:::

## Familiar Statistics Built From Order Statistics

- **Sample median:**
$$M = \begin{cases} X_{\left(\frac{n+1}{2}\right)} & n \text{ odd} \\[4pt] \frac{1}{2}\!\left(X_{\left(\frac{n}{2}\right)} + X_{\left(\frac{n}{2}+1\right)}\right) & n \text{ even} \end{cases}$$

- **Sample range:** $X_{(n)} - X_{(1)}$ — measures spread

- **Sample midrange:** $\frac{X_{(n)} + X_{(1)}}{2}$ — alternative center

## Distribution of the Maximum

Let $X_1, \ldots, X_n \overset{\text{iid}}{\sim} F$ (continuous). Let $G$ be the CDF of $X_{(n)}$.

The maximum is $\leq t$ iff **every** observation is $\leq t$:

$$G(t) = P(X_{(n)} \leq t) = \prod_{i=1}^{n} P(X_i \leq t) = F^n(t)$$

Differentiating:

$$\boxed{g(t) = n\, F^{n-1}(t)\, f(t)}$$

## Distribution of the Minimum

The minimum is $\geq t$ iff **every** observation is $\geq t$:

$$H(t) = P(X_{(1)} \leq t) = 1 - (1 - F(t))^n$$

Differentiating:

$$\boxed{h(t) = n\,(1-F(t))^{n-1}\, f(t)}$$

## General $r^{\text{th}}$ Order Statistic

$$\boxed{f_{X_{(r)}}(x) = \frac{n!}{(r-1)!\,(n-r)!}\, F^{r-1}(x)\,(1-F(x))^{n-r}\, f(x)}$$

**Memory aid:** Think of it as a multinomial:

- $(r-1)$ observations **below** $x$: probability $F^{r-1}(x)$
- **1** observation **at** $x$: probability $f(x)\,dx$
- $(n-r)$ observations **above** $x$: probability $(1-F(x))^{n-r}$

::: {.callout-tip}
When $F = \text{Uniform}(0,1)$, this simplifies to a **Beta$(r,\, n-r+1)$** density.  
This fact is the key to the KS distribution-free property!
:::

## Simulating Order Statistic Distributions

```{r order-stat-demo}
#| fig-height: 3.5
#| code-fold: true
set.seed(42); n <- 10; B <- 50000
sims <- replicate(B, sort(runif(n)))

par(mfrow = c(1, 3), mar = c(4, 4, 3, 1))
hist(sims[1,],  breaks=50, freq=FALSE, col="#861F4155", border="white",
     main=expression(X[(1)]), xlab="x")
curve(dbeta(x,1,n), add=TRUE, col="#E5751F", lwd=2)

hist(sims[5,],  breaks=50, freq=FALSE, col="#861F4155", border="white",
     main=expression(X[(5)]), xlab="x")
curve(dbeta(x,5,6), add=TRUE, col="#E5751F", lwd=2)

hist(sims[10,], breaks=50, freq=FALSE, col="#861F4155", border="white",
     main=expression(X[(10)]), xlab="x")
curve(dbeta(x,n,1), add=TRUE, col="#E5751F", lwd=2)
par(mfrow=c(1,1))
```

Red curves: theoretical Beta densities. Histograms: simulation.

# Part 2: Ranks {background-color="#861F41"}

## Ranks — Definition

The **rank** of $X_i$ in a sample $X_1, \ldots, X_n$ is the number of observations $\leq X_i$:

$$\text{rank}(X_i) = n \times S_n(X_i) = \sum_{j=1}^{n} \mathbf{1}(X_j \leq X_i)$$

**Key facts:**
- Ranks are integers in $\{1, 2, \ldots, n\}$
- The $r^{\text{th}}$ order statistic has rank $r$: $\;\text{rank}(X_{(r)}) = r$
- Ties → use **mid-ranks** (average of tied positions)

::: {.callout-note}
Ranks discard the actual values — only relative ordering is retained.  
This is the foundation of **nonparametric statistics**.
:::

## Two-Sample Ranks

With $m$ observations $X_1,\ldots,X_m$ (group 1) and $n$ observations $Y_1,\ldots,Y_n$ (group 2), define ranks in the **combined sample** of $m+n$:

$$\text{rank}(Y_{(j)}) = \underbrace{\sum_{i=1}^{m} \mathbf{1}(X_i \leq Y_{(j)})}_{\text{how many } X\text{'s are} \leq Y_{(j)}} + \underbrace{j}_{\text{how many } Y\text{'s are} \leq Y_{(j)}}$$

This structure underlies the **Wilcoxon rank-sum test** and related two-sample procedures.

# Part 3: The Empirical CDF {background-color="#861F41"}

## Empirical Distribution Function (EDF)

Given i.i.d. $X_1, \ldots, X_n \sim F_X$, the **EDF** is:

$$S_n(x) = \frac{1}{n}\sum_{i=1}^{n} \mathbf{1}(X_i \leq x)$$

In terms of order statistics:

$$S_n(x) = \begin{cases} 0 & x < X_{(1)} \\ i/n & X_{(i)} \leq x < X_{(i+1)} \\ 1 & x \geq X_{(n)} \end{cases}$$

A **right-continuous step function** with jumps of $1/n$ at each observation.

## Statistical Properties of the EDF

For each fixed $x$, $\;T_n(x) = nS_n(x) \sim \text{Binomial}(n,\, F_X(x))$.

Therefore:

$$E[S_n(x)] = F_X(x) \qquad \text{(unbiasedness)}$$

$$\text{Var}[S_n(x)] = \frac{F_X(x)(1-F_X(x))}{n} \qquad \text{(consistency)}$$

As $n \to \infty$, $S_n(x) \xrightarrow{p} F_X(x)$ for each fixed $x$.

But can we say something **uniform** over all $x$?

## Glivenko–Cantelli Theorem

::: {.callout-important icon=false}
## Theorem (Glivenko–Cantelli — "Fundamental Theorem of Statistics")
$$P\!\left[\lim_{n\to\infty} \sup_{x\in\mathbb{R}} |S_n(x) - F_X(x)| = 0\right] = 1$$
:::

$S_n$ converges to $F_X$ **uniformly** over all $x$, **almost surely**.

**TL;DR:** With enough IID data, $\sup_x |S_n(x) - F_X(x)|$ is small.  
This is the key motivation for the KS test. ↓

## EDF in Action

```{r edf-demo}
#| fig-height: 3.5
#| code-fold: true
set.seed(7)
par(mfrow=c(1,2))
for(n in c(20,50)){
  x <- rnorm(n)
  plot(ecdf(x), main=paste0("n = ", n), xlab="x",
       ylab=expression(S[n](x)), col="#861F41", lwd=2, xlim=c(-3,3))
  curve(pnorm(x), add=TRUE, col="#E5751F", lwd=2, lty=2)
  legend("topleft", c("EDF","True CDF N(0,1)"),
         col=c("#861F41","#E5751F"), lwd=2, lty=c(1,2), bty="n", cex=0.8)
}
par(mfrow=c(1,1))
```

# Part 4: Quantile Transformation & RV Generation {background-color="#861F41"}

## The Probability Integral Transform

::: {.callout-important icon=false}
## Theorem (PIT / "The Golden Rule")
Let $X$ have a **continuous** CDF $F$. Then:
$$U = F(X) \sim \text{Uniform}(0,1)$$
:::

**Proof:** For $u \in (0,1)$,
$$P(F(X) \leq u) = P(X \leq F^{-1}(u)) = F(F^{-1}(u)) = u \quad \checkmark$$

**Shorthand to remember:**
$$X \sim F \;\Rightarrow\; F(X) \sim U[0,1] \;\Rightarrow\; F^{-1}(U) \sim F$$

::: {.callout-warning}
Continuity of $F$ is **essential**. For discrete $F$, $F(X)$ is stochastically larger than Uniform — this has consequences for the KS test!
:::

## The Inverse CDF Method

::: {.callout-note icon=false}
## Algorithm
1. Generate $U_1, \ldots, U_n \overset{\text{iid}}{\sim} \text{Uniform}(0,1)$
2. Set $X_i = F^{-1}(U_i)$

Then $X_1, \ldots, X_n \overset{\text{iid}}{\sim} F$
:::

**Why it works:** For any $x$,
$$P(F^{-1}(U) \leq x) = P(U \leq F(x)) = F(x) \quad \checkmark$$

**Interactive demo:**  
[https://dattahub.github.io/iqs/rvgen.html](https://dattahub.github.io/iqs/rvgen.html)

## Example: Exponential Distribution

$X \sim \text{Exp}(\lambda)$: $\quad f(x) = \lambda e^{-\lambda x}$, $\quad F(x) = 1 - e^{-\lambda x}$

**Step 1:** Solve $F(F^{-1}(u)) = u$:
$$1 - e^{-\lambda F^{-1}(u)} = u \;\Rightarrow\; F^{-1}(u) = -\frac{1}{\lambda}\log(1-u)$$

**Step 2:** Generate $U_i \sim U[0,1]$, set $X_i = -\frac{1}{\lambda}\log(1-U_i)$

```r
lambda <- 3
U <- runif(1000)
X <- -(1/lambda) * log(1 - U)
```

(R uses `rexp()` internally, which does exactly this!)

## Example: Non-Standard CDF

Consider $F(x) = 1 - \exp(-\sqrt{x})$ for $x \geq 0$.

**Derive $F^{-1}(u)$:**

$$F(F^{-1}(u)) = u$$
$$\Rightarrow 1 - \exp\!\bigl(-\sqrt{F^{-1}(u)}\bigr) = u$$
$$\Rightarrow \exp\!\bigl(-\sqrt{F^{-1}(u)}\bigr) = 1-u$$
$$\Rightarrow \boxed{F^{-1}(u) = \{\log(1/(1-u))\}^2}$$

```r
U <- runif(5000)
X <- (log(1/(1 - U)))^2
```

## Verification

```{r nonstandard-gen}
#| fig-height: 3.2
#| code-fold: true
set.seed(42)
U <- runif(5000)
X <- (log(1/(1-U)))^2
f_true <- function(x) (1/(2*sqrt(x))) * exp(-sqrt(x))

hist(X, breaks=80, freq=FALSE, col="#861F4155", border="white",
     main=expression("F(x) = 1 - exp("*-sqrt(x)*")"),
     xlab="x", xlim=c(0,20))
curve(f_true(x), from=0.01, to=20, add=TRUE, col="#E5751F", lwd=2)
legend("topright","True density", col="#E5751F", lwd=2, bty="n")
```

## Special Case: Generating Normal Variables

$\Phi(x)$ (standard Normal CDF) has **no closed-form inverse** — can't apply inverse CDF directly.

**Box–Müller Transform** uses polar coordinates:

If $(X_1, X_2) \sim N(0,I_2)$ with polar coordinates $(R, \Theta)$:
$$R^2 = X_1^2 + X_2^2 \sim \chi^2_2 \equiv \text{Exp}(1/2), \qquad \Theta \sim \text{Uniform}(0, 2\pi)$$

::: {.callout-note icon=false}
## Algorithm (Box–Müller)
1. Generate $U_1, U_2 \overset{\text{iid}}{\sim} U[0,1]$
2. $X_1 = \sqrt{-2\log U_1}\,\cos(2\pi U_2)$
3. $X_2 = \sqrt{-2\log U_1}\,\sin(2\pi U_2)$

Then $X_1, X_2 \overset{\text{iid}}{\sim} N(0,1)$.
:::

# Part 5: The Kolmogorov–Smirnov Test {background-color="#861F41"}

## Setup: What Are We Testing?

We observe $X_1, \ldots, X_n$ and want to test:
$$H_0: F_X(x) = F_0(x) \quad \text{for all } x \in \mathbb{R}$$
where $F_0$ is a **completely specified** continuous CDF.

**Idea:** By Glivenko–Cantelli, $S_n \approx F_X$.  
Under $H_0$, $S_n \approx F_0$, so $\sup_x |S_n(x) - F_0(x)|$ should be **small**.

**Large values of this supremum → evidence against $H_0$.**

## The KS Statistics

$$D_n^+ = \sup_{x}\, [S_n(x) - F_0(x)] \qquad \text{(upper one-sided)}$$

$$D_n^- = \sup_{x}\, [F_0(x) - S_n(x)] \qquad \text{(lower one-sided)}$$

$$\boxed{D_n = \sup_{x}\, |S_n(x) - F_0(x)| = \max(D_n^+,\, D_n^-)}$$

$D_n$ is the **two-sided KS statistic** — the one we almost always use.

Reject $H_0$ when $D_n$ is **large** (exceeds the critical value).

## Remarkable Property: Distribution-Free!

::: {.callout-important icon=false}
## Theorem
Under $H_0$, the distributions of $D_n$, $D_n^+$, $D_n^-$ are **independent of $F_0$**, provided $F_0$ is continuous.
:::

**Is this surprising?** The test statistic measures how far $S_n$ is from $F_0$ — yet its distribution doesn't depend on $F_0$ at all!

To understand why, we need the PIT + a careful look at how $D_n^+$ reduces...

## Proof: $D_n^+$ is Distribution-Free {.smaller}

Using the step-function form of $S_n$:

$$D_n^+ = \sup_{x} [S_n(x) - F_0(x)]$$

$$= \max_{0\leq i\leq n} \left[\sup_{X_{(i)}\leq x < X_{(i+1)}} \left\{\frac{i}{n} - F_0(x)\right\}\right]$$

$$= \max_{0\leq i\leq n} \left[\frac{i}{n} - \inf_{X_{(i)}\leq x < X_{(i+1)}} F_0(x)\right]$$

$$= \max_{0\leq i\leq n} \left[\frac{i}{n} - F_0(X_{(i)})\right]$$

Since $X_{(i)}$ are order stats from $F_0$, by the **PIT**, $F_0(X_{(i)})$ behave as order stats from $U[0,1]$ — **regardless of $F_0$**! $\square$

## Two Critical Consequences

From the proof:

::: {.incremental}
1. **Distribution-free** ✓  
   The same critical value table works for **any** continuous null distribution.

2. **$F_0$ must be continuous**  
   The PIT requires continuity. Applying KS to discrete data makes the test **overly conservative**.

3. **$F_0$ must be completely known**  
   We need to evaluate $F_0(X_{(i)})$ exactly. If parameters are estimated from the data, the critical values are **wrong**.
:::

## Computing $D_n$ in Practice

**Convenient formula:**

$$D_n = \max_{i=1,\ldots,n}\left\{ \left|\frac{i}{n} - F_0(X_{(i)})\right|,\; \left|\frac{i-1}{n} - F_0(X_{(i)})\right| \right\}$$

The two terms capture the jump from both sides of the step function.

**In R:**
```r
ks.test(x, "pnorm", mean = 0, sd = 1)   # test against N(0,1)
ks.test(x, "pexp",  rate = 2)            # test against Exp(2)
```

## Worked Example {.smaller}

20 values from $U(0,1)$. Test: do their **square roots** also follow $U(0,1)$?

Under $H_0$: $F_0(x) = x$.

```{r ks-example}
x_obs <- c(0.0123, 0.1039, 0.1954, 0.2621, 0.2802,
           0.3217, 0.3645, 0.3919, 0.4240, 0.4814,
           0.5139, 0.5846, 0.6275, 0.6541, 0.6889,
           0.7621, 0.8320, 0.8871, 0.9249, 0.9634)
y <- sqrt(x_obs)
ks.test(y, "punif")
```

$D_{20} = 0.36 > 0.352$ (critical value at $\alpha=0.01$) → **Reject $H_0$**

## Visualising the KS Statistic

```{r ks-plot}
#| fig-height: 3.5
#| code-fold: true
x_obs <- c(0.0123,0.1039,0.1954,0.2621,0.2802,0.3217,0.3645,
           0.3919,0.4240,0.4814,0.5139,0.5846,0.6275,0.6541,
           0.6889,0.7621,0.8320,0.8871,0.9249,0.9634)
y <- sqrt(x_obs)

plot(ecdf(y), col="#861F41", lwd=2,
     main=expression("EDF of"~sqrt(x)~"vs Uniform(0,1)"),
     xlab="x", ylab="CDF")
abline(0, 1, col="#E5751F", lwd=2, lty=2)

# Mark the maximum gap
i_max <- which.max(abs((1:20)/20 - y))
segments(y[i_max], (i_max-1)/20, y[i_max], y[i_max],
         col="#003366", lwd=3)
text(y[i_max]+0.05, ((i_max-1)/20 + y[i_max])/2,
     expression(D[n]~"= 0.36"), col="#003366", cex=0.9)

legend("topleft", c("EDF","Uniform CDF"),
       col=c("#861F41","#E5751F"), lwd=2, lty=c(1,2), bty="n")
```

## Asymptotic Distribution

::: {.callout-note icon=false}
## Theorem (Kolmogorov 1933, Smirnov 1939)
For any continuous $F_X$ and $d > 0$:
$$\lim_{n\to\infty} P\!\left(D_n \leq \frac{d}{\sqrt{n}}\right) = L(d) = 1 - \sum_{i=1}^{\infty}(-1)^{i-1} e^{-2i^2 d^2}$$
:::

The approximation is accurate for $n \geq 35$.

**For $n > 40$**, critical values at common $\alpha$:

| $\alpha$ | 0.20 | 0.10 | 0.05 | 0.02 | 0.01 |
|:--------:|:----:|:----:|:----:|:----:|:----:|
| $c_\alpha / \sqrt{n}$ | $1.07/\sqrt{n}$ | $1.22/\sqrt{n}$ | $1.36/\sqrt{n}$ | $1.52/\sqrt{n}$ | $1.63/\sqrt{n}$ |

# Part 6: The Lilliefors Test {background-color="#861F41"}

## The Problem With Estimated Parameters

Suppose data come from $N(\mu, \sigma^2)$ with **unknown** $\mu, \sigma$.

**Tempting idea:** plug in $\hat\mu = \bar X$, $\hat\sigma = S$, run KS against $N(\hat\mu, \hat\sigma^2)$.

**Why this fails:**
- The fitted distribution is forced closer to the data
- The KS statistic is systematically **smaller** than under a truly specified $F_0$
- Standard KS critical values are **too large** → test is **conservative**
- Actual type I error $\ll$ nominal $\alpha$

## Lilliefors' Fix

**Lilliefors (1967)** derived corrected critical values by simulation specifically for the case of testing normality with estimated parameters.

::: {.callout-note icon=false}
## Lilliefors Test Procedure
1. Compute $\hat\mu = \bar X$, $\hat\sigma = S$
2. Form the KS statistic $D_n$ against $N(\hat\mu, \hat\sigma^2)$
3. Compare to **Lilliefors critical values** (smaller than standard KS values)
:::

Extensions exist for testing exponentiality, uniformity, etc., with estimated parameters.

## KS vs Lilliefors: When To Use Each

| Scenario | Test |
|----------|------|
| $H_0$: $F = F_0$ fully specified | **Kolmogorov–Smirnov** |
| $H_0$: $F$ is Normal, $\mu,\sigma$ unknown | **Lilliefors** |
| $H_0$: $F$ is Exponential, $\lambda$ unknown | **Lilliefors (exp variant)** |
| Discrete data | **Neither** — use $\chi^2$ GoF |

## IQ Score Example

IQ scores of 18 girls:  
`114, 81, 87, 114, 113, 87, 111, 89, 93, 108, 99, 93, 100, 95, 93, 95, 106, 108`

**(a)** Test normality with **unknown** $\mu, \sigma$ → Lilliefors

**(b)** Test normality with $\mu = 100$, $\sigma^2 = 100$ → KS

**Why?**  
In (a), parameters must be estimated from the data → Lilliefors.  
In (b), $F_0 = N(100, 10^2)$ is completely specified → KS.

## IQ Example in R

```{r iq-r}
if (!requireNamespace("nortest", quietly=TRUE)) install.packages("nortest")
library(nortest)

iq <- c(114, 81, 87, 114, 113, 87, 111, 89, 93, 108,
        99, 93, 100, 95, 93, 95, 106, 108)

# (a) Lilliefors
lillie.test(iq)

# (b) KS with specified parameters
ks.test(iq, "pnorm", mean=100, sd=10)
```

# Part 7: Other Goodness-of-Fit Tests {background-color="#861F41"}

## The Chi-Squared GoF Test

For **discrete** distributions or **binned continuous** data:

$$\chi^2 = \sum_{i=1}^{k} \frac{(O_i - E_i)^2}{E_i} \xrightarrow{d} \chi^2_{k-1-m}$$

where $E_i = np_i^{(0)}$ (expected counts), $m$ = number of estimated parameters.

**Rules of thumb:**
- Each $E_i \geq 5$ (merge cells if needed)
- Result depends on binning choice — a known weakness
- Works for **discrete** distributions where KS is inappropriate

```r
chisq.test(table(observed_counts))
```

## The Anderson–Darling Test

**Weighted** version of the KS test giving more attention to the tails:

$$A^2 = -n - \sum_{i=1}^{n} \frac{2i-1}{n}\left[\log F_0(X_{(i)}) + \log(1-F_0(X_{(n+1-i)}))\right]$$

**Advantages over KS:**
- More sensitive to **tail departures**
- Uniformly more powerful for many distributions
- Corrected versions for estimated parameters (normality, exponentiality, etc.)

```{r}
library(nortest)
ad.test(rnorm(20))   # Anderson-Darling normality test
```


## The Cramér–von Mises Test

Measures the **integrated squared** difference between $S_n$ and $F_0$:

$$W^2 = n\int_{-\infty}^{\infty}(S_n(x)-F_0(x))^2\,dF_0(x) = \frac{1}{12n} + \sum_{i=1}^{n}\!\left(F_0(X_{(i)}) - \frac{2i-1}{2n}\right)^2$$

- Distribution-free (like KS) when $F_0$ is continuous and fully specified
- Weights all parts of the distribution **equally** (unlike Anderson–Darling)

```{r}
library(goftest)
cvm.test(rnorm(20), "pnorm")
```




## The Shapiro–Wilk Test

**Normality-specific** — most powerful test for normality at small/moderate $n$.

$$W = \frac{\left(\sum_{i=1}^{n} a_i X_{(i)}\right)^2}{\sum_{i=1}^{n}(X_i - \bar{X})^2}$$

The $a_i$ coefficients come from the expected values and covariance matrix of standard normal order statistics.

- $W \in (0, 1]$; values close to 1 → consistent with normality
- Recommended for $n \leq 2000$

```{r}
shapiro.test(rnorm(20))
```


## Comparing All the Tests

| Test | Null $H_0$ | Continuous? | Params known? | Tail focus |
|------|-----------|:-----------:|:-------------:|:----------:|
| $\chi^2$ GoF | Any (binned) | No | Yes/No | Low |
| KS | Specific $F_0$ | **Yes** | **Yes** | Middle |
| Lilliefors | Normal/Exp family | Yes | **No** | Middle |
| Cramér–von Mises | Specific $F_0$ | Yes | Yes | Uniform |
| Anderson–Darling | Specific $F_0$ | Yes | Yes | **High** |
| Shapiro–Wilk | **Normality only** | Yes | No | **High** |

## Power Comparison in R

```{r power-compare}
#| code-fold: true
#| fig-height: 3.5
library(nortest)
set.seed(42)
x_t3 <- rt(50, df = 5)   # heavy-tailed, not normal

results <- data.frame(
  Test = c("KS (Lilliefors)", "Anderson-Darling", "Shapiro-Wilk", "Cramer-von Mises"),
  p_value = c(
    lillie.test(x_t3)$p.value,
    ad.test(x_t3)$p.value,
    shapiro.test(x_t3)$p.value,
    if(requireNamespace("goftest",quietly=TRUE))
      goftest::cvm.test(x_t3,"pnorm",mean=mean(x_t3),sd=sd(x_t3))$p.value else NA
  )
)

barplot(results$p_value, names.arg = results$Test,
        col = ifelse(results$p_value < 0.05, "#861F41", "#C0C0C0"),
        main = expression("p-values for H"[0]*": t"[5]*" data is Normal"),
        ylab = "p-value", las = 2, cex.names = 0.75)
abline(h = 0.05, col = "#E5751F", lwd = 2, lty = 2)
text(0.3, 0.07, "α = 0.05", col = "#E5751F", cex = 0.8)
```

# Summary {background-color="#861F41"}

## The Big Picture

::: {.incremental}
1. **Order statistics** $X_{(1)} \leq \cdots \leq X_{(n)}$ have known distributions (Beta family for Uniform parent)

2. **Ranks** strip values to relative order — the basis of nonparametric inference

3. **EDF** $S_n(x)$ estimates $F_X$ nonparametrically; **Glivenko–Cantelli** guarantees uniform convergence

4. **Probability Integral Transform** ($X \sim F \Rightarrow F(X) \sim U[0,1]$) powers both RV generation and the KS distribution-free property

5. **KS test** $D_n = \sup_x|S_n(x)-F_0(x)|$ requires $F_0$ continuous and **fully specified**

6. **Lilliefors** corrects KS when parameters are estimated; **Anderson–Darling** and **Shapiro–Wilk** offer greater power in specific settings
:::

## Practical Decision Tree

```
Is H₀ fully specified?
├── Yes → Is F₀ continuous?
│         ├── Yes → KS test (or CvM / AD for more power)
│         └── No  → χ² goodness-of-fit
└── No  → What family?
          ├── Normal → Lilliefors (or Shapiro-Wilk)
          └── Other  → Lilliefors variant / parametric bootstrap
```

## Exercises

1. Derive the density of $X_{(3)}$ for $n=10$, $X_i \sim \text{Exp}(2)$. Verify by simulation.

2. For $F(x) = 1 - (\beta/x)^\alpha$ (Pareto), derive $F^{-1}(u)$ and write R code to generate samples.

3. Given 10 observations (0.12, 0.28, …, 0.95), test $H_0: F = U(0,1)$ by computing $D_{10}$ by hand.

4. Why does plugging in $\hat\mu, \hat\sigma$ into the KS test make it **conservative**? Which direction does the size deviate from $\alpha$?

5. Generate 30 observations from $t_3$. Apply KS, Lilliefors, AD, and Shapiro–Wilk for normality. Which detects non-normality first?

---

## References and Further Reading

- Gibbons, J.D. & Chakraborti, S. (2011). *Nonparametric Statistical Inference*, 5th ed. CRC Press.
- Kolmogorov, A.N. (1933). Sulla determinazione empirica di una legge di distribuzione. *Giornale dell'Istituto Italiano degli Attuari*, 4, 83–91.
- Lilliefors, H.W. (1967). On the Kolmogorov-Smirnov test for normality with mean and variance unknown. *JASA*, 62, 399–402.
- R packages: `nortest`, `goftest`, `KSgeneral`
- **Interactive demo:** [https://dattahub.github.io/iqs/rvgen.html](https://dattahub.github.io/iqs/rvgen.html)
