---
title: "Lecture 7: Wilcoxon Signed-Rank Test and Q-Q Plots"
subtitle: "Nonparametric Tests for Location and Goodness of Fit"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    fig-align: center
    html-math-method: mathjax
    lang: en
    fig-cap-location: bottom
  pdf:
    toc: true
    number-sections: true
    fig-align: center
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
set.seed(42)
```

# The Wilcoxon Signed-Rank Test

The Wilcoxon signed-rank test is a test for a centrality parameter (location). It is a nonparametric alternative to both the sign test and the parametric z-test/t-test, applicable to both single-sample and paired-sample problems.

---

## Review: Sign Test

### Situation

A sample of size $n$, $(x_1, x_2, \ldots, x_n)$, is drawn from an unknown distribution and we wish to test
$$H_0: \theta = \theta_0 \quad \text{vs} \quad H_A: \theta \neq \theta_0 \quad \text{(two-sided, or one-sided)},$$
where $\theta$ is the center (median) of the distribution.

For the **sign test**, we count $S$, the number of positive values of $(x_1 - \theta_0, \ldots, x_n - \theta_0)$. We reject $H_0$ if $S$ is not close to $n/2$.

### Steps of the Sign Test

1. Compute $S$ = the number of observations greater than $\theta_0$.
2. Let $s_\text{obs}$ be the observed value of $S$.
3. Compute the p-value:
   - One-sided: $P[S \geq s_\text{obs}]$
   - Two-sided: $2 \times P[S \geq s_\text{obs}]$
4. Conclude $H_A$ (reject $H_0$) if p-value $< 0.05$ (or $0.01$).

In R: use `binom.test()` with $p = 1/2$ and $n$ = sample size. For large $n$, use the CLT approximation via `pnorm()`.

### Drawback of the Sign Test

The sign test **completely ignores the magnitudes** of the observations — only the signs of $(x_i - \theta_0)$ are used. This discards information in the data and can lead to **lack of power**.

> **Power** $= 1 - P(\text{Type II error})$ = the ability to detect a departure from the null when $H_0$ is false.

**Question:** How do we improve on the sign test?

**Wilcoxon's answer:** Use the *ranks* of the observations (based on their absolute values), along with the sign.

---

## Wilcoxon Signed-Rank Test

### Setup and Assumptions

**Basic assumptions:**

- $X_1, \ldots, X_n$ are independently and identically distributed, continuous, and **symmetric** about the common median $\theta$.
- Test $H_0: \theta = \theta_0$ versus $H_a: \theta \neq \theta_0$, where $\theta_0$ is a prespecified hypothesized value.

The Binomial (sign) test uses only the signs of the differences $D_i = X_i - \theta_0$. The Wilcoxon test uses both the signs *and* the ranks of $|D_i|$.

### Constructing the Test Statistic

**Step 1.** Define $D_i = X_i - \theta_0$ and rank the absolute values $|D_i|$.

**Step 2.** Define $R_i$ as the rank of $|D_i|$ (rank 1 = smallest, rank $n$ = largest).

**Step 3.** Define the indicator
$$S_i = \begin{cases} 1 & \text{if } D_i > 0 \\ 0 & \text{otherwise.}\end{cases}$$

**Step 4.** Define the **Wilcoxon signed-rank statistic**:
$$SR^+ = \sum_{i=1}^{n} S_i R_i,$$
the sum of ranks associated with positive differences.

Similarly, $SR^-$ is the sum of ranks associated with negative differences.

**Key identity:**
$$SR^+ + SR^- = 1 + 2 + 3 + \cdots + n = \frac{n(n+1)}{2}$$
so knowing one determines the other.

### Behaviour Under $H_0$ and $H_A$

- Under $H_0$: $SR^+ \approx SR^- \approx \dfrac{n(n+1)}{4}$, since each rank is equally likely to be positive or negative.
- Under $H_A$: Either $SR^+$ will be large ($SR^-$ small), or $SR^+$ will be small ($SR^-$ large).

**Rejection rules:**

- Two-sided $H_a: \theta \neq \theta_0$: reject $H_0$ if $SR^+$ is either too large or too small.
- Upper-tailed $H_a: \theta > \theta_0$: reject $H_0$ if $SR^+$ is too large.
- Lower-tailed $H_a: \theta < \theta_0$: reject $H_0$ if $SR^+$ is too small.

**Ties.** The continuity assumption implies no ties in theory, but in practice ties occur. When they do, replace ranks with **mid-ranks** of the tied observations. Discard any $D_i = 0$ and recalculate ranks on the remainder.

---

## Example: Hypnotic Susceptibility (Cooper et al., 1967)

**Objective:** Test whether hypnotic susceptibility can be increased with training (paired replicate data).

$H_0$: no change in hypnotic susceptibility vs. $H_A$: susceptibility can be increased (one-sided, upper).

```{r}
#| label: tbl-hypnosis
#| tbl-cap: "Average Scores of Hypnotic Susceptibility (before and after training)."

hypnosis <- data.frame(
  Subject  = 1:6,
  Before   = c(10.5, 19.5, 7.5, 4.0, 4.5, 2.0),
  After    = c(18.5, 24.5, 11.0, 2.5, 5.5, 3.5)
)
hypnosis$D  <- hypnosis$After - hypnosis$Before
hypnosis$R  <- rank(abs(hypnosis$D))
hypnosis$S  <- as.integer(hypnosis$D > 0)
knitr::kable(hypnosis, col.names = c("Subject", "$X_i$ (before)", "$Y_i$ (after)",
                                      "$D_i = Y_i - X_i$", "$R_i$", "$S_i$"),
             align = "crrrrr")
```

```{r}
#| label: hypnosis-test

D <- c(8, 5, 3.5, -1.5, 1, 1.5)
SR_plus <- sum(rank(abs(D))[D > 0])
cat("SR+ =", SR_plus, "\n")

wilcox.test(D, alternative = "greater")
```

The `V = 18.5` reported by R is $SR^+$. Since the p-value $\approx 0.057 > 0.05$, we do not quite reject $H_0$ at the 5% level.

---

## Null Distribution of the Signed-Rank Statistic

### Exact Null Distribution

Under $H_0$, each sign attached to any rank is equally likely to be $+$ or $-$ (probability $1/2$ each), independently. For sample size $n$, there are $2^n$ equally probable sign assignments.

### Example: $n = 4$

With $n = 4$ there are $2^4 = 16$ equally probable sign assignments:

```{r}
#| label: tbl-null-dist-n4
#| tbl-cap: "All $2^4 = 16$ sign assignments for $n = 4$, with $SR^+$, $SR^-$, and probability $1/16$."

signs <- expand.grid(s1 = c(-1,1), s2 = c(-1,1),
                     s3 = c(-1,1), s4 = c(-1,1))
signs$SR_plus  <- apply(signs, 1, function(s) sum((1:4)[s > 0]))
signs$SR_minus <- apply(signs, 1, function(s) sum((1:4)[s < 0]))
signs$Prob <- "1/16"

tab_display <- signs
tab_display[,1:4] <- ifelse(signs[,1:4] > 0, "+", "-")
knitr::kable(tab_display,
             col.names = c("$s_1$","$s_2$","$s_3$","$s_4$",
                           "$SR^+$","$SR^-$","Prob"),
             align = "ccccccc")
```

### PMF and CDF of $SR^+$ for $n = 4$

```{r}
#| label: tbl-pmf-cdf-n4
#| tbl-cap: "PMF and CDF of $SR^+$ for $n = 4$, verified against `psignrank()`."

sr_vals <- 0:10
pmf <- table(factor(signs$SR_plus, levels = sr_vals)) / 16
cdf <- cumsum(pmf)

dist_tab <- data.frame(
  T     = sr_vals,
  PMF   = as.numeric(pmf),
  CDF   = as.numeric(cdf),
  R_CDF = psignrank(sr_vals, n = 4)
)
knitr::kable(dist_tab,
             col.names = c("$t$", "$P[T = t]$", "$P[T \\leq t]$ (manual)",
                           "$P[T \\leq t]$ (`psignrank`)"),
             digits = 4, align = "crrrr")
```

The CDF values match `psignrank(0:10, n = 4)` exactly. The size of the enumeration table grows as $2^n$, so for larger $n$ we use R or the normal approximation.

```{r}
#| label: fig-null-dist
#| fig-cap: "PMF (left) and CDF (right) of $SR^+$ under $H_0$ for $n = 4$, computed by complete enumeration."
#| fig-alt: "Left: bar chart showing the probability mass function of SR+ for n=4, symmetric around 5. Right: step function showing the CDF."
#| fig-width: 8
#| fig-height: 4

par(mfrow = c(1, 2), mar = c(4.5, 4.5, 3, 1))

barplot(pmf, names.arg = sr_vals,
        xlab = expression(SR^"+"), ylab = "Probability",
        main = expression("PMF of " * SR^"+" * ", n = 4"),
        col = "#aec7e8", border = "white", las = 1)

plot(sr_vals, cdf, type = "s", lwd = 2, col = "#1f77b4",
     xlab = expression(SR^"+"), ylab = expression(P(SR^"+" <= t)),
     main = expression("CDF of " * SR^"+" * ", n = 4"),
     las = 1)
points(sr_vals, cdf, pch = 16, col = "#1f77b4")

par(mfrow = c(1, 1))
```

---

## CLT Approximation for Large $n$

For $n > 12$, the statistic $T$ ($W^+$ or $W^-$) is approximately normal:
$$\mu_T = \frac{n(n+1)}{4}, \qquad \sigma_T = \sqrt{\frac{n(n+1)(2n+1)}{24}}.$$

### Example: Cystic Fibrosis FVC Study

**Context.** FVC (Forced Vital Capacity) = volume of air expelled from the lungs in 6 seconds; reduced over time in cystic fibrosis patients.

**Research question:** Is the reduction in FVC less when a new experimental drug is administered?

**Design.** Matched-pair design: pairs of patients matched on initial FVC. One receives the drug, one a placebo. Compute $x_i = \text{Reduction (placebo)} - \text{Reduction (drug)}$. Large $SR^+$ (small $SR^-$) supports the drug being effective.

```{r}
#| label: tbl-fvc
#| tbl-cap: "Reduction in forced vital capacity (FVC) for a matched-pair sample of cystic fibrosis patients."

fvc <- data.frame(
  Subject     = 1:14,
  Placebo     = c(224, 8, 75, 541, 74, 85, 293, -23, 525, -38, 508, 255, 525, 1023),
  Drug        = c(213, 95, 33, 440, -32, -28, 445, -178, 367, 140, 323, 10, 65, 343)
)
fvc$xi    <- fvc$Placebo - fvc$Drug
fvc$Rank  <- rank(abs(fvc$xi))
fvc$SR_pos <- ifelse(fvc$xi > 0, fvc$Rank, NA)
fvc$SR_neg <- ifelse(fvc$xi < 0, fvc$Rank, NA)

knitr::kable(fvc,
             col.names = c("Subject","Placebo","Drug","$x_i$","Rank","$SR^+$","$SR^-$"),
             na.rm = FALSE, align = "crrrrrr")
```

```{r}
#| label: fvc-clt

n  <- 14
SR_plus  <- sum(fvc$Rank[fvc$xi > 0])
SR_minus <- sum(fvc$Rank[fvc$xi < 0])
cat(sprintf("SR+ = %g,  SR- = %g\n", SR_plus, SR_minus))

mu_T    <- n * (n + 1) / 4
sigma_T <- sqrt(n * (n + 1) * (2*n + 1) / 24)
cat(sprintf("Mean = n(n+1)/4 = %g\n", mu_T))
cat(sprintf("SD   = sqrt(n(n+1)(2n+1)/24) = %.5f\n", sigma_T))

# CLT p-value (using SR-)
z_stat  <- (SR_minus - mu_T) / sigma_T
p_clt   <- pnorm(z_stat)
cat(sprintf("z = (SR- - mu)/sigma = (19 - %.1f) / %.5f = %.4f\n",
            mu_T, sigma_T, z_stat))
cat(sprintf("P-value (CLT) = P[Z <= %.4f] = %.4f\n", z_stat, p_clt))
```

Since the p-value $< 0.05$, we conclude the drug is effective in reducing the deterioration of FVC.

```{r}
#| label: fvc-exact

# Exact test using R
wilcox.test(fvc$xi, alternative = "greater")
```

---

## Summary: Carrying Out the Wilcoxon Signed-Rank Test

1. Compute $T = SR^+$ or $SR^-$ (often take the smaller of the two).
2. Let $t_\text{obs}$ be the observed value of $T$.
3. Compute the p-value:
   - $P[T \leq t_\text{obs}]$ (one-tailed), or $2 P[T \leq t_\text{obs}]$ (two-tailed).
4. For $n \leq 12$: use the exact distribution table or `psignrank()`.
5. For $n > 12$: use the normal approximation above.
6. In R (any sample size): `wilcox.test(x, alternative = "greater"/"less"/"two.sided")`.
7. Conclude $H_A$ (reject $H_0$) if p-value $< 0.05$ (or $0.01$).

---

## Comparison of Tests

```{r}
#| label: tbl-comparison
#| tbl-cap: "Assumptions and properties of the sign test, Wilcoxon signed-rank test, and t-test."

comp <- data.frame(
  Method = c("Sign test", "Wilcoxon signed-rank test", "t-test / z-test"),
  Assumptions = c(
    "i.i.d., any continuous distribution",
    "i.i.d., continuous and **symmetric** distribution",
    "i.i.d., **normal** distribution"
  ),
  Notes = c(
    "Weakest assumption; ignores magnitude entirely",
    "Uses ranks — more powerful than sign test when symmetry holds",
    "Most powerful when normality holds (UMP test)"
  )
)
knitr::kable(comp, align = "lll")
```

- The **sign test** has the weakest assumptions and is always valid for continuous data, but is the least powerful.
- The **t-test** is most powerful under normality (uniformly most powerful for normal location families), but the Type I error rate may be distorted if normality fails.
- The **Wilcoxon test** sits between them: more powerful than the sign test when symmetry holds; no normality required.

> **How restrictive is symmetry?** Fairly restrictive. A serious violation invalidates the Wilcoxon signed-rank test, but the sign test remains valid. In practice, you must assess symmetry (e.g., via a Q-Q plot) before choosing.

---

## Handling Ties

Real data almost always contain ties due to rounding or measurement limitations. The standard procedure is to use **mid-ranks**:

```{r}
#| label: ties-example

# Sample with hypothesized median theta_0 = 3
x      <- c(1, 1, 5, 5, 8, 8, 8)
theta0 <- 3
d      <- x - theta0
cat("Differences d_i:", d, "\n")
cat("Mid-ranks of |d_i|:", rank(abs(d)), "\n")
```

The differences $d_i = x_i - \theta_0$ are $-2, -2, 2, 2, 5, 5, 5$.

- Ranks 1–4 are tied at $|d| = 2$ → mid-rank = 2.5
- Ranks 5–7 are tied at $|d| = 5$ → mid-rank = 6

```{r}
#| label: ties-wilcox

suppressWarnings(wilcox.test(x, mu = theta0, conf.int = TRUE))
```

R automatically applies mid-ranks and uses a continuity correction when ties are present.

---

## Univariate Example: Water Content Data

```{r}
#| label: water-example

x2 <- c(5.6, 6.1, 6.3, 6.4, 6.5, 6.6, 7.0, 7.5, 7.9,
        8.0, 8.0, 8.1, 8.1, 8.2, 8.4, 8.5, 8.7, 9.4, 14.3, 26.0)

# Wilcoxon signed-rank test: H0: theta = 9 vs H1: theta > 9
suppressWarnings(wilcox.test(x2, mu = 9, alternative = "greater", conf.int = TRUE))
```

```{r}
#| label: water-alternatives

# Sign test
binom.test(sum(x2 > 9), n = length(x2), p = 0.5, alternative = "greater")

# t-test
t.test(x2, mu = 9, alternative = "greater")
```

Notice the discrepancy: the Wilcoxon and sign tests reject $H_0$, but the t-test does not (p-value $\approx 0.827$). This is because the data contain extreme outliers (14.3, 26.0) that inflate $s^2$ and render the t-test insensitive here — but violate the symmetry assumption needed for the Wilcoxon test too. The sign test, requiring only continuity, is the most defensible choice for this dataset.

---

# Quantiles and Q-Q Plots

## Quantiles

The **quantile** $Q(p)$ corresponding to probability $p \in (0,1)$ is defined as:
$$Q(p) = F^{-1}(p) = \inf\{x : F(x) \geq p\}.$$

- **Population (true) CDF:** $F(x) = P(X \leq x)$
- **Empirical CDF:** $F_n(x) = \dfrac{\#\{X_i \leq x\}}{n}$

$F_n$ is an estimator of $F$ (just as $\bar{X}$ estimates $\mu$). In fact, there is a stronger result: $\sup_x |F_n(x) - F(x)| \to 0$ uniformly (Glivenko–Cantelli theorem), and the scaled deviation $\sqrt{n}\sup_x |F_n(x) - F(x)|$ converges to the Kolmogorov distribution.

```{r}
#| label: fig-ecdf
#| fig-cap: "Empirical CDF (step function) vs. true normal CDF (smooth curve) for $n = 20$ observations. The right panel shows the 90th percentile read off from the empirical CDF."
#| fig-alt: "Two side-by-side plots. Left: step-function ECDF overlaid with smooth normal CDF. Right: same plot with a horizontal line at y=0.9 and a vertical red line indicating the empirical 90th percentile."
#| fig-width: 9
#| fig-height: 4.5

set.seed(2025)
n_ecdf <- 20
x_samp <- rnorm(n_ecdf)
Fn     <- ecdf(x_samp)
x_grid <- seq(-3, 3, length.out = 300)

par(mfrow = c(1, 2), mar = c(4.5, 4.5, 3, 1))

# Left: ECDF vs true CDF
plot(Fn, main = "Empirical and True Distribution",
     xlab = "x", ylab = expression(F[n](x)),
     col = "black", lwd = 1.5, las = 1,
     xlim = c(-3, 3), do.points = TRUE, pch = 16, cex = 0.6)
lines(x_grid, pnorm(x_grid), col = "#d62728", lwd = 1.5)
abline(h = c(0, 1), lty = 2, col = "grey70")
legend("topleft", legend = c("ECDF", "True CDF"),
       col = c("black","#d62728"), lwd = 1.5, bty = "n", cex = 0.85)

# Right: reading off the 90th percentile
q90 <- quantile(x_samp, 0.9)
plot(Fn, main = "Reading Off a Quantile",
     xlab = "x", ylab = expression(F[n](x)),
     col = "black", lwd = 1.5, las = 1,
     xlim = c(-3, 3), do.points = TRUE, pch = 16, cex = 0.6)
lines(x_grid, pnorm(x_grid), col = "#d62728", lwd = 1.5)
abline(h = 0.9, lty = 1, col = "grey50")
abline(v = q90, col = "#d62728", lwd = 1.5)
abline(h = c(0, 1), lty = 2, col = "grey70")
text(q90 + 0.1, 0.15, sprintf("Q(0.9) = %.2f", q90),
     col = "#d62728", adj = 0, cex = 0.85)

par(mfrow = c(1, 1))
```

In R: `qnorm()` for normal quantiles, `quantile()` for sample quantiles.

---

## Q-Q Plots

For a sample $x_1, \ldots, x_n$, a **Q-Q plot** computes sample quantiles and plots them against either:

(a) **theoretical quantiles** of a hypothesized distribution, or
(b) **quantiles of a second sample**.

If the points lie approximately along the diagonal ($y = x$ line), this is consistent with the sample following the hypothesized distribution, or the two samples coming from the same distribution.

```{r}
#| label: fig-qqplots
#| fig-cap: "Three Q-Q plots. Left: 100 samples from $\\mathcal{N}(0,1)$ plotted against theoretical normal quantiles — near-perfect diagonal. Centre: two independent samples from $\\mathcal{N}(0,1)$ plotted against each other — also near-diagonal. Right: one sample from $\\mathcal{N}(0,1)$ vs one from a Cauchy distribution — heavy tails cause extreme deviations from the diagonal."
#| fig-alt: "Three Q-Q scatter plots side by side. Left panel shows points tightly following the diagonal reference line. Middle panel similarly follows the diagonal. Right panel shows severe S-shaped deviation from the diagonal, indicating heavy-tailed Cauchy data compared to normal."
#| fig-width: 10
#| fig-height: 4

set.seed(101)
n_qq <- 100
s1   <- rnorm(n_qq)
s2   <- rnorm(n_qq)
s3   <- rcauchy(n_qq)

par(mfrow = c(1, 3), mar = c(4.5, 4.5, 3.5, 1))

# Plot 1: normal sample vs theoretical normal quantiles
qqnorm(s1, main = "Normal Q-Q Plot",
       xlab = "Theoretical Quantiles", ylab = "Sample Quantiles",
       pch = 1, cex = 0.7)
qqline(s1, col = "#d62728", lwd = 1.5)

# Plot 2: two N(0,1) samples
qqplot(s1, s2,
       main = "Two-sample Q-Q Plot\n(both N(0,1))",
       xlab = "Sample 1 Quantiles", ylab = "Sample 2 Quantiles",
       pch = 1, cex = 0.7, col = "#1f77b4")
abline(0, 1, col = "#d62728", lwd = 1.5)

# Plot 3: N(0,1) vs Cauchy
qqplot(s1, s3,
       main = "Two-sample Q-Q Plot\n(N(0,1) vs Cauchy)",
       xlab = "Sample 1 Quantiles (Normal)", ylab = "Sample 2 Quantiles (Cauchy)",
       pch = 1, cex = 0.7, col = "#1f77b4")
abline(0, 1, col = "#d62728", lwd = 1.5)

par(mfrow = c(1, 1))
```

Note the pronounced deviation in the right panel: the Cauchy distribution is notoriously heavy-tailed, so its quantiles diverge sharply from the normal reference line.

> See also: [Wikipedia — Normality test](https://en.wikipedia.org/wiki/Normality_test)

---

## Next: Goodness-of-Fit Tests

Q-Q plots are a visual tool. The next module covers formal **goodness-of-fit tests** — how to test whether your data follow a given distribution (both discrete and continuous cases).
