---
title: "Lecture: One-Sample t-Test"
subtitle: "From z-Tests to Student's t"
author: "Jyotishka Datta"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    code-fold: false
    fig-align: center
    html-math-method: mathjax
    lang: en
    # Accessibility
    fig-cap-location: bottom
  pdf:
    toc: true
    number-sections: true
    fig-align: center
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(ggplot2)
library(dplyr)
theme_set(theme_classic(base_size = 13))
```

## Setup and Motivation

Suppose we have an i.i.d. sample:
$$Y_1, Y_2, \ldots, Y_n \sim \mathcal{N}(\mu, \sigma^2).$$

We want to estimate (and test hypotheses about) $\mu$ and $\sigma^2$.

### Maximum Likelihood Estimates (MLE)

The MLEs are:
$$\hat{\mu} = \bar{y} = \frac{1}{n}\sum_{i=1}^{n} y_i, \qquad
\hat{\sigma}^2 = \frac{1}{n}\sum_{i=1}^{n}(y_i - \bar{y})^2 \quad \text{(biased)}.$$

In practice we prefer the **unbiased** sample variance (requires $n \geq 2$):
$$s^2 = \frac{1}{n-1}\sum_{i=1}^{n}(y_i - \bar{y})^2.$$

> **Note (n = 1 edge case):** If $n = 1$, then $\hat{\mu} = \bar{y} = y_1$ and
> $\hat{\sigma}^2 = (y_1 - \bar{y})^2 = 0$, which is degenerate. Always ensure $n \geq 2$.

---

## The z-Test (Known Variance)

**Hypotheses:**
$$H_0: \mu = 0 \quad \text{vs} \quad H_1: \mu \neq 0.$$

**Test statistic:** The sample mean satisfies
$$\mathbb{E}[\bar{Y}] = \mu, \qquad \operatorname{Var}(\bar{Y}) = \frac{\sigma^2}{n},$$
so after standardising:
$$Z = \frac{\bar{Y} - \mu_0}{\sigma / \sqrt{n}} \sim \mathcal{N}(0,1) \quad \text{when } \sigma = \sigma_0 \text{ is known}.$$

The **p-value** for the two-sided test is $P(|Z| \geq z_\text{obs})$.

```{r}
#| label: fig-z-dist
#| fig-cap: "Standard normal distribution with critical region (two-sided, α = 0.05) shaded. The observed test statistic z_obs = 2.5 is shown."
#| fig-alt: "A standard normal bell curve. The two tails beyond ±1.96 are shaded in red indicating the rejection region. A vertical dashed blue line marks z_obs = 2.5."

z_obs <- 2.5
z_crit <- qnorm(0.975)

x <- seq(-4, 4, length.out = 500)
df_z <- data.frame(x = x, y = dnorm(x))

ggplot(df_z, aes(x, y)) +
  geom_line(linewidth = 0.9) +
  # Shade left tail
  geom_ribbon(
    data = subset(df_z, x <= -z_crit),
    aes(ymin = 0, ymax = y), fill = "#d62728", alpha = 0.4
  ) +
  # Shade right tail
  geom_ribbon(
    data = subset(df_z, x >= z_crit),
    aes(ymin = 0, ymax = y), fill = "#d62728", alpha = 0.4
  ) +
  # Observed statistic
  geom_vline(xintercept = z_obs, linetype = "dashed", colour = "#1f77b4", linewidth = 0.9) +
  geom_vline(xintercept = -z_obs, linetype = "dashed", colour = "#1f77b4",
             linewidth = 0.9, alpha = 0.5) +
  annotate("text", x = z_obs + 0.25, y = 0.32, label = expression(z[obs]),
           hjust = 0, colour = "#1f77b4", size = 4) +
  annotate("text", x =  z_crit, y = 0.05,
           label = expression(-z[obs] ~ "" ~ z[obs]), hjust = 0, size = 3.5) +
  labs(x = "z", y = "Density",
       title = "Standard Normal: Critical Region for Two-Sided z-Test") +
  scale_x_continuous(
    breaks = c(-z_obs, -z_crit, 0, z_crit, z_obs),
    labels = c(expression(-z[obs]), expression(-z[crit]), "0",
               expression(z[crit]), expression(z[obs]))
  )
```

> **Practical limitation:** z-tests are rarely useful because $\sigma^2$ is almost never known in practice.

---

## The t-Test (Unknown Variance)

### Plug-in estimator

When $\sigma^2$ is unknown, we substitute $s^2$ for $\sigma^2$:
$$T = \frac{\bar{Y} - 0}{\sqrt{s^2/n}} = \frac{\bar{Y}}{s/\sqrt{n}}.$$

This statistic is **no longer** $\mathcal{N}(0,1)$-distributed.

### William Gosset ("Student") and the t-distribution

Gosset showed that if $\bar{Y} \sim \mathcal{N}(\mu, \sigma^2/n)$ and
$s^2 = \frac{1}{n-1}\sum_{i=1}^n (y_i - \bar{y})^2$, then:

$$\boxed{\frac{\bar{Y} - \mu}{s/\sqrt{n}} \sim t_{n-1}}$$

where $t_{n-1}$ denotes the **Student's t-distribution with $n-1$ degrees of freedom**.

As $n \to \infty$ (degrees of freedom $\to \infty$), $t_\nu \to \mathcal{N}(0,1)$.

```{r}
#| label: fig-t-dist
#| fig-cap: "t-distributions for increasing degrees of freedom compared to the standard normal. As df increases, the t-distribution converges to N(0,1), but has heavier tails for small df."
#| fig-alt: "Overlaid density curves for t distributions with 1, 5, 30 degrees of freedom and the standard normal. Heavier tails are visible for low degrees of freedom."

x <- seq(-5, 5, length.out = 500)

df_tdist <- bind_rows(
  data.frame(x = x, y = dt(x, df = 1),  label = "t, df = 1"),
  data.frame(x = x, y = dt(x, df = 5),  label = "t, df = 5"),
  data.frame(x = x, y = dt(x, df = 30), label = "t, df = 30"),
  data.frame(x = x, y = dnorm(x),       label = "N(0,1)")
)

ggplot(df_tdist, aes(x, y, colour = label, linetype = label)) +
  geom_line(linewidth = 0.9) +
  labs(x = "t", y = "Density",
       title = "t-Distributions vs Standard Normal",
       colour = "Distribution", linetype = "Distribution") +
  scale_colour_manual(values = c(
    "t, df = 1"  = "#d62728",
    "t, df = 5"  = "#ff7f0e",
    "t, df = 30" = "#2ca02c",
    "N(0,1)"     = "#1f77b4"
  )) +
  scale_linetype_manual(values = c(
    "t, df = 1"  = "dotted",
    "t, df = 5"  = "dashed",
    "t, df = 30" = "longdash",
    "N(0,1)"     = "solid"
  ))
```

### Computing p-values in R

In R, the t-distribution functions follow the standard `d/p/q/r` naming convention:

| Prefix | Function | Purpose |
|--------|----------|---------|
| `d` | `dt()` | Density |
| `p` | `pt()` | CDF / Probability |
| `q` | `qt()` | Quantile |
| `r` | `rt()` | Random sample |

For a two-sided test the p-value is:

$$p\text{-value} = P(|T| \geq t_\text{obs}) = 2 \times P(T \geq |t_\text{obs}|) = 2 \times (1 - \texttt{pt}(t_\text{obs},\, n-1))$$

### Example 

Read the `gpa_study_hours` data set from  [https://www.openintro.org/data/index.php?data=gpa_study_hours](https://www.openintro.org/data/index.php?data=gpa_study_hours). This dataset has 193 rows and 2 columns. The columns represent the variables `gpa` and `study_hours` for a sample of 193 undergraduate students who took an introductory statistics course in 2012 at a private US university.

We are interested in testing hypothesis about the mean `study_hours`, i.e. if students in this particular class studied more than a specified number of hours, e.g., we can ask if they have studied for more than 15 hours? 


```{r}
gpa_study_hours <- read.csv("https://www.openintro.org/data/csv/gpa_study_hours.csv")
str(gpa_study_hours)

mu0 = 15
y = gpa_study_hours$study_hours
n = length(y)
ybar = mean(y)
s = sd(y)

se = s/(n^(1/2))
t_stat = (ybar - mu0)/se

cat("sample size:", n, "sample mean:", ybar, "standard error:", se, "t:" , t_stat, "\n")

```


```{r}
#| label: pvalue-example
#| echo: true


p_upper    <- 1 - pt(t_stat, df = n - 1)   # one-sided (H1: mu > mu_0)
p_twosided <- 2 * p_upper                  # two-sided (H1: mu ≠ mu_0)

cat(sprintf("One-sided p-value  : %.4f\n", p_upper))
cat(sprintf("Two-sided p-value  : %.4f\n", p_twosided))
```

```{r}
#| label: fig-t-pvalue
#| fig-cap: "t-distribution with n-1s degrees of freedom. The shaded regions represent the two-sided p-value corresponding to t_stat = 3.015799."
#| fig-alt: "Bell-shaped t distribution curve with n-1 degrees of freedom. Both tails beyond ±t_stat are shaded in red, representing the p-value. A dashed blue line marks t_stat."

x <- seq(-4, 4, length.out = 500)
df_t80 <- data.frame(x = x, y = dt(x, df = n-1))

ggplot(df_t80, aes(x, y)) +
  geom_line(linewidth = 0.9) +
  geom_ribbon(data = subset(df_t80, x >= t_stat),
              aes(ymin = 0, ymax = y), fill = "#d62728", alpha = 0.4) +
  geom_ribbon(data = subset(df_t80, x <= -t_stat),
              aes(ymin = 0, ymax = y), fill = "#d62728", alpha = 0.4) +
  geom_vline(xintercept =  t_stat, linetype = "dashed", colour = "#1f77b4", linewidth = 0.9) +
  geom_vline(xintercept = -t_stat, linetype = "dashed", colour = "#1f77b4",
             linewidth = 0.9, alpha = 0.5) +
  annotate("text", x = t_stat + 0.1, y = 0.3,
           label = expression(t[obs]), hjust = 0, colour = "#1f77b4", size = 4) +
  annotate("text", x = -t_stat - 0.1, y = 0.3,
           label = expression(-t[obs]), hjust = 1, colour = "#1f77b4", size = 4) +
  labs(x = "t", y = "Density",
       title = expression("t"[80] ~ "distribution: two-sided p-value region"))
```

---

## Distribution of $s^2$: The Chi-Squared Connection

To derive the t-distribution formally, we need the sampling distribution of $s^2$.

### Chi-squared distribution

If $Y \sim \mathcal{N}(0,1)$, then $Y^2 \sim \chi^2_1$.

More generally, if $Y_1, \ldots, Y_n \overset{\text{iid}}{\sim} \mathcal{N}(0,1)$, then:
$$\sum_{i=1}^n Y_i^2 \sim \chi^2_n.$$

```{r}
#| label: fig-chisq
#| fig-cap: "Chi-squared densities for increasing degrees of freedom. The distribution is right-skewed for low df and becomes more symmetric as df grows."
#| fig-alt: "Overlaid chi-squared density curves for 1, 2, 5, and 10 degrees of freedom. The curve for df=1 has a sharp peak near zero; higher df curves shift right and become more bell-shaped."

x <- seq(0.01, 25, length.out = 500)

df_chi <- bind_rows(
  data.frame(x = x, y = dchisq(x, df = 1),  label = "df = 1"),
  data.frame(x = x, y = dchisq(x, df = 2),  label = "df = 2"),
  data.frame(x = x, y = dchisq(x, df = 5),  label = "df = 5"),
  data.frame(x = x, y = dchisq(x, df = 10), label = "df = 10")
)

ggplot(df_chi, aes(x, y, colour = label)) +
  geom_line(linewidth = 0.9) +
  coord_cartesian(ylim = c(0, 0.5)) +
  labs(x = expression(chi^2), y = "Density",
       title = expression("Chi-Squared Distributions" ~ chi[df]^2),
       colour = "Degrees of freedom") +
  scale_colour_brewer(palette = "Set1")
```

### Decomposition of the total sum of squares

For $Y_i \sim \mathcal{N}(\mu, \sigma^2)$, we know $\sum_{i=1}^n \left(\frac{Y_i - \mu}{\sigma}\right)^2 \sim \chi^2_n$.

By adding and subtracting $\bar{y}$:
$$\sum_{i=1}^n \left(\frac{Y_i - \mu}{\sigma}\right)^2
= \sum_{i=1}^n \frac{(Y_i - \bar{Y})^2}{\sigma^2} + \left(\frac{\bar{Y} - \mu}{\sigma/\sqrt{n}}\right)^2
= \frac{(n-1)s^2}{\sigma^2} + \underbrace{\left(\frac{\bar{Y} - \mu}{\sigma/\sqrt{n}}\right)^2}_{\sim\, \chi^2_1}.$$

The cross-term $\sum_i (y_i - \bar{y})(\bar{y} - \mu) = 0$ by construction, so by independence of the two components:

$$\boxed{\frac{(n-1)s^2}{\sigma^2} \sim \chi^2_{n-1}}$$

Equivalently:
$$s^2 \sim \frac{\sigma^2}{n-1} \cdot \chi^2_{n-1}, \qquad
\sigma^2 \sim \frac{(n-1)s^2}{\chi^2_{n-1}}.$$

---

## Summary: The One-Sample t-Test

**Setup:** $Y_1, \ldots, Y_n \overset{\text{iid}}{\sim} \mathcal{N}(\mu, \sigma^2)$, both $\mu$ and $\sigma^2$ unknown.

**Hypotheses:**
$$H_0: \mu = \mu_0 \quad \text{vs} \quad H_1: \mu \neq \mu_0 \quad (\text{two-sided}).$$

**Combining results:** Since $\bar{Y} \sim \mathcal{N}(\mu, \sigma^2/n)$ and $\frac{(n-1)s^2}{\sigma^2} \sim \chi^2_{n-1}$ are **independent**, their ratio yields:

$$\bar{Y} \sim t_{n-1}\!\left(\mu,\, \frac{s^2}{n}\right), \qquad
T = \frac{\bar{Y} - \mu_0}{s/\sqrt{n}} \sim t_{n-1} \quad \text{under } H_0.$$

**p-values:**

$$p\text{-value} = \begin{cases} P(T > t_\text{obs}) & \text{if } H_1: \mu > \mu_0, \\ P(|T| > t_\text{obs}) & \text{if } H_1: \mu \neq \mu_0. \end{cases}$$

**95% Confidence interval:**
$$\bar{Y} \pm t_{n-1,\, 0.975} \cdot \frac{s}{\sqrt{n}} \approx \bar{Y} \pm 1.96 \cdot \frac{s}{\sqrt{n}} \quad \text{(large } n).$$

**In R:**
```{r}
#| label: ttest-example
#| echo: true

# Simulate some data and run a one-sample t-test
set.seed(42)
y <- rnorm(30, mean = 0.3, sd = 1)

result <- t.test(y, mu = 0)
print(result)
```

### Caveat: Gaussianity assumption

The exact result $\frac{\bar{Y} - \mu}{s/\sqrt{n}} \sim t_{n-1}$ holds **only** when $Y_1, \ldots, Y_n$ are Gaussian. For non-Gaussian data, alternatives include:

- **Nonparametric tests** (e.g., Wilcoxon signed-rank test).
- **Monte Carlo / permutation approaches** to obtain the sampling distribution of the test statistic empirically.

---

## Problem: Gaussianity Assumption and Alternatives

```{r}
#| label: fig-sampling-dist
#| echo: true
#| fig-cap: "Empirical sampling distribution of the t-statistic under the null, obtained via Monte Carlo simulation (n = 30, B = 10,000 replicates), overlaid with the theoretical t_{29} density. The close agreement illustrates how the t-distribution arises in practice."
#| fig-alt: "A histogram of simulated t-statistics overlaid with a red t distribution curve. The empirical distribution closely follows the theoretical curve, validating the t-test under normality."


B <- 10000
n <- 30
t_stats <- replicate(B, {
  y_sim <- rnorm(n, mean = 0, sd = 1)
  t.test(y_sim, mu = 0)$statistic
})

x_seq <- seq(-4, 4, length.out = 300)
theory <- data.frame(x = x_seq, y = dt(x_seq, df = n - 1))

ggplot(data.frame(t = t_stats), aes(x = t)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 60, fill = "#aec7e8", colour = "white") +
  geom_line(data = theory, aes(x, y), colour = "#d62728", linewidth = 1) +
  labs(x = "t-statistic", y = "Density",
       title = expression("Sampling Distribution of T under H"[0] ~ "(Monte Carlo vs" ~ t[29] ~ ")"))
```
