---
title: "Reading QQ-Plots"
format:
  revealjs:
    theme: simple
    slide-number: true
    fig-align: center
    smaller: false
    self-contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
set.seed(42)
n <- 500

make_slide_plots <- function(x, y, caption) {
  probs <- seq(0.01, 0.99, length.out = 200)
  qx <- quantile(x, probs)
  qy <- quantile(y, probs)
  rng <- range(c(qx, qy))
  
  oldpar <- par(no.readonly = TRUE)
  on.exit(par(oldpar))
  
  # Device is 9 x 5.5 inches. We want:
  #   Left half  (x: 0.00-0.50): square QQ plot
  #   Top-right  (x: 0.50-1.00, y: 0.50-1.00): histogram of X
  #   Bot-right  (x: 0.50-1.00, y: 0.00-0.50): histogram of Y
  
  # QQ plot — square panel on the left
  par(fig = c(0.00, 0.50, 0.08, 0.97), mar = c(4, 4, 1, 1), new = FALSE)
  plot(qx, qy, type = "l", lwd = 1.5,
       xlab = "X", ylab = "Y",
       xlim = rng, ylim = rng, asp = 1)
  abline(0, 1, lty = 2)
  
  # Histogram of X — top right
  par(fig = c(0.50, 1.00, 0.50, 0.97), mar = c(4, 3, 1, 1), new = TRUE)
  hist(x, breaks = 20, col = "white", border = "black",
       main = "", xlab = "X", ylab = "")
  
  # Histogram of Y — bottom right
  par(fig = c(0.50, 1.00, 0.08, 0.55), mar = c(4, 3, 1, 1), new = TRUE)
  hist(y, breaks = 20, col = "white", border = "black",
       main = "", xlab = "Y", ylab = "")
  
  mtext(caption, side = 1, outer = TRUE, line = -1.5, cex = 1.5, font = 3)
}
```

## Why two-sample QQ plots? {.center}

Two-sample QQ plots let you compare distributions **without assuming any parametric form**.
By plotting the quantiles of $X$ against the quantiles of $Y$, deviations from the
45° reference line reveal *what kind* of difference exists between the groups, more than just
*whether* one exists.


---

## Where and how 


Unlike a hypothesis test, which returns a single $p$-value, the QQ plot tells you
**where** and **how** the distributions differ — upper tail only, throughout the
whole range, or just in location. 

This makes it an essential diagnostic before
choosing a two-sample test: a pure location shift supports the $t$-test or
Wilcoxon, but tail differences or unequal spread may call for something else.



---

## Y stochastically greater than X {.center}

```{r slide1, fig.width=9, fig.height=5.5}
x1 <- rnorm(n, mean = 0, sd = 1)
y1 <- rnorm(n, mean = 0.75, sd = 1)
make_slide_plots(x1, y1,
  "Y is stochastically greater than X.")
```

---

## X stochastically greater than Y {.center}

```{r slide2, fig.width=9, fig.height=5.5}
# X is Poisson(2), Y is Poisson(1) — right-skewed, lower quantiles close
x2 <- rpois(n, lambda = 2)
y2 <- rpois(n, lambda = 1)
make_slide_plots(x2, y2,
  "X is stochastically greater than Y, but their lower quantiles are very close.")
```

---

## X more variable than Y {.center}

```{r slide3, fig.width=9, fig.height=5.5}
x3 <- rnorm(n, mean = 0, sd = 2)
y3 <- rnorm(n, mean = 0, sd = 1)
make_slide_plots(x3, y3,
  "X is more variable than Y.")
```

---

## Y more variable than X {.center}

```{r slide4, fig.width=9, fig.height=5.5}
x4 <- rnorm(n, mean = 0, sd = 1)
y4 <- rnorm(n, mean = 0, sd = 2)
make_slide_plots(x4, y4,
  "Y is more variable than X.")
```

---

## Same distribution {.center}

```{r slide5, fig.width=9, fig.height=5.5}
x5 <- rnorm(n, mean = 0, sd = 1)
y5 <- rnorm(n, mean = 0, sd = 1)
make_slide_plots(x5, y5,
  "X and Y have the same distribution.")
```

---

## X has a longer right tail {.center}

```{r slide6, fig.width=9, fig.height=5.5}
x6 <- c(rnorm(round(n * 0.75), 0, 1), rexp(round(n * 0.25), 0.5))
y6 <- rnorm(n, mean = 0, sd = 1)
make_slide_plots(x6, y6,
  "X has a longer right tail.")
```

---

## X has longer right tail and shorter left tail {.center}

```{r slide7, fig.width=9, fig.height=5.5}
# X is right-skewed (log-normal shifted), Y is symmetric
x7 <- rlnorm(n, meanlog = 0, sdlog = 0.8) - 0.5
y7 <- rnorm(n, mean = 0.5, sd = 1)
make_slide_plots(x7, y7,
  "X has a longer right tail and a shorter left tail.")
```


---

## Summary of patterns


| Pattern | What the QQ plot shows |
|---|---|
| Curve above/below the diagonal | Stochastic ordering (one group tends larger) |
| Slope $\neq 1$ through the origin | Difference in spread/variability |
| S-shape | One distribution has heavier tails |
| Shifted S-shape | Skewness differs between groups |
| Hugs the diagonal | Distributions are similar |


