---
title: "Collapse of the Horseshoe Global Scale: A Small Simulation"
author: "Jyotishka Datta"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                       fig.align = "center")
```

## Purpose

This note reproduces, on a small and fast scale, the two qualitative patterns behind the companion theory note: as the proportion of nonzero coordinates grows (past 1/2) (i) the horseshoe global scale $\tau$ collapses toward zero, and (ii) the horseshoe misclassification probability rises sharply. 

```{r libs}
suppressMessages(library(horseshoe))
suppressMessages(library(ggplot2))
```

## Simulation setup

We use an $n\times p$ design with i.i.d. $N(0,2)$ entries, a signal of
size 5 on a proportion `prop` of the coordinates and zero elsewhere, and
$N(0,5)$ noise:

```{r helpers}
make_X <- function(n, p) matrix(rnorm(n * p, 0, sqrt(2)), nrow = n, ncol = p)

make_beta_y <- function(X, prop, signal = 5, noise_sd = sqrt(5)) {
  n <- nrow(X); p <- ncol(X)
  k <- round(prop * p)
  beta <- c(rep(signal, k), rep(0, p - k))
  y <- as.vector(X %*% beta + rnorm(n, 0, noise_sd))
  list(beta = beta, y = y)
}

n <- 100
p <- 100
props <- seq(0.1, 0.9, by = 0.05)
```

## Figure 1: collapse of $\tau$

For a single design $X$, we fit the horseshoe at each sparsity level and
keep the post-burn-in draws of $\tau$. This mirrors the `tau_hs`/`tau_box`
computation in `DL_Simulation.R`, with the square-root Dirichlet--Laplace
part removed.

```{r tau-collapse, cache=TRUE}
set.seed(2026)
X <- make_X(n, p)

tau_df <- do.call(rbind, lapply(props, function(prop) {
  dat <- make_beta_y(X, prop)
  invisible(capture.output(
    res <- horseshoe(dat$y, X, method.tau = "truncatedCauchy",
                      method.sigma = "Jeffreys", burn = 1000, nmc = 3000)
  ))
  data.frame(sparsity = prop, tau = res$TauSamples)
}))
tau_df$sparsity <- factor(tau_df$sparsity)
```

```{r tau-collapse-plot, fig.width=6, fig.height=4}
tau_plot <- ggplot(tau_df, aes(x = sparsity, y = tau)) +
  geom_boxplot(outlier.size = 0.5, fill = "grey85") +
  labs(x = "Proportion of nonzero coordinates",
       y = expression(tau),
       title = "Collapse of the horseshoe global scale") +
  theme_bw()
tau_plot
ggsave("Tau_hs.pdf", tau_plot, width = 6, height = 4)
```

## Figure 2: misclassification probability

For each sparsity level, we draw a fresh design $X$ per replicate , fit the horseshoe, classify coordinates with `HS.var.select(..., method = "intervals")`, and average the misclassification proportion over replicates.

```{r misclass, cache=TRUE}
set.seed(2026)
n_rep <- 20

misclass_mat <- matrix(NA_real_, nrow = n_rep, ncol = length(props))
mse_mat      <- matrix(NA_real_, nrow = n_rep, ncol = length(props))

for (r in 1:n_rep) {
  Xr <- make_X(n, p)
  for (j in seq_along(props)) {
    dat <- make_beta_y(Xr, props[j])
    invisible(capture.output(
      res <- horseshoe(dat$y, Xr, method.tau = "truncatedCauchy",
                        method.sigma = "Jeffreys", burn = 500, nmc = 1500)
    ))
    index <- as.numeric(dat$beta != 0)
    hs_ind <- HS.var.select(res, dat$y, method = "intervals")
    misclass_mat[r, j] <- mean(hs_ind != index)
    mse_mat[r, j]      <- mean((res$BetaHat - dat$beta)^2)
  }
}

summarize_reps <- function(mat, props, name) {
  m  <- colMeans(mat)
  se <- apply(mat, 2, sd) / sqrt(nrow(mat))
  data.frame(prop = props, mean = m, se = se,
             ymin = m - se, ymax = m + se, metric = name)
}

misclass_summary <- summarize_reps(misclass_mat, props, "Misclassification probability")
mse_summary      <- summarize_reps(mse_mat,      props, "Mean squared error")
```

```{r misclass-plot, fig.width=6, fig.height=4}
mc_plot <- ggplot(misclass_summary, aes(x = prop, y = mean)) +
  geom_ribbon(aes(ymin = pmax(ymin, 0), ymax = pmin(ymax, 1)),
              fill = "steelblue", alpha = 0.2) +
  geom_line(color = "steelblue") +
  geom_point(size = 3, color = "steelblue") +
  scale_x_continuous(breaks = props) +
  ylim(0, 1) +
  labs(x = "Proportion of nonzero coordinates",
       y = "Misclassification probability",
       title = "Horseshoe misclassification probability",
       subtitle = paste0("Mean \u00b1 1 SE over ", n_rep, " replicates")) +
  theme_bw()
mc_plot
ggsave("Sparsity_MSP_n=p.pdf", mc_plot, width = 6, height = 4)
```

```{r mse-plot, fig.width=6, fig.height=4}
mse_plot <- ggplot(mse_summary, aes(x = prop, y = mean)) +
  geom_ribbon(aes(ymin = pmax(ymin, 0), ymax = ymax),
              fill = "darkorange", alpha = 0.2) +
  geom_line(color = "darkorange") +
  geom_point(size = 3, color = "darkorange") +
  scale_x_continuous(breaks = props) +
  labs(x = "Proportion of nonzero coordinates",
       y = "Mean squared error",
       title = "Horseshoe mean squared error",
       subtitle = paste0("Mean \u00b1 1 SE over ", n_rep, " replicates")) +
  theme_bw()
mse_plot
ggsave("MSE_n=p.pdf", mse_plot, width = 6, height = 4)
```



## Notes

* `n_rep` and the MCMC lengths are kept small so the document knits in a few minutes; increase both for a smoother, lower-variance curve if you have the time before submission.
* Both code chunks are cached (`cache=TRUE`); re-knitting without changing the code will not re-run the MCMC.
