---
title: "The Central Limit Theorem"
author: "Jyotishka Datta"
institute: "Virginia Tech"
format:
  revealjs:
    theme: simple
    slide-number: true
    chalkboard: true
    preview-links: auto
    transition: fade
    incremental: false
    width: 1280
    height: 720
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
header-includes: |
  <style>
  :root { --vt-maroon: #861F41; --vt-orange: #E5751F; }
  .reveal h1, .reveal h2 { color: var(--vt-maroon); }
  .reveal strong { color: var(--vt-maroon); }
  .reveal .slide-number { color: var(--vt-maroon); }
  .callout { border-left: 0.35rem solid var(--vt-orange); padding-left: 1rem; }
  .small { font-size: 0.72em; }
  .center { text-align: center; }
  </style>
---

## Today's question

Suppose the population is skewed, discrete, or otherwise very non-Normal.

::: {.callout}
Why does the sample mean often have an approximately Normal distribution?
:::

By the end, you should be able to:

- state the Central Limit Theorem;
- distinguish a population distribution from a sampling distribution;
- compute probabilities for sample means, sums, and proportions;
- judge when a Normal approximation is reasonable.

---

## Three distributions—not one

1. **Population distribution:** the distribution of one observation, $X$.
2. **Sample distribution:** the values $X_1,\ldots,X_n$ in one observed sample.
3. **Sampling distribution:** the distribution of a statistic, such as $\bar X_n$, over repeated samples.

::: {.fragment}
The CLT is a statement about the **sampling distribution** of a standardized sum or mean.
:::

---

## A thought experiment

Start with a strongly right-skewed population:

```{r}
#| label: population-shape
#| fig-width: 7
#| fig-height: 4
#| echo: false
x <- seq(0, 10, length.out = 500)
plot(x, dexp(x), type = "l", lwd = 4, col = "#861F41",
     xlab = "x", ylab = "Density", main = "Population: Exponential(1)")
polygon(c(x, rev(x)), c(dexp(x), rep(0, length(x))),
        col = adjustcolor("#E5751F", 0.25), border = NA)
lines(x, dexp(x), lwd = 4, col = "#861F41")
```

If we repeatedly average $n$ independent observations, what shape will those averages have?

---

## Repeated sampling

```{r}
#| label: sampling-distributions
#| fig-width: 10
#| fig-height: 4.7
#| fig-align: center
#| echo: false
set.seed(3504)
B <- 30000
ns <- c(1, 5, 30)
oldpar <- par(mfrow = c(1, 3), mar = c(4, 4, 3, 1))
for (n in ns) {
  means <- replicate(B, mean(rexp(n)))
  hist(means, probability = TRUE, breaks = 45,
       col = adjustcolor("#861F41", 0.68), border = "white",
       main = paste("n =", n), xlab = expression(bar(X)))
  curve(dnorm(x, mean = 1, sd = 1/sqrt(n)), add = TRUE,
        col = "#E5751F", lwd = 3)
}
par(oldpar)
```

The center stays at $\mu=1$, the spread shrinks, and the shape becomes more Normal.

---

## The Central Limit Theorem

Let $X_1,X_2,\ldots$ be IID with

$$
E(X_i)=\mu, \qquad \operatorname{Var}(X_i)=\sigma^2<\infty.
$$

Then

$$
Z_n=\frac{\bar X_n-\mu}{\sigma/\sqrt n}
\xrightarrow{d} N(0,1)
\qquad \text{as }n\to\infty.
$$

Equivalently, for sufficiently large $n$,

$$
\bar X_n \mathrel{\dot\sim} N\!\left(\mu,\frac{\sigma^2}{n}\right).
$$

::: {.fragment}
The population itself does **not** become Normal. The distribution of the standardized mean does.
:::

---

## What changes as $n$ grows?

For $\bar X_n$:

$$
E(\bar X_n)=\mu,
\qquad
\operatorname{Var}(\bar X_n)=\frac{\sigma^2}{n},
\qquad
\operatorname{SE}(\bar X_n)=\frac{\sigma}{\sqrt n}.
$$

- **Center:** remains at $\mu$.
- **Spread:** decreases at rate $1/\sqrt n$.
- **Shape:** becomes approximately Normal under the CLT conditions.

Doubling $n$ does not halve the standard error. To halve it, multiply $n$ by $4$.

---

## The standardized picture

```{r}
#| label: standardized-means
#| fig-width: 8
#| fig-height: 4.6
#| echo: false
set.seed(17)
B <- 40000
n <- 30
z <- replicate(B, {
  xb <- mean(rexp(n))
  (xb - 1)/(1/sqrt(n))
})
hist(z, probability = TRUE, breaks = 50,
     col = adjustcolor("#861F41", 0.68), border = "white",
     xlab = expression(Z[n]), main = "Standardized means from an Exponential population")
curve(dnorm(x), add = TRUE, col = "#E5751F", lwd = 4)
legend("topright", c("Simulation", "N(0,1)"),
       fill = c("#861F41", NA), border = c("#861F41", NA),
       lty = c(NA, 1), lwd = c(NA, 4), col = c(NA, "#E5751F"), bty = "n")
```

---

## A practical CLT workflow

For a probability involving a sample mean:

1. Identify $\mu$, $\sigma$, and $n$.
2. Check independence and whether the sample size is plausible.
3. Use
   $$\bar X\mathrel{\dot\sim}N\!\left(\mu,\frac{\sigma^2}{n}\right).$$
4. Standardize the boundary:
   $$z=\frac{\text{boundary}-\mu}{\sigma/\sqrt n}.$$
5. Use $\Phi$, a Normal table, or `pnorm()`.
6. Interpret the probability in context.

---

## Worked example: coffee fill volume

A machine fills cups with mean $\mu=12.0$ oz and standard deviation $\sigma=0.6$ oz. Individual fills are somewhat skewed. For a random sample of $n=36$ cups, approximate

$$P(11.9<\bar X<12.2).$$

The standard error is

$$
\frac{0.6}{\sqrt{36}}=0.1.
$$

Therefore,

$$
P(11.9<\bar X<12.2)
\approx P\left(-1<Z<2\right)
=\Phi(2)-\Phi(-1).
$$

```{r}
pnorm(2) - pnorm(-1)
```

---

## Sums are covered too

Let $S_n=X_1+\cdots+X_n$. Since $S_n=n\bar X_n$,

$$
\frac{S_n-n\mu}{\sigma\sqrt n}
\xrightarrow{d}N(0,1),
$$

so

$$
S_n\mathrel{\dot\sim}N(n\mu,n\sigma^2).
$$

::: {.callout}
For a **mean**, the standard deviation is $\sigma/\sqrt n$.  
For a **sum**, the standard deviation is $\sigma\sqrt n$.
:::

---

## Worked example: total demand

Daily demand has mean $42$ units and standard deviation $9$ units. Assume demands on different days are IID. Over $50$ days, approximate

$$P(S_{50}>2200).
$$

Under the CLT,

$$
S_{50}\mathrel{\dot\sim}N(50\cdot42,50\cdot9^2)
=N(2100,4050).
$$

Thus

$$
z=\frac{2200-2100}{9\sqrt{50}}\approx 1.571.
$$

```{r}
1 - pnorm((2200 - 50*42)/(9*sqrt(50)))
```

---

## Bernoulli variables and sample proportions

If $X_i\sim\operatorname{Bernoulli}(p)$, then

$$
\bar X=\hat p, \qquad E(\hat p)=p,
\qquad \operatorname{SE}(\hat p)=\sqrt{\frac{p(1-p)}{n}}.
$$

The CLT gives

$$
\frac{\hat p-p}{\sqrt{p(1-p)/n}}
\mathrel{\dot\sim}N(0,1).
$$

A common rule of thumb is that both $np$ and $n(1-p)$ should be at least about $10$ for a comfortable Normal approximation.

---

## De Moivre–Laplace: Binomial CLT

If $X\sim\operatorname{Bin}(n,p)$, then

$$
\frac{X-np}{\sqrt{np(1-p)}}
\xrightarrow{d}N(0,1).
$$

Therefore,

$$
X\mathrel{\dot\sim}N\bigl(np,np(1-p)\bigr).
$$

Because $X$ is discrete and the Normal distribution is continuous, a **continuity correction** usually improves the approximation.

---

## Continuity correction

Let $Y\sim N(np,np(1-p))$. Translate integer outcomes into intervals of width $1$:

| Binomial probability | Normal approximation |
|---|---|
| $P(X\le k)$ | $P(Y\le k+0.5)$ |
| $P(X<k)$ | $P(Y<k-0.5)$ |
| $P(X\ge k)$ | $P(Y\ge k-0.5)$ |
| $P(X>k)$ | $P(Y>k+0.5)$ |
| $P(X=k)$ | $P(k-0.5<Y<k+0.5)$ |

**Memory aid:** move the Normal boundary outward so that it covers exactly the integer outcomes you want.

---

## Binomial example

Let $X\sim\operatorname{Bin}(25,0.4)$. Approximate $P(X\le8)$.

$$
\mu=np=10,
\qquad
\sigma=\sqrt{np(1-p)}=\sqrt6.
$$

With continuity correction,

$$
P(X\le8)
\approx
\Phi\!\left(\frac{8.5-10}{\sqrt6}\right).
$$

```{r}
c(
  exact = pbinom(8, 25, 0.4),
  normal_no_correction = pnorm(8, 10, sqrt(6)),
  normal_with_correction = pnorm(8.5, 10, sqrt(6))
)
```

---

## How large must $n$ be?

There is no universal cutoff.

- Nearly symmetric, light-tailed populations: modest $n$ may work well.
- Strong skewness: larger $n$ is needed.
- Rare-event Bernoulli data: check both $np$ and $n(1-p)$.
- Heavy tails: convergence may be very slow.
- Infinite variance: the classical CLT does not apply.

::: {.fragment}
“$n\ge30$” is a rough classroom mnemonic—not a theorem.
:::

---

## Conditions and warning signs

The classical version assumes:

- observations are independent;
- observations are identically distributed;
- the common mean exists;
- the common variance is finite.

Be cautious with:

- time series or spatial dependence;
- a few extreme observations;
- samples dominated by one cluster;
- very skewed populations with small $n$;
- Cauchy-like heavy tails.

---

## What the CLT does **not** say

- It does not say the raw data are Normal.
- It does not guarantee a good approximation for every $n$.
- It does not make dependence disappear.
- It does not justify ignoring infinite or unstable variance.
- It does not say $\bar X$ is exactly Normal—unless the population is Normal.

---

## Quick check

A population has mean $80$ and standard deviation $24$. For samples of size $64$:

1. What is the approximate distribution of $\bar X$?
2. What is $P(77<\bar X<85)$?
3. If the sample size becomes $256$, what happens to the standard error?

. . .

**Answers:**

$$
\bar X\mathrel{\dot\sim}N(80,3^2),
$$

$$
P(77<\bar X<85)\approx\Phi(5/3)-\Phi(-1),
$$

and the standard error falls from $3$ to $1.5$.

---

## Takeaways

1. The CLT concerns the distribution of a **standardized sum or mean**.
2. For large $n$,
   $$\bar X\mathrel{\dot\sim}N\!\left(\mu,\frac{\sigma^2}{n}\right).$$
3. The mean stays centered at $\mu$ while its standard error shrinks as $1/\sqrt n$.
4. Sums, proportions, and Binomial counts are direct applications.
5. Independence, finite variance, skewness, and tail behavior determine whether the approximation is credible.

## Questions?
