---
title: "Order Statistics, Ranks, and the KS Test"
subtitle: "Nonparametric Statistics"
author: "Jyotishka Datta"
institute: "Virginia Tech"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [default, default-fonts, custom.css]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.retina = 3, fig.align = "center",
  fig.width = 8, fig.height = 4
)
library(nortest)
library(goftest)
```

class: inverse, center, middle

# Part 1: Order Statistics

---

## What Are Order Statistics?

Given a random sample $X_1, X_2, \ldots, X_n$, the **order statistics** are
the sample values arranged in non-decreasing order:

$$X_{(1)} \leq X_{(2)} \leq \cdots \leq X_{(n)}$$

- $X_{(1)}$ = **sample minimum**
- $X_{(n)}$ = **sample maximum**
- $X_{(i)}$ = $i^{\text{th}}$ smallest value

--

**Derived summaries:**

- **Median:** $M = X_{((n+1)/2)}$ for $n$ odd; average of two middle values for $n$ even
- **Sample range:** $X_{(n)} - X_{(1)}$
- **Midrange:** $(X_{(n)} + X_{(1)})/2$

> Order statistics are **random quantities**, i.e., their distribution depends on the parent $F$.

---

## Distribution of the Maximum

Let $X_1, \ldots, X_n \overset{\text{iid}}{\sim} F$ (continuous). Let $G$ be the CDF of $X_{(n)}$.

The maximum is $\leq t$ iff **every** observation is $\leq t$:

$$G(t) = P(X_{(n)} \leq t) = F^n(t) \qquad \Rightarrow \qquad \boxed{g(t) = n\,F^{n-1}(t)\,f(t)}$$

--

For the **minimum** $X_{(1)}$:

$$H(t) = 1 - (1 - F(t))^n \qquad \Rightarrow \qquad \boxed{h(t) = n\,(1-F(t))^{n-1}\,f(t)}$$

---

## General $r^{\text{th}}$ Order Statistic

$$\boxed{f_{X_{(r)}}(x) = \frac{n!}{(r-1)!\,(n-r)!}\,F^{r-1}(x)\,(1-F(x))^{n-r}\,f(x)}$$

**Memory aid:** Think multinomial: 

- $(r-1)$ observations **below** $x$: probability $F^{r-1}(x)$
- **One** observation **at** $x$: probability $f(x)dx$
- $(n-r)$ observations **above** $x$: probability $(1-F(x))^{n-r}$

--

.callout[
**Key fact:** When $F = \text{Uniform}(0,1)$, this is **Beta$(r,\; n-r+1)$**.  
This is central to the distribution-free property of the KS test.
]

---

## Simulating Order Statistic Densities

```{r order-stat-demo, fig.height=3.5}
set.seed(42); n <- 10; B <- 50000
sims <- replicate(B, sort(runif(n)))
par(mfrow = c(1, 3), mar = c(4, 3, 2.5, 1))
hist(sims[1,],  breaks=40, freq=FALSE, col="#22222255", border="white",
     main=expression(X[(1)]~"~ Beta(1,10)"), xlab="x")
curve(dbeta(x,1,n), add=TRUE, col="#1a1a1a", lwd=2)
hist(sims[5,],  breaks=40, freq=FALSE, col="#22222255", border="white",
     main=expression(X[(5)]~"~ Beta(5,6)"), xlab="x")
curve(dbeta(x,5,6), add=TRUE, col="#1a1a1a", lwd=2)
hist(sims[10,], breaks=40, freq=FALSE, col="#22222255", border="white",
     main=expression(X[(10)]~"~ Beta(10,1)"), xlab="x")
curve(dbeta(x,n,1), add=TRUE, col="#1a1a1a", lwd=2)
par(mfrow=c(1,1))
```

---

class: inverse, center, middle

# Part 2: Ranks

---

## Ranks 

Definition:

The **rank** of $X_i$ in a sample $X_1, \ldots, X_n$ is the number of
observations $\leq X_i$:

$$\text{rank}(X_i) = n \times S_n(X_i) = \sum_{j=1}^{n} \mathbf{1}(X_j \leq X_i)$$

**Key facts:**
- Ranks are integers in $\{1, 2, \ldots, n\}$
- The $r^{\text{th}}$ order statistic has rank $r$: $\text{rank}(X_{(r)}) = r$
- Ties $\to$ use **mid-ranks** (average of tied positions)

--

> Ranks discard the actual values and retain only relative ordering.  
> This is the foundation of **nonparametric statistics**.

---

## Two-Sample Ranks

With $m$ observations $X_1,\ldots,X_m$ (group 1) and $n$ observations $Y_1,\ldots,Y_n$ (group 2),
rank in the **combined sample** of $m+n$:

$$\text{rank}(Y_{(j)}) = \underbrace{\sum_{i=1}^{m} \mathbf{1}(X_i \leq Y_{(j)})}_{\text{how many } X\text{'s are} \leq Y_{(j)}} + \underbrace{j}_{\text{how many } Y\text{'s are} \leq Y_{(j)}}$$

This is the structure behind the **Wilcoxon rank-sum test** (coming up).

---

class: inverse, center, middle

# Part 3: The Empirical CDF

---

## Empirical Distribution Function (EDF)

Given i.i.d. $X_1, \ldots, X_n \sim F_X$, the **EDF** is:

$$S_n(x) = \frac{1}{n}\sum_{i=1}^{n} \mathbf{1}(X_i \leq x)$$

In terms of order statistics:

$$S_n(x) = \begin{cases} 0 & x < X_{(1)} \\ i/n & X_{(i)} \leq x < X_{(i+1)} \\ 1 & x \geq X_{(n)} \end{cases}$$

A **right-continuous step function** with jumps of $1/n$ at each observation.

---

## Properties of the EDF

For each fixed $x$, $T_n(x) = nS_n(x) \sim \text{Binomial}(n, F_X(x))$. Therefore:

$$\mathbb{E}[S_n(x)] = F_X(x) \quad \text{(unbiased)}$$

$$\text{Var}[S_n(x)] = \frac{F_X(x)(1-F_X(x))}{n} \to 0 \quad \text{(consistent)}$$

--

**Glivenko–Cantelli Theorem (Fundamental Theorem of Statistics):**

$$P\!\left[\lim_{n\to\infty} \sup_{x\in\mathbb{R}} |S_n(x) - F_X(x)| = 0\right] = 1$$

$S_n$ converges to $F_X$ **uniformly** over all $x$, **almost surely**.

> **Under $H_0: F_X = F_0$**, the supremum $\sup_x |S_n(x) - F_0(x)|$ should be **small** for large $n$.  
> This motivates the KS test. 

---

## EDF in Action

```{r edf-demo, fig.height=3.5}
set.seed(7)
par(mfrow=c(1,2))
for(n in c(20, 50)){
  x <- rnorm(n)
  plot(ecdf(x), main=paste0("n = ", n), xlab="x",
       ylab=expression(S[n](x)), col="#1a1a1a", lwd=2, xlim=c(-3,3))
  curve(pnorm(x), add=TRUE, col="#888888", lwd=2, lty=2)
  legend("topleft", c("EDF", "True CDF N(0,1)"),
         col=c("#1a1a1a","#888888"), lwd=2, lty=c(1,2), bty="n", cex=0.85)
}
par(mfrow=c(1,1))
```

---

class: inverse, center, middle

# Part 4: Probability Integral Transform

---

## The PIT 

We need one key result before proving the KS distribution-free property.

**Theorem (Probability Integral Transform).**
Let $X$ have a **continuous** CDF $F$. Then:
$$U = F(X) \sim \text{Uniform}(0,1)$$

**Proof.** For $u \in (0,1)$:
$$P(F(X) \leq u) = P(X \leq F^{-1}(u)) = F(F^{-1}(u)) = u \quad \checkmark$$

--

**Shorthand to remember:**
$$X \sim F \;\Rightarrow\; F(X) \sim U[0,1] \;\Rightarrow\; F^{-1}(U) \sim F$$

> **Continuity is essential.** For discrete $F$, $F(X)$ is *not* uniform. This has direct consequences for the KS test.

---

## The Inverse CDF Method (Recap)

We have studied this in detail previously. The key result:

**Algorithm (Inverse CDF Method)**
1. Generate $U_1, \ldots, U_n \overset{\text{iid}}{\sim} \text{Uniform}(0,1)$
2. Set $X_i = F^{-1}(U_i)$

Then $X_1, \ldots, X_n \overset{\text{iid}}{\sim} F$.

**Interactive demo:** https://dattahub.github.io/iqs/rvgen.html


> We will now use the PIT in a completely different way: to prove that the KS statistic is **distribution-free**.

---

class: inverse, center, middle

# Part 5: The Kolmogorov–Smirnov Test

---

## Setup

We observe $X_1, \ldots, X_n$ and want to test:

$$H_0: F_X(x) = F_0(x) \quad \text{for all } x \in \mathbb{R}$$

where $F_0$ is a **completely specified** continuous CDF.

**Idea:** By Glivenko–Cantelli, $S_n \approx F_X$.  
Under $H_0$, $S_n \approx F_0$, so $\sup_x |S_n(x) - F_0(x)|$ should be **small**.

**Large values $\to$ evidence against $H_0$.**

---

## The KS Statistics

$$D_n^+ = \sup_{x}\, [S_n(x) - F_0(x)] \qquad \text{(upper one-sided)}$$

$$D_n^- = \sup_{x}\, [F_0(x) - S_n(x)] \qquad \text{(lower one-sided)}$$

$$\boxed{D_n = \sup_{x}\, |S_n(x) - F_0(x)| = \max(D_n^+,\, D_n^-)}$$

$D_n$ is the **two-sided** statistic — the one we almost always use.

Reject $H_0$ when $D_n$ is **large** (exceeds the critical value).

---

## The Distribution-Free Property

**Theorem.** Under $H_0$, the distributions of $D_n$, $D_n^+$, $D_n^-$ are **independent of $F_0$**, provided $F_0$ is continuous.

**Proof sketch for $D_n^+$.** Using the step-function form of $S_n$:

$$D_n^+ = \max_{0\leq i\leq n}\left[\frac{i}{n} - F_0(X_{(i)})\right]$$

--

Since $X_{(i)}$ are order statistics from $F_0$, by the **PIT**:

$$F_0(X_{(i)}) \sim \text{order statistics from } U[0,1]$$

**regardless of what $F_0$ is.**

So $D_n^+$ depends on $F_0$ only through uniform order statistics $\Rightarrow$ distribution-free. $\square$

---

## Critical Consequences

**1. Universal critical values.**  
The same table works for *any* continuous null distribution.

**2. Requirements that must hold:**

| Requirement | What breaks if violated |
|-------------|------------------------|
| $F_0$ must be **continuous** | PIT fails; KS table invalid for discrete data |
| $F_0$ must be **completely specified** | If parameters estimated from data, critical values too large $\to$ test is conservative |

---

## Hand Calculation Formula

$$D_n = \max_{i=1,\ldots,n}\left\{
  \left|\frac{i}{n} - F_0(X_{(i)})\right|,\;
  \left|\frac{i-1}{n} - F_0(X_{(i)})\right|
\right\}$$

The two terms capture the jump from both sides of the step function.

**In R:**

```r
ks.test(x, "pnorm", mean = 0, sd = 1)   # test against N(0,1)
ks.test(x, "pexp",  rate = 2)            # test against Exp(2)
```

---

## Worked Example

```{r ks-example}
x_obs <- c(0.0123,0.1039,0.1954,0.2621,0.2802,
           0.3217,0.3645,0.3919,0.4240,0.4814,
           0.5139,0.5846,0.6275,0.6541,0.6889,
           0.7621,0.8320,0.8871,0.9249,0.9634)
y <- sqrt(x_obs)   # test if sqrt(x) ~ Uniform(0,1)
ks.test(y, "punif")
```

$D_{20} = 0.36 > 0.352$ (critical value at $\alpha = 0.01$) $\Rightarrow$ **reject $H_0$**

---

## Visualising the KS Statistic

```{r ks-plot, fig.height=3.8}
plot(ecdf(y), col="#1a1a1a", lwd=2,
     main=expression("EDF of"~sqrt(x)~"vs Uniform(0,1)"),
     xlab="x", ylab="CDF")
abline(0, 1, col="#888888", lwd=2, lty=2)
# Mark the maximum gap
i_max <- which.max(pmax(abs((1:20)/20 - y), abs((0:19)/20 - y)))
segments(y[i_max], (i_max-1)/20, y[i_max], y[i_max],
         col="#1a1a1a", lwd=3)
text(y[i_max]+0.05, ((i_max-1)/20+y[i_max])/2,
     expression(D[n]), col="#1a1a1a", cex=1.1)
legend("topleft", c("EDF of sqrt(x)","Uniform CDF"),
       col=c("#1a1a1a","#888888"), lwd=2, lty=c(1,2), bty="n")
```

---

## Asymptotic Distribution

**Theorem (Kolmogorov 1933, Smirnov 1939).** For any continuous $F_X$ and $d > 0$:
$$\lim_{n\to\infty} P\!\left(D_n \leq \frac{d}{\sqrt{n}}\right) = L(d) = 1 - \sum_{i=1}^{\infty}(-1)^{i-1} e^{-2i^2 d^2}$$

Good approximation for $n \geq 35$.

**For $n > 40$, critical values:**

| $\alpha$ | 0.20 | 0.10 | 0.05 | 0.02 | 0.01 |
|:---:|:---:|:---:|:---:|:---:|:---:|
| Critical value | $1.07/\sqrt{n}$ | $1.22/\sqrt{n}$ | $1.36/\sqrt{n}$ | $1.52/\sqrt{n}$ | $1.63/\sqrt{n}$ |

---

class: inverse, center, middle

# Part 6: Simulation-Based p-values for KS

---

## The Problem with the Asymptotic Table

The Kolmogorov–Smirnov table assumes $n$ is large and $F_0$ is fully specified.

But what if:
- $n$ is small (say $n = 15$)?
- We are unsure whether the approximation is adequate?
- We want to understand where the p-value actually comes from?

--

**Solution:** Generate the null distribution of $D_n$ by **simulation**.

> This is the philosophy of permutation tests: if we know the null hypothesis, we can simulate from it directly.

---

## Simulation-Based KS p-value: Algorithm

**Setup:** Observed data $x_1, \ldots, x_n$. Null hypothesis $H_0: F_X = F_0$.

**Step 1.** Compute the observed KS statistic:
$$D_n^{\text{obs}} = \sup_x |S_n(x) - F_0(x)|$$

**Step 2.** Repeat $B$ times (e.g. $B = 10{,}000$):
- Generate $x_1^*, \ldots, x_n^* \overset{\text{iid}}{\sim} F_0$
- Compute $D_n^{(b)} = \sup_x |S_n^*(x) - F_0(x)|$

**Step 3.** The simulation-based p-value is:
$$\hat{p} = \frac{1}{B}\sum_{b=1}^{B} \mathbf{1}\!\left(D_n^{(b)} \geq D_n^{\text{obs}}\right)$$

---

## Implementation in R

```{r ks-sim-pvalue}
x_data <- rexp(20, rate = 1.5)   # 20 observations
D_obs  <- ks.test(x_data, "pexp", rate = 1.5)$statistic
cat("Observed D_n:", round(D_obs, 4), "\n")

# Simulate null distribution
B <- 10000; n <- length(x_data)
D_null <- replicate(B, {
  x_star <- rexp(n, rate = 1.5)    # draw from F_0
  ks.test(x_star, "pexp", rate = 1.5)$statistic
})
# Simulation-based p-value
(p_sim <- mean(D_null >= D_obs))
# Compare with asymptotic
(p_asy <- ks.test(x_data, "pexp", rate = 1.5)$p.value)
```

---

## Visualising the Null Distribution

```{r ks-null-dist, fig.height=3.8}
hist(D_null, breaks=60, freq=FALSE,
     col="#88888855", border="white",
     main=expression("Null distribution of D"[n]~"(n=20, B=10000)"),
     xlab=expression(D[n]))
abline(v=D_obs, col="#1a1a1a", lwd=2.5, lty=2)
text(D_obs+0.01, par("usr")[4]*0.85,paste0("D[obs] = ", round(D_obs,3),"\np = ", round(p_sim,3)),
     adj=0, cex=0.9)
# shade the p-value region
x_shade <- seq(D_obs, max(D_null), length.out=200)
d_shade <- density(D_null, from=D_obs)
polygon(c(D_obs, d_shade$x[d_shade$x >= D_obs], max(D_null)),c(0, d_shade$y[d_shade$x >= D_obs], 0),col="#1a1a1a44", border=NA)
```

---

## When Is Simulation-Based Better?

.pull-left[
**Use simulation when:**
- $n$ is small (the asymptotic approximation needs $n \geq 35$)
- You want an exact finite-sample p-value
- Also makes the p-value conceptually transparent


]

.pull-right[
**Use the asymptotic table / `ks.test` when:**
- $n \geq 35$ (approximation is good)
- $F_0$ is a standard distribution
- Speed matters (simulation takes $B \times n$ draws)

```{r ks-compare-table, echo=FALSE}
df <- data.frame(
  Method = c("Asymptotic", "Simulation (B=1000)",
             "Simulation (B=10000)"),
  p_value = c(
    round(ks.test(x_data,"pexp",rate=1.5)$p.value,4),
    round(mean(replicate(1000,{
      ks.test(rexp(n,1.5),"pexp",rate=1.5)$statistic})>=D_obs),4),
    round(p_sim,4))
)
knitr::kable(df, col.names=c("Method","p-value"), align="lr")
```
]

---

## The Big Picture: Two Ways to Get a p-value

```{r two-ways, echo=FALSE, fig.height=6}
par(mfrow=c(1,2))

# Left: asymptotic
d_seq <- seq(0, 0.5, length.out=300)
# Kolmogorov distribution approximation
kol_dens <- function(d, n) {
  x <- d * sqrt(n)
  2 * sum(sapply(1:50, function(k)
    (-1)^(k-1) * k^2 * exp(-2*k^2*x^2))) * sqrt(n)
}
approx_d <- sapply(d_seq, function(d) kol_dens(d, n))
approx_d <- pmax(approx_d, 0)

plot(d_seq, approx_d/max(approx_d)*8, type="l",
     col="#1a1a1a", lwd=2,
     main="Asymptotic: Kolmogorov distribution",
     xlab=expression(D[n]), ylab="Density (scaled)")
abline(v=D_obs, lty=2, col="#888888", lwd=1.5)
text(D_obs+0.01, 6, "p-value\n(right tail)", cex=0.8, adj=0)

# Right: simulation
hist(D_null, breaks=50, freq=FALSE,
     col="#88888855", border="white",
     main="Simulation: empirical null distribution",
     xlab=expression(D[n]))
abline(v=D_obs, lty=2, col="#1a1a1a", lwd=1.5)
text(D_obs+0.01, par("usr")[4]*0.7, "p-value\n(fraction\nto the right)",
     cex=0.8, adj=0)
par(mfrow=c(1,1))
```

---

class: inverse, center, middle

# Part 7: The Lilliefors Test

---

## The Problem with Estimated Parameters

Suppose data come from $N(\mu, \sigma^2)$ with **unknown** $\mu, \sigma$.

**Tempting:** plug in $\hat\mu = \bar X$, $\hat\sigma = S$, run KS against $N(\hat\mu, \hat\sigma^2)$.

**Why this fails:**
- The fitted distribution is forced closer to the data
- KS statistic is systematically **smaller** than it would be under a truly specified $F_0$
- Standard KS critical values are **too large** $\to$ test is **conservative**

--

The simulation approach gives us a way to understand this directly:

```r
# WRONG: uses standard KS table with estimated parameters
ks.test(x, "pnorm", mean = mean(x), sd = sd(x))

# RIGHT: Lilliefors test with corrected critical values
library(nortest)
lillie.test(x)
```

---

## Lilliefors' Fix

**Lilliefors (1967)** derived corrected critical values by simulation. So, essentially the same idea as what we just did, but for the composite null $H_0$: $F$ is Normal (parameters unknown).

**Procedure:**
1. Compute $\hat\mu = \bar X$, $\hat\sigma = S$
2. Form the KS statistic $D_n$ against $N(\hat\mu, \hat\sigma^2)$
3. Compare to **Lilliefors critical values** (smaller than standard KS)

---

## Lilliefors' Fix


```{r lilliefors-demo}
set.seed(2024)
x_norm <- rnorm(50, mean=100, sd=10)
x_exp  <- rexp(50, rate=0.1)
lillie.test(x_norm)   # should NOT reject
lillie.test(x_exp)    # should reject
```

---

## KS vs Lilliefors: Decision Table

| Scenario | Test |
|----------|------|
| $H_0$: $F = F_0$ fully specified | **Kolmogorov–Smirnov** |
| $H_0$: $F$ is Normal, $\mu,\sigma$ unknown | **Lilliefors** |
| $H_0$: $F$ is Exponential, $\lambda$ unknown | **Lilliefors (exp variant)** |
| Discrete data | **Neither** — use $\chi^2$ GoF |


---

## KS vs Lilliefors: IQ Example

**IQ Example:** Scores of 18 girls: 114, 81, 87, 114, 113, 87, 111, 89, 93, 108, 99, 93, 100, 95, 93, 95, 106, 108

```{r iq-example}
iq <- c(114,81,87,114,113,87,111,89,93,108,99,93,100,95,93,95,106,108)
lillie.test(iq)                           # (a) unknown mu, sigma
ks.test(iq, "pnorm", mean=100, sd=10)    # (b) mu=100, sigma=10 specified
```

---

class: inverse, center, middle

# Part 8: Other GoF Tests

---

## Anderson–Darling: Tail-Sensitive KS

$$A^2 = -n - \sum_{i=1}^{n} \frac{2i-1}{n}\left[\log F_0(X_{(i)}) + \log(1-F_0(X_{(n+1-i)}))\right]$$

- Gives **more weight to the tails** of the distribution
- Uniformly more powerful than KS for many alternatives
- Available with parameter estimation corrections

```{r ad-test}
library(nortest)
set.seed(42); x <- rnorm(100)
ad.test(x)
```

---

## Cramér–von Mises and Shapiro–Wilk

**Cramér–von Mises**; integrated squared gap:
$$W^2 = \frac{1}{12n} + \sum_{i=1}^{n}\!\left(F_0(X_{(i)}) - \frac{2i-1}{2n}\right)^2$$

```{r cvm-shapiro, results='hold'}
library(goftest)
set.seed(42); x <- rnorm(100)
cvm.test(x, "pnorm")$p.value      # Cramer-von Mises
shapiro.test(x)$p.value           # Shapiro-Wilk (normality only)
```

**Shapiro–Wilk** : most powerful normality test for $n \leq 2000$:
$$W = \frac{\left(\sum a_i X_{(i)}\right)^2}{\sum(X_i-\bar X)^2} \in (0,1]$$

Values close to 1 $\Rightarrow$ consistent with normality.

---

## Comparison Table

| Test | Null $H_0$ | Needs continuous $F$? | Params known? | Tail focus |
|------|-----------|:-:|:-:|:-:|
| $\chi^2$ GoF | Any (binned) | No | Yes/No | Low |
| KS | Specific $F_0$ | **Yes** | **Yes** | Middle |
| Lilliefors | Normal/Exp | Yes | **No** | Middle |
| Cramér–von Mises | Specific $F_0$ | Yes | Yes | Uniform |
| Anderson–Darling | Specific $F_0$ | Yes | Yes | **High** |
| Shapiro–Wilk | **Normality only** | Yes | No | **High** |

**Practical guide:**
- Normality, small $n$: **Shapiro–Wilk**
- Normality, unknown params: **Lilliefors** or **Anderson–Darling**
- Fully specified $F_0$: **KS** or **Cramér–von Mises**
- Discrete data: **$\chi^2$ GoF**

---

## Summary

.pull-left[
**Order statistics** $X_{(1)} \leq \cdots \leq X_{(n)}$

**Ranks** strip values to relative order — basis of nonparametric inference

**EDF** $S_n(x)$ estimates $F_X$; Glivenko–Cantelli gives uniform convergence

**PIT** $X \sim F \Rightarrow F(X) \sim U[0,1]$: powers both RV generation and the KS distribution-free property
]

.pull-right[
**KS test** $D_n = \sup_x|S_n(x)-F_0(x)|$ requires $F_0$ continuous and **fully specified**

**Simulation-based p-values**: generate samples from $F_0$, compute $D_n$ each time, see where observed $D_n$ lands; exact for any $n$

**Lilliefors** corrects KS for estimated parameters

**AD / SW** offer greater power for normality testing
]

---

## Exercises

1. Derive the density of $X_{(3)}$ for $n=10$, $X_i \sim \text{Exp}(2)$. Verify by simulation.

2. Given 10 observations $(0.12, 0.28, 0.37, 0.45, 0.52, 0.61, 0.70, 0.79, 0.88, 0.95)$, test $H_0: F = U(0,1)$ by computing $D_{10}$ by hand.

3. Why does plugging in $\hat\mu, \hat\sigma$ into the KS test make it **conservative**? Which direction does the actual type I error deviate from $\alpha$?

4. Write R code to compute a simulation-based p-value for testing $H_0: F = N(5, 4)$ using $n=25$ observations and $B=5000$ replicates.

5. Generate 30 observations from $t_3$. Apply KS, Lilliefors, AD, and Shapiro–Wilk for normality. Which is most powerful here? Why?
