---
title: "Matrix Operations in R"
author: "Jyotishka Datta"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
  css: mystyle.css
lib_dir: libs
nature:
  highlightStyle: github
highlightLines: true
countIncrementalSlides: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = TRUE)
```

class: middle
count: false

## Introduction

- Matrix in R tutorial by Prof. Giovanni Petris. 
- `doParallel` guide by Weston and Calaway available [here](https://cran.r-project.org/web/packages/doParallel/vignettes/gettingstartedParallel.pdf). 
- Miscellaneous internet resources. 
- Before you run a module or chunk, make sure that you have all the packages installed. 
- We need the R package 'Matrix' and 'doParallel'.

---
## Matrices 

- To arrange values into a matrix, we use the `matrix()` function:
```{r, echo = TRUE}
a <- matrix(1 : 6, nrow = 2, ncol = 3)
a
```
- Individual entries can be referred to using a pair of indices. For example, the element in the second row, third column can be printed as:

```{r, echo = TRUE}
a[2, 3]
```

---
## Rows, Columns 
-  An entire row or column can be retrieved by specifying its index and leaving the other index empty:

```{r, echo = TRUE}
a[2, ]
```
-  Similarly to the way R works with vectors, you can use negative indices to exclude row or colums of a matrix:

```{r, echo = TRUE}
a[-2,]
```

---
## Matrix indices 

- Note how the values 1,2,3,4,5,6 are used to create the matrix: the first two are used to fill the first column, then the next two to fill the second column, and so on.

- R allows matrices to be indexed by a single number, e.g.,

```{r, echo = TRUE}
a[5]
```
- The fifth element of a is the fifth element of the vector obtained by "unrolling" the matrix, column by column. 

---
## Fill by row 

-  Sometimes you need to fill a matrix row by row, instead than column by column. 
-  You can change the default behavior as follows:

```{R,echo = TRUE}
a <- matrix(1 : 6, nrow = 2, ncol = 3, byrow = TRUE)
a
```

---
## Diagonals 

- We have seen rbind() and cbind() before. 
- e.g. We can use cbind() to construct a 3 x 3 Hilbert matrix. The (i, j) entry is
1/(i + j - 1).
```{r, echo = TRUE}
H3 <- 1 / cbind(1 : 3, 2 : 4, 3 : 5)
H3
```

---
## Diagonals and Determinants

- The diagonal of a matrix can be extracted using `diag()`.
```{r, echo = TRUE}
diag(H3)
```
- The determinant of a square matrix can be obtained with `det()`.

```{r, echo = TRUE}
det(H3)
```


---
## Triangular 

-  The functions `lower.tri()` and `upper.tri()` can be used to obtain the lower and upper triangular parts of matrices. 
The output of the functions is a matrix of logical elements, with TRUE representing the relevant triangular elements. For example,

```{r, echo = TRUE}
lower.tri(H3)
```

---
## Triangular 

- A typical use of these functions is to set the upper or lower triangular part of a matrix to zero, thus constructing a triangular matrix.

```{r, echo = TRUE}
Htri <- H3
Htri[lower.tri(Htri)] <- 0
Htri
```

---
## Multiplication 

- Matrix multiplication in R is performed by the operator %*%. For example,
```{r, echo = TRUE}
A <- matrix(c(1, 2, 0, 1), 2)
B <- matrix(c(1, 3, 3, 1, 4, 5), nr = 2)
A
B
```
- Matrices must conform, i.e. multiplication must make sense !

---
## Multiplication 

```{r, echo = TRUE}
A %*% B
dim(A)
ncol(A)
nrow(B)
```


---
## Multiplication 

With $A$ and $B$ defined as above the matrix product $BA$ in not defined, but the product $B^T A$ is ($B^T$ denotes the transpose of $B$). 
- The transpose of a matrix can be obtained with the function `t()`.
```{r, echo = TRUE}
t(B) %*% A
```


---
## Crossprod

-  A much more efficient way of computing the same matrix product is via the function `crossprod()`.
```{r, echo = TRUE}
all.equal(crossprod(B, A), t(B) %*% A)
```
- A similar function, tcrossprod(), computes the matrix product $AB^T$ whenever the product is well defined.


---
## Exercise 

- Consider the matrix
```{r, echo = TRUE}
X <- cbind(1, seq(-1, 1, length = 11))
X
```
- Use crossprod() and tcrossprod() to find $X^T X$ and $X X^T$. 
- Note the dimensions of the resulting matrices.


---
## Matrix inversion

- The inverse of a matrix can be found using `solve()`:
```{r, echo = TRUE}
Ainv <- solve(A)
Ainv
all.equal(Ainv %*% A, diag(2))
```


---
## Solve linear system 

- Inverse in R uses QR decomposition, that we might briefly talk about at the very end (if time permits)! 

- For linear systems we don't need to compute inverses. We can use Gaussian Elimination. 

```{r, echo = TRUE}
b <- c(5, 3)
solve(A, b) # use this
Ainv %*% b # don't use this
```

---
## Time comparison 

- Compare execution times for the different methods of solving the Least Squares problem when the number of observations is large.

-  Let us start by making up our own random data: $n \times p$ design matrix $X$ and an $n \times 1$ response vector $y$. 
-  Take $n = 5 \times 10^5$. 

```{r, echo = TRUE, cache = TRUE}
set.seed(123)
n <- 5e5; p <- 20
X <- matrix(rnorm(n * p, mean = 1 : p, sd = 10), nr = n, nc = p, byrow = TRUE)
y <- rowSums(X) + rnorm(n)
```

---
## OLS 

- The naive approach to solve the LS problem is to compute: 
$\hat{\beta} = (X^T X)^{-1}X^T y$

```{r, echo = TRUE, cache = TRUE}
system.time(bHat1 <- solve(t(X) %*% X) %*% t(X) %*% y)
```

-  Not so naive: use `crossprod`.

```{r, echo = TRUE, cache = TRUE}
system.time(bHat2 <- solve(crossprod(X), crossprod(X, y)))
```

---
class: middle
count: false

# More advanced topics  


---
## Inverting a matrix using Cholesky helps

- A symmetric positive definite matrix can be factorized as: 
$$
A = L L^T
$$

```{r, echo = TRUE, cache = TRUE}
p = 1000; A = array(rnorm(p*p), c(p,p))
A = crossprod(A)
ptm <- proc.time(); B = solve(A); proc.time()-ptm
ptm <- proc.time(); B = chol2inv(A); proc.time()-ptm
```

---
## Storage of sparse matrices 

- Sparse matrices appear in many areas of Statistics and Machine learning. 
  1. For large Markov chain, the transition matrix becomes sparse. 
  2. Large design matrix / contingency tables. 
  3. Modern dataset with 0 as imputation for missing values. 

- If matrix $A_{n\times n}$ has average $k$ non-zero entries in each row and $k << n$, then storage reduces from $O(n^2)$ to $O(n)$. 


---
## Storage of unstructured sparse matrices 

```{r, echo = TRUE, cache = TRUE}
# Demo 2: saving sparse matrices in R can save memory
library(Matrix)
m1 = matrix(0, nrow=1000, ncol=1000)
m2 = Matrix(0, nrow=1000, ncol=1000, sparse=TRUE)
object.size(m1)
object.size(m2)
```

