---
title: "Nonparametric Statistics"
subtitle: "Introduction: Lecture 0"
author: "Jyotishka Datta"
institute: "Virginia Tech"
date: today
format:
  revealjs:
    theme: default
    slide-number: true
    progress: true
    chalkboard: true
    html-math-method: mathjax
    transition: slide
    background-transition: fade
    footer: "STAT 3504 | Virginia Tech | Jyotishka Datta"
    scrollable: true
    smaller: true
execute:
  echo: false
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

## About Me

-   Associate Professor at VT Statistics.
-   Research area: Statistics / Machine Learning / (Applied)
    Probability.
    -   Statistical methods for high-dimensional data with structures.
    -   Applications: human genomics, crime forecasting, infectious
        disease.
-   Born in Calcutta (Kolkata), India. Attended Indian Statistical
    Institute.
-   Mumbai, India — Barclays Bank, PLC — modelling credit card default
    and fraud detection.
-   Purdue University: Ph.D. in Statistics.
-   Postdoc at Duke University and SAMSI (Statistics and Applied
    Mathematical Sciences Institute).

# Syllabus {background-color="#861F41"}

## Outline of This Course — Part I

**Foundations and One/Two-Sample Methods**

1.  Why nonparametric? Parametric vs. distribution-free inference;
    Pitman efficiency
2.  Permutation tests: the unifying idea behind most of what follows
3.  Simulation and resampling, basics of Monte Carlo, Bootstrap
    hypothesis tests
4.  Ranks and order statistics (just enough theory to understand what
    follows)
5.  One-sample and paired-sample tests: sign test, Wilcoxon signed-rank
    test
6.  Two-sample tests: Mann–Whitney, Wilcoxon rank-sum, and
    permutation-based alternatives

::: callout-note
If you have taken CMDA 2006 with me, some of these will feel familiar,
but the treatment will be faster and more theoretical throughout. The
two courses share some common material alongside topics unique to each.
This is only a tentative list of topics and may change to accommodate
this class.
:::

## Outline of This Course — Part II

**More Advanced Topics**

1.  Rank-based linear models — ranks inside `lm()`, Spearman and Kendall
    correlation
2.  Goodness-of-fit tests: KS, Lilliefors, Anderson–Darling
3.  Kernel Density Estimation and $k$-nearest neighbors
4.  Survival Analysis: Kaplan–Meier estimator and log-rank test

Implement all of the above in `R`, with and without packages

::: callout-note
If you have taken CMDA 2006 with me, some of these will feel familiar,
but the treatment will be faster and more theoretical throughout. The
two courses share some common material alongside topics unique to each.
This is only a tentative list of topics and may change to accommodate
this class.
:::

## How to Reach Me

-   **Office:** Hutcheson 409C
-   **Email:** jyotishka\@vt.edu\
    (Subject line: STAT 3504 + \[Your query\])
-   **Class time:** TR 2:00–3:15 pm
-   **Office Hours:** Will be updated on Canvas
-   **Graduate Teaching Assistant (GTA):** Jacob Gareis
    -   Email: jacobgareis\@vt.edu
    -   Office Hours: Will be updated on Canvas

## Grades

**Grade Distribution:**

-   Two statistics midterms: 25% each
-   Homework: 25%
-   In-class activities / exercises: 25%
-   Final Exam: 25%

| Interval  | Grade | Interval  | Grade |
|-----------|-------|-----------|-------|
| \[93, ∞)  | A     | \[73, 77) | C     |
| \[90, 93) | A−    | \[70, 73) | C−    |
| \[87, 90) | B+    | \[67, 70) | D+    |
| \[83, 87) | B     | \[63, 67) | D     |
| \[80, 83) | B−    | \[60, 63) | D−    |
| \[77, 80) | C+    | \[0, 60)  | F     |

## Homework Rules

-   Homework assignments will be frequent and will require using R.
-   Extensions will **not** be granted. Assignments accepted up to 24
    hours late with a 20% penalty.
-   You may discuss assignments with others, but **what you turn in must
    be your own work**.
-   No sharing of written work or solutions.
-   We grade **solutions, not answers**. Show your work. Answers without
    justification receive no credit.
-   All homework will have one week to submit unless otherwise
    announced.
-   The **lowest homework grade** will be dropped.

## Policy on Use of Generative AI {.smaller}

::: callout-warning
## Generative AI Policy

You may use tools such as ChatGPT or Copilot to **support learning, not
replace your authorship**. When I grade an assignment, I should be
grading **your** work.

**Permitted:** Brainstorm, outline, debug, revise, or generate examples
that you verify and rewrite. Use AI only when the assignment allows it.

**Not permitted:** Submitting AI text or code as your own; using AI on
quizzes or exams unless explicitly allowed; entering confidential or
restricted data.

**Disclosure:** If you used AI, add a one-line "AI use" note at the end
listing the tool, what you asked, and how you changed or verified the
output.\
*Example:* "AI use: ChatGPT for an outline; I rewrote sections and
checked citations."

**Accountability:** You are responsible for accuracy, originality, and
citation. Work may be checked through follow-up questions or originality
tools. Your voice and reasoning must remain prominent.
:::

## My personal take

![A slide for Terence Tao's talk: Mathematics in the age of
AI](images/Screenshot 2026-08-06 114704.png){fig-alt="A slide for Terence Tao's talk: Mathematics and AI"}

## My personal take (contd.)

![Slide from the same talk showing Leiden declaration
(<https://leidendeclaration.ai/>)](images/Screenshot 2026-08-06 114818.png){fig-alt="Slide from Terence Tao's talk: Leiden declaration"}

## Goodhart's Law

![Goodhart's Law (Source: [sketchplanation
blog](https://sketchplanations.com/goodharts-law "https://sketchplanations.com/goodharts-law"))](images/goodharts_law.avif){fig-alt="An image showing Goodhart's Law"}

## Back to logistics: References

-   No required textbook. Following the lecture slides should be
    sufficient for tests.
-   Optional:\
    1. *Nonparametric Statistical Inference***, 5th ed.** Gibbons &
    Chakraborti\
    2. *Monty the Null Hippopotamus, Exploring Statistical Foundations
    Through Simulation,* Robert B. Gramacy. Chapman
    Hall/CRC. <https://bobby.gramacy.com/hipp0/>
-   Other references:
    1.  Higgins. *Introduction to Modern Nonparametric Statistics.*
    2.  Hollander and Wolfe. *Nonparametric Statistical Methods.*
    3.  Sprent and Smeeton. *Applied Nonparametric Statistical Methods.*
    4.  Conover. *Practical Nonparametric Statistics.*
-   You are responsible for taking notes, including from class
    discussions you may have missed.

## Software

-   We will use **R** throughout the course. Install it on your laptop
    (or computing resource of choice).
-   Available free from: <http://cran.r-project.org/>
-   Recommended interface: **RStudio** — <https://www.rstudio.com/>
-   Useful online resources:
    1.  ATS UCLA R Page: <http://statistics.ats.ucla.edu/stat/r/>
    2.  Quick R: <http://www.statmethods.net/>
    3.  R-Bloggers: <https://www.r-bloggers.com/>
    4.  R for Data Science: <http://r4ds.had.co.nz/>

# Introduction {background-color="#861F41"}

## Parametric and Nonparametric Statistics

-   **Parameter:** A constant (usually unspecified) that characterizes a
    population: broadly includes any description of population
    characteristics.
-   **Statistic:** A function of observations that does not depend on
    unknown parameters.
-   **Inference:** Using statistics to infer something about the
    population: mainly estimation and hypothesis testing. What makes it
    *scientific* is the ability to make statements about accuracy or
    reliability.

. . .

-   Some statistics are more popular than others, e.g. sample mean vs.
    median. **Why?**

. . .

-   **Central Limit Theorem:** The sample mean from a large sample is
    approximately normally distributed.

## Central Limit Theorem

-   This is the reason we see Normal distributions everywhere!
-   The **Central Limit Theorem** is one of the most important results
    in all of mathematics.
-   The sum of $N$ independent, identically distributed random variables
    is approximately **Normally distributed**.
-   Example: if you flip a coin many times, the probability of getting a
    given number of heads follows a Normal curve with mean equal to half
    the total number of flips (De Moivre's classic result).

<https://jdatta.shinyapps.io/demoivre/>

## General CLT {.center}

![Illustration of the Central Limit Theorem — details in
slide](art/clt-1.png){fig-alt="Details in slide" width="90%"}

##  {.center}

![Statistical cartoon — details in
slide](art/cartoon16.jpg){fig-alt="Details in slide" width="70%"}

# Parametric versus Nonparametric {background-color="#861F41"}

## Parametric Inference

-   **Parametric Inference:** Estimation and inference based on an
    assumed form of the distribution.
-   **Example** (modified from G&C 4.13): Scores on a certain IQ test
    are assumed normally distributed,
    $X_i \sim N(\mu,\, \sigma^2 = 100)$. Eighteen girls were tested with
    scores:\
    114, 81, 87, 114, 113, 87, 111, 89, 93, 108, 99, 93, 100, 95, 93,
    95, 106, 108.
-   **Question:** Is the mean IQ significantly greater than 100?
-   **Hypotheses:** $H_0: \mu = 100$, $H_A: \mu > 100$
-   We can carry out a $z$-test based on the normality assumption.

. . .

-   Conclusions are valid **as long as the assumptions are valid.**

## Questions to Ask About Any Test

-   What are the assumptions behind a particular test?

. . .

-   For example, the $z$-test assumes IID, normally distributed data and
    known variance.

. . .

-   How do we test if those assumptions are valid?\
    → **Goodness-of-fit tests**, e.g. Lilliefors or Shapiro–Wilk tests
    for normality.

. . .

-   Can we make other, weaker assumptions? (e.g. symmetry or
    independence?)

. . .

-   How do we test the hypothesis **without** the distributional
    assumptions?\
    → **Nonparametric alternatives.**

. . .

-   You can also ask a more basic question: are the numbers really
    random? There are methods for that too.

## One-Sample $Z$-Test

-   **Assumption:**
    $X_1, X_2, \ldots, X_n \overset{\text{iid}}{\sim} N(\mu,\, \sigma^2)$,
    $\mu$ unknown, $\sigma^2$ known.
-   **Null hypothesis:** $H_0: \mu = \mu_0$
-   **Test statistic:**

$$Z = \frac{\bar{X} - \mu_0}{\sigma / \sqrt{n}}$$

Under $H_0$, $Z \sim N(0, 1)$.

-   If the observed $Z$ is "too large," it is unlikely to have come from
    $N(0,1)$.
-   We control how large is "too large" by fixing the **Type I error**:

$$\text{Type I error} = P(\text{Reject } H_0 \mid H_0 \text{ true}) \leq \alpha$$

## Z-Test, Continued

$$Z = \frac{\bar{X} - \mu_0}{\sigma / \sqrt{n}}$$

-   P-values are exact as long as $\bar{X}$ is normally distributed —
    guaranteed when $X_i \sim N(\mu, \sigma^2)$, giving
    $\bar{X} \sim N(\mu,\, \sigma^2/n)$.

-   But this is a strong assumption! What if $X_i$'s are bounded,
    discrete, or skewed?

. . .

-   **The answer is "yes, you can still apply these tests" — if your
    sample size is large.**

-   You don't need individual $X_i$'s to be normal, as long as $\bar{X}$
    is **approximately** normal.

-   This is what the CLT guarantees — but you need a large enough
    sample, and the CLT conditions must hold.

## Tests Based on Normality

There is nothing exceptional about the $z$-test.

-   If $\sigma^2$ is unknown and estimated by $S^2$ → $t$-test
-   Single sample → **paired** $t$-test → **two-sample** $t$-test →
    **ANOVA (**$K$ samples)

All of these are parametric tests that are:

-   **Exact** under normality
-   **Approximately correct** for large samples (by CLT)

So one good reason to look for alternative methods is **small sample
size**.

## Sample Size Requirements for Parametric Tests with Non-Normal Data {.smaller}

Parametric tests can still be used with non-normal data, given
sufficient sample size:

| Parametric Analysis          | Sample Size Required (Non-Normal Data) |
|------------------------------|----------------------------------------|
| 1-sample $t$-test            | $n > 20$                               |
| 2-sample $t$-test            | Each group: $n \geq 15$                |
| One-Way ANOVA (2–9 groups)   | Each group: $n \geq 15$                |
| One-Way ANOVA (10–12 groups) | Each group: $n \geq 20$                |

But small sample size is **not the only** reason to switch to
nonparametric methods.

## Non-Parametric Methods

-   The main attraction: **fewer assumptions** about the underlying
    distribution.
-   **Advantage:** Robustness (inference remains valid under model
    misspecification.)

. . .

-   **Price to pay:** If distributional assumptions are true, parametric
    methods often have more **power**.
    -   If data truly come from $N(\mu, \sigma^2)$ and you use a test
        based on that, you have a higher chance of correctly detecting
        the alternative.

. . .

-   On the other hand, if assumptions are *wrong* (e.g. data follow a
    Cauchy but you assume Normal), nonparametric methods can be
    *infinitely more* powerful.

. . .

> "*The basic idea of nonparametric inference is to use data to infer an
> unknown quantity while making as few assumptions as possible.*"

## Distribution-Free?

-   "Nonparametric" is actually a **misnomer**. A more accurate term:
    **Distribution-Free Statistics**.
-   Some prefer the term **infinite-dimensional inference**.
-   We will still estimate and test hypotheses about parameters but the
    **form of the distribution is not assumed**.

. . .

-   Often we only assume that random variables are **independent and
    identically distributed (IID)**.

. . .

-   Test and estimation procedures require **relatively fewer
    assumptions** about the population distribution.

# History of Nonparametric Statistics {background-color="#861F41"}

## The First Nonparametric Test

These anecdotes are mainly from Sprent & Smeeton's textbook and *The
Lady Tasting Tea* by David Salsburg.

The first chapter of the Book of Daniel records that on the orders of
Nebuchadnezzar certain favoured children were to be specially fed on the
king's meat and wine for 3 years. Reluctant to defile himself with such
luxuries, Daniel pleaded that he and three of his brethren be fed
instead on pulse for 10 days. After that time the four were declared
'fairer and fatter in flesh than all of the children which did eat the
portion of the king's meat'. This evidence was taken on commonsense
grounds to prove the superiority of a diet of pulse.

Although the biblical analysis is informal, it contains the seed of a
nonparametric test.

## Early History

**John Arbuthnot (1710)** observed that in each year from 1629 to 1710,
the number of males christened in London exceeded the number of females.
He used this to test whether the birth ratio was 1:1. (**Binomial
test?**)

Data: <https://www.openintro.org/data/index.php?data=arbuthnot>

**Francis Galton (1892)** developed a measure, which he called a
"centissimal scale", to assess agreement between patterns (categorical
data) on corresponding fingertips on left and right hands.

**Karl Pearson (1900)** proposed the well-known **chi-squared
goodness-of-fit test** for any discrete distribution.

**C. Spearman (1904)** defined the **rank correlation coefficient** that
bears his name.

## Wilcoxon and Mann–Whitney

-   **Frank Wilcoxon**, working at American Cyanamide in the 1940s, was
    frustrated by the sensitivity of $t$-tests and ANOVA to outliers.
-   He developed a nonparametric test that did not rely on distribution
    parameters, published in **Biometrics (1945)**.
-   He reportedly submitted the paper hoping it would be rejected so
    referees could point out his mistakes but it was accepted, because
    no such methods had existed before.

. . .

-   Unbeknownst to Wilcoxon, economists Henry B. **Mann** and graduate
    student D. Ransom **Whitney** were working on similar problems.
-   They published in **1947**, establishing a connection to Wilcoxon's
    work.
-   You will learn these as the **Wilcoxon rank-sum test** and
    **Mann–Whitney U test**.

## Kolmogorov and Smirnov

-   In the **1930s**, Andrei **Kolmogorov** and his student **Smirnov**
    developed a different test for comparing two distributions without
    using any parameters.
-   This is the **KS test**, which you will study in detail.

Both tests allowed analysis **without estimating distribution
parameters**, marking a shift in statistical methodology from Pearson's
era.

## Pitman Efficiency and ARE

-   After Wilcoxon, Mann, and Whitney, some questions remained:

    -   If data truly follow a known parametric distribution (like
        Normal), how bad is it to ignore normality and use nonparametric
        tests instead?

-   **Pitman (1948)** addressed this, demonstrating that nonparametric
    tests perform better with **slight deviations** from parametric
    models, and suggested a preference for nonparametric tests in
    hypothesis testing.

-   We will study **Pitman Efficiency** and **Asymptotic Relative
    Efficiency (ARE)**.

::: callout-tip
The large-sample efficiency of the Wilcoxon test compared to the
$t$-test is $\dfrac{3}{\pi} \approx 0.955$.
:::

## Further Developments

-   The field grew in the **1960s** with researchers exploring various
    nonparametric tests.
-   **Jaroslav Hájek** published a unifying textbook in **1971**,
    linking nonparametric tests to central limit theorem conditions.
    Much of the theory of linear rank tests is due to Hájek.

**This course covers everything up to this point.**

## Challenges: The Bahadur–Savage Result

-   The first major challenge to nonparametric methodology came from
    **Raj Bahadur** at the University of Chicago.
-   **R.R. Bahadur and L.J. Jimmie Savage (1956)** identified issues
    with outliers affecting nonparametric test reliability.

> "*If outliers are **rare and completely wrong observations**, then
> nonparametric methods **reduce their influence** on the analysis. If
> outliers are **part of a systemic contamination** of the data, then
> shifting to nonparametric will **make matters worse**.*"

I will describe the Bahadur–Savage result because it is striking and
beautiful but the mathematical tools needed to prove it are
(unfortunately!) beyond the scope of this course.

## Survival Analysis

-   Another widely used nonparametric method: **Kaplan–Meier survival
    analysis**.
-   **Survival Analysis:** A set of methods for analysing data where the
    outcome variable is the **time until the occurrence of an event of
    interest**.
    -   Events: death, disease occurrence, equipment failure, marriage,
        etc.
    -   Literally anything where we ask: *How long does it take to…?*

. . .

![Kaplan-Meier citation count — details in
slide](art/KM_cite.jpg){fig-alt="Details in slide" height="180px"}

. . .

…and it's a **nonparametric method!**

## Survival Analysis, Continued {.smaller}

What's the second most cited article?

::::: columns
::: {.column width="50%"}
![Top 100 most cited papers — details in
slide](art/top100_cited.jpg){fig-alt="Details in slide" width="90%"}
:::

::: {.column width="50%"}
![Cox's proportional hazard paper — details in
slide](art/Stiegler.jpg){fig-alt="Details in slide" width="90%"}
:::
:::::

-   Later in the course, we will learn the **Bootstrap**.
-   At least 10 papers in this list use Bootstrap or a version of it
    applied to biological data.
-   The most well-known: *Confidence limits on phylogenies: an approach
    using the bootstrap*, Felsenstein (1985). Cited by 37,710.

# Parametric vs. Nonparametric {background-color="#861F41"}

## Parametric Inference: the IQ Example

**Example:** Scores on a certain IQ test are normally distributed with
$\sigma^2 = 100$. Eighteen girls of similar age had scores:

$$114,\ 81,\ 87,\ 114,\ 113,\ 87,\ 111,\ 89,\ 93,\ 108,\ 99,\ 93,\ 100,\ 95,\ 93,\ 95,\ 106,\ 108$$

**Question:** Is the mean IQ significantly greater than 100?

$H_0$: $\mu = 100$ $\quad$ $H_A$: $\mu > 100$

→ Carry out a $z$-test based on the normality assumption.

Conclusions are valid **as long as the assumptions are valid.**

## IQ Example — Nonparametric Version

-   The IQ scores of 18 girls were:\
    114, 81, 87, 114, 113, 87, 111, 89, 93, 108, 99, 93, 100, 95, 93,
    95, 106, 108.
-   **We do not assume the scores are Normal, but we do assume the
    scores are IID.**

. . .

-   **Changed Question:** Is the ~~mean~~ **median** IQ significantly
    greater than 100?

. . .

-   We can use **one-sample nonparametric tests** (first module of this
    course).

## Why Study Nonparametric Statistics?

-   In many applications, there is **no prior knowledge** of the
    underlying distribution.
-   If parametric assumptions are violated, parametric test procedures
    can give **misleading or wrong results**.
-   For studies with **small sample sizes (**$n \leq 30$), Normal
    approximations do not work well.

. . .

-   Therefore, we need methods that require **very little
    model/distributional assumptions**,

. . .

-   or methods that are **robust/insensitive** to model/distributional
    assumptions,

. . .

-   and that are **insensitive to outliers**.

. . .

> **Nonparametric Statistics offers a solution to all of these
> problems!**

# When to Use Nonparametric Methods? {background-color="#861F41"}

## Simplest Method: The Sign Test

-   With correct assumptions (e.g. normality), parametric methods are
    more efficient but often not by as much as you'd think.[^1]
-   Many nonparametric methods convert raw values to **ranks** and
    analyse those which can be "wasteful" if a valid parametric
    alternative exists.

[^1]: The large-sample efficiency of the Wilcoxon test vs. $t$-test is
    $3/\pi \approx 0.955$.

**The simplest nonparametric test: Sign test**

For the IQ scores of 18 girls (median = 100), roughly half of them
should lie above 100 and half below if the true median is 100.

-   Let $K_{100}$ = number of observations **above** 100.
-   **If** $K_{100}$ is too high or too low, the median is away from
    100.

## Sign Test – Calculation {.smaller}

| Obs  | 114 | 81  | 87  | 114 | 113 | 87  | 111 | 89  | 93  |
|------|-----|-----|-----|-----|-----|-----|-----|-----|-----|
| Diff | 14  | −19 | −13 | 14  | 13  | −13 | 11  | −11 | −7  |
| Sign | \+  | −   | −   | \+  | \+  | −   | \+  | −   | −   |

| Obs  | 108 | 99  | 93  | 100 | 95  | 93  | 95  | 106 | 108 |
|------|-----|-----|-----|-----|-----|-----|-----|-----|-----|
| Diff | 8   | −1  | −7  | 0   | −5  | −7  | −5  | 6   | 8   |
| Sign | \+  | −   | −   | \+  | −   | −   | −   | \+  | \+  |

$K_{100}$ = Total number of positives = **7 out of 18**.

. . .

Does this look too high or too low?

. . .

Calculate the probability of observing 7 or fewer heads in 18 tosses of
a fair coin — or just use R:

``` r
binom.test(7, 18)
```

## Sign Test — R Output

```         
binom.test(7, 18)

    Exact binomial test

data:  7 and 18
number of successes = 7, number of trials = 18, p-value = 0.4807
alternative hypothesis: true probability of success is not equal to 0.5
95 percent confidence interval: 0.1729859 0.6425488
sample estimates:
probability of success
             0.3888889
```

-   P-value =
    $P(\text{successes} \leq 7 \mid H_0\text{: Median IQ} = 100) = 0.4807$
-   We **fail to reject** $H_0$: not enough evidence of a difference.
-   The sign test is intuitive enough that anyone can understand the
    idea — we will revisit it in more detail.

## Parametric Inference (Arbuthnot Example)

**Example:** John Arbuthnot (1710) recorded the number of male and
female christenings in London for each year from 1629 to 1710. The data
are available at:
<https://www.openintro.org/data/index.php?data=arbuthnot>

```{r}
#| echo: true
arbuthnot <- read.csv("https://www.openintro.org/data/csv/arbuthnot.csv")
head(arbuthnot)
```

**Question:** Is the proportion of male christenings significantly
greater than 0.5?

**Hypotheses:** $H_0: p = 0.5$ $\quad$ $H_A: p > 0.5$

→ Carry out a one-sample $z$-test for a proportion based on the
normality assumption (CLT for large $n$).

Conclusions are valid **as long as the assumptions are valid.**

## Arbuthnot

-   In each year from 1629 to 1710, the number of male christenings
    **exceeded** the number of female christenings.
-   **We do not assume any particular distribution, but we do assume the
    yearly counts are IID.**

. . .

-   **Changed Question:** Is the probability that males outnumber
    females in a given year significantly greater than 0.5?

. . .

-   Let $K$ = number of years in which male christenings exceeded female
    christenings. Under $H_0: p = 0.5$, we expect $K \approx 41$ out of
    82 years.

## Sign test

-   We can use a **sign test** — or equivalently, an exact binomial
    test:

```{r}
#| echo: true
arbuthnot <- read.csv("https://www.openintro.org/data/csv/arbuthnot.csv")
K <- sum(arbuthnot$boys > arbuthnot$girls)
binom.test(K, nrow(arbuthnot), alternative = "greater")
```

## Sign Test — Drawback

::: r-fit-text
**Question:** Can you name one drawback of the Sign Test?
:::

. . .

::: r-fit-text
**Answer:** It completely ignores the **actual magnitudes**: might
result in low power (inefficient)!
:::

## Parametric Tests and Their Nonparametric Alternatives

| Parametric Test | Nonparametric Alternative |
|-----------------------------|-------------------------------------------|
| 1-sample $t$-test, Paired $t$-test | Sign test, Wilcoxon signed-rank test |
| 2-sample $t$-test | Mann–Whitney U test, Wilcoxon rank-sum test |
| One-Way ANOVA | Kruskal–Wallis test, Mood's median test |
| Factorial DOE | Friedman test |
| Pearson's correlation | Spearman's rank correlation or Kendall's $\tau$ |

## Other Nonparametric Methods

-   Most of these methods are very simple, in fact you can often derive
    them by **replacing an observation with its rank** (either within
    its own sample or a pooled sample).

-   I will show how many of these methods can be derived using ideas
    from **linear regression** via the simple `lm()` function in R after
    we have understood them as standalone methods.

-   Some popular nonparametric procedures are **not alternatives** but
    standalone tools. For example, the **Bootstrap**, which we will
    learn later.

-   There are also more advanced methods I won't be able to cover:
    nonparametric regression, wavelets, nonparametric Bayes (e.g.
    Dirichlet processes).

## Reasons to Use Nonparametric vs. Parametric Methods

::::: columns
::: {.column width="50%"}
**Reasons for Nonparametric**

1.  Small sample size *(double-edged sword!)*
2.  Ordinal data, ranked data, or outliers you cannot remove
3.  Data distribution doesn't follow any known parametric distribution
4.  Median is a better representative than mean (e.g. income, grade
    distributions)
5.  Usually simpler tests, less affected by departures from ideal
    assumptions — but limited in power and conclusions
:::

::: {.column width="50%"}
**Reasons for Parametric**

1.  Large sample — statistic is approximately Normal (CLT)
2.  Normality doesn't hold, but other parametric tests apply, or data
    can be suitably transformed
3.  Can draw more conclusions at the population level
:::
:::::

## What's Next

. . .

## Please also fill out the **survey about your R background** on Canvas.
