---
title: "Tests of Randomness: The Runs Test"
subtitle: "Lecture notes for STAT 3504"
author: "Jyotishka Datta"
institute: "Virginia Tech"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: show
    code-tools: true
    theme: cosmo
    df-print: paged
execute:
  warning: false
  message: false
  freeze: auto
header-includes: |
  <style>
  :root { --vt-maroon: #861F41; --vt-orange: #E5751F; }
  h1, h2, h3 { color: var(--vt-maroon); }
  .callout-important { border-left-color: var(--vt-orange); }
  </style>
---

# Why test randomness?

Many methods begin with an assumption that observations arrive in a random order or are serially independent. A dataset can have a reasonable histogram and still contain a strong sequential pattern.

Examples include:

- winning and losing days for a stock;
- defective and nondefective items on a production line;
- treatment successes and failures over time;
- measurements above and below a target;
- residuals from a fitted model.

A **runs test** asks whether the ordering contains too much clustering or too much alternation to be compatible with randomness.

::: {.callout-important}
A runs test is about **order**. If the observations are shuffled, their numerical values do not change, but the number of runs usually does.
:::

# Runs in a binary sequence

A **run** is a maximal consecutive block of identical symbols.

For example,

$$
+\,+\,+\;-\,-\;+\;-\,-\,-\;+\,+
$$

has the five runs

$$
(+++)\quad(--)\quad(+)\quad(---)\quad(++).
$$

Let

- $n_1$ be the number of symbols of type 1 (for example, $+$);
- $n_2$ be the number of symbols of type 2 (for example, $-$);
- $n=n_1+n_2$;
- $R$ be the total number of runs.

Interpretation:

- unusually **small** $R$: clustering or positive serial association;
- unusually **large** $R$: alternation or negative serial association;
- moderate $R$: compatible with random ordering.

# Hypotheses and null model

For the one-sample binary runs test, the null hypothesis is

$$
H_0:\text{ conditional on }n_1\text{ and }n_2,\text{ every ordering is equally likely.}
$$

Common alternatives are:

- lower-tailed: the sequence is clustered, so $R$ is too small;
- upper-tailed: the sequence alternates too often, so $R$ is too large;
- two-sided: the ordering is nonrandom in either direction.

Conditioning on $n_1$ and $n_2$ is important. Under $H_0$, there are

$$
\binom{n}{n_1}
$$

possible binary orderings, each with the same probability.

# Exact null distribution

The exact distribution depends only on $n_1$ and $n_2$.

For an even number of runs, $R=2k$,

$$
P(R=2k)
=
\frac{2\binom{n_1-1}{k-1}\binom{n_2-1}{k-1}}
{\binom{n}{n_1}}.
$$

For an odd number of runs, $R=2k+1$,

$$
P(R=2k+1)
=
\frac{
\binom{n_1-1}{k}\binom{n_2-1}{k-1}
+
\binom{n_1-1}{k-1}\binom{n_2-1}{k}
}
{\binom{n}{n_1}}.
$$

The possible values satisfy

$$
2\le R\le
\begin{cases}
2\min(n_1,n_2)+1, & n_1\ne n_2,\\
2n_1, & n_1=n_2.
\end{cases}
$$

If one symbol never occurs, then $R=1$ and no meaningful two-symbol runs test is possible.

## Why the formulas have this form

Suppose $R=2k$. The sequence must start with one symbol and end with the other, so each symbol forms exactly $k$ positive-length blocks. Splitting $n_1$ objects into $k$ nonempty blocks can be done in

$$
\binom{n_1-1}{k-1}
$$

ways, and similarly for the second symbol. The factor $2$ chooses which symbol starts the sequence.

For $R=2k+1$, the sequence starts and ends with the same symbol. One symbol has $k+1$ blocks and the other has $k$ blocks; the two terms correspond to which symbol has the extra block.

# Mean and variance under randomness

Under $H_0$,

$$
E(R)=\mu_R=1+\frac{2n_1n_2}{n},
$$

and

$$
\operatorname{Var}(R)=\sigma_R^2
=
\frac{2n_1n_2\left(2n_1n_2-n\right)}
{n^2(n-1)}.
$$

For sufficiently large $n_1$ and $n_2$,

$$
Z=\frac{R-\mu_R}{\sigma_R}
\mathrel{\dot\sim}N(0,1).
$$

A common classroom guideline is to prefer the exact distribution when either group is small and to regard the Normal approximation as comfortable when both $n_1$ and $n_2$ exceed about $10$. This is a rule of thumb, not a theorem.

## Optional continuity correction

Since $R$ is discrete, a continuity correction may be used:

$$
Z_{\mathrm{cc}}=
\begin{cases}
\dfrac{R+0.5-\mu_R}{\sigma_R}, & R<\mu_R,\\[8pt]
\dfrac{R-0.5-\mu_R}{\sigma_R}, & R>\mu_R,\\[8pt]
0, & R=\mu_R.
\end{cases}
$$

Software packages differ in whether they use a correction. Consequently, two implementations can report slightly different large-sample p-values.

# A complete hand calculation

Consider the sequence of stock-market directions over $20$ days:

$$
U\,U\,D\,D\,D\,U\,U\,U\,D\,U\,D\,D\,U\,U\,U\,U\,D\,D\,U\,D.
$$

The run decomposition is

$$
(UU)(DDD)(UUU)(D)(U)(DD)(UUUU)(DD)(U)(D),
$$

so

$$
R=10,\qquad n_1=11,\qquad n_2=9,\qquad n=20.
$$

The null mean and standard deviation are

$$
\mu_R=1+\frac{2(11)(9)}{20}=10.9,
$$

$$
\sigma_R
=
\sqrt{
\frac{2(11)(9)\{2(11)(9)-20\}}
{20^2(19)}
}
\approx2.1535.
$$

Without continuity correction,

$$
Z=\frac{10-10.9}{2.1535}\approx-0.4179,
$$

which gives a two-sided Normal p-value of approximately $0.676$. There is no evidence here of a nonrandom ordering.

```{r}
#| label: stock-example
seq_raw <- c("U","U","D","D","D","U","U","U","D","U",
             "D","D","U","U","U","U","D","D","U","D")

n1 <- sum(seq_raw == "U")
n2 <- sum(seq_raw == "D")
n  <- n1 + n2
R  <- length(rle(seq_raw)$lengths)

mu_R <- 1 + 2*n1*n2/n
var_R <- 2*n1*n2*(2*n1*n2 - n)/(n^2*(n - 1))
sd_R <- sqrt(var_R)
z <- (R - mu_R)/sd_R
p_normal <- 2*pnorm(-abs(z))

c(n1 = n1, n2 = n2, runs = R,
  expected_runs = mu_R, sd_runs = sd_R,
  z = z, p_value = p_normal)
```

::: {.callout-note}
The variance denominator is $n^2(n-1)$. Writing the algebra incorrectly can produce an impossible standard deviation larger than the entire range of $R$.
:::

# Base-R implementation

The following functions count runs, compute the exact null distribution, and perform exact or Normal tests without requiring an external package.

```{r}
#| label: runs-functions
count_runs <- function(x) {
  if (length(x) == 0) return(0)
  1 + sum(x[-1] != x[-length(x)])
}

runs_moments <- function(n1, n2) {
  n <- n1 + n2
  if (n1 < 1 || n2 < 1) stop("Both symbol types must occur.")
  mu <- 1 + 2*n1*n2/n
  variance <- 2*n1*n2*(2*n1*n2 - n)/(n^2*(n - 1))
  c(mean = mu, variance = variance, sd = sqrt(variance))
}

runs_pmf <- function(n1, n2) {
  if (n1 < 1 || n2 < 1) stop("Both symbol types must occur.")
  n <- n1 + n2
  choose0 <- function(a, b) {
    if (a < 0 || b < 0 || b > a) 0 else choose(a, b)
  }
  r_max <- if (n1 == n2) 2*n1 else 2*min(n1, n2) + 1
  r_values <- 2:r_max
  denom <- choose(n, n1)

  probability <- vapply(r_values, function(r) {
    if (r %% 2 == 0) {
      k <- r/2
      2*choose0(n1 - 1, k - 1)*choose0(n2 - 1, k - 1)/denom
    } else {
      k <- (r - 1)/2
      (choose0(n1 - 1, k)*choose0(n2 - 1, k - 1) +
       choose0(n1 - 1, k - 1)*choose0(n2 - 1, k))/denom
    }
  }, numeric(1))

  data.frame(runs = r_values, probability = probability)
}

runs_test_binary <- function(x,
                             alternative = c("two.sided", "less", "greater"),
                             method = c("exact", "normal"),
                             continuity = FALSE) {
  alternative <- match.arg(alternative)
  method <- match.arg(method)
  if (length(unique(x)) != 2L) stop("x must contain exactly two symbol types.")

  x <- as.factor(x)
  counts <- table(x)
  n1 <- unname(counts[1L])
  n2 <- unname(counts[2L])
  R <- count_runs(x)
  moments <- runs_moments(n1, n2)

  if (method == "exact") {
    tab <- runs_pmf(n1, n2)
    lower <- sum(tab$probability[tab$runs <= R])
    upper <- sum(tab$probability[tab$runs >= R])
    p <- switch(alternative,
      less = lower,
      greater = upper,
      two.sided = min(1, 2*min(lower, upper))
    )
    statistic <- R
  } else {
    shift <- 0
    if (continuity && R < moments["mean"]) shift <- 0.5
    if (continuity && R > moments["mean"]) shift <- -0.5
    z <- (R + shift - moments["mean"])/moments["sd"]
    p <- switch(alternative,
      less = pnorm(z),
      greater = pnorm(z, lower.tail = FALSE),
      two.sided = 2*pnorm(-abs(z))
    )
    statistic <- z
  }

  list(
    method = method,
    alternative = alternative,
    n1 = n1,
    n2 = n2,
    runs = R,
    null_mean = unname(moments["mean"]),
    null_sd = unname(moments["sd"]),
    statistic = unname(statistic),
    p_value = unname(p)
  )
}
```

Apply both methods to the stock sequence:

```{r}
runs_test_binary(seq_raw, method = "exact")
runs_test_binary(seq_raw, method = "normal")
runs_test_binary(seq_raw, method = "normal", continuity = TRUE)
```

## A note on two-sided exact p-values

For a discrete statistic, “two-sided” is not unique. Common conventions include:

1. twice the smaller tail probability;
2. summing probabilities of outcomes whose null probability is no larger than that of the observed outcome;
3. ordering outcomes by distance from $E(R)$.

The function above uses twice the smaller tail, capped at $1$. A software package may use a different convention, so exact two-sided p-values can differ slightly even when all calculations are correct. One-sided exact p-values are unambiguous once the direction is specified.

# Visualizing runs

```{r}
#| label: visualize-runs
#| fig-width: 9
#| fig-height: 3.8
#| echo: false
x <- ifelse(seq_raw == "U", 1, 0)
cols <- c("#E5751F", "#861F41")
plot(seq_along(x), x, type = "s", lwd = 3, col = "#861F41",
     xaxt = "n", yaxt = "n", ylim = c(-0.25, 1.25),
     xlab = "Day", ylab = "Direction", main = "Runs in the observed sequence")
axis(1, at = seq_along(x))
axis(2, at = c(0, 1), labels = c("D", "U"), las = 1)
points(seq_along(x), x, pch = 21, bg = cols[x + 1], col = "white", cex = 1.6)
abline(v = which(diff(x) != 0) + 0.5, col = "gray65", lty = 3)
```

Vertical dotted lines mark changes from one run to the next.

# Simulating the null distribution

Under $H_0$, we can hold the counts fixed and randomly permute the sequence.

```{r}
#| label: null-simulation
#| fig-width: 8
#| fig-height: 4.8
set.seed(3504)
n1 <- 10
n2 <- 10
base_sequence <- c(rep(1, n1), rep(0, n2))

R_null <- replicate(20000, count_runs(sample(base_sequence)))

hist(R_null, probability = TRUE,
     breaks = seq(min(R_null) - 0.5, max(R_null) + 0.5, by = 1),
     col = "#861F41", border = "white",
     xlab = "Number of runs", main = "Permutation null distribution")

mom <- runs_moments(n1, n2)
curve(dnorm(x, mom["mean"], mom["sd"]), add = TRUE,
      col = "#E5751F", lwd = 3)
```

This simulation is not needed when the exact combinatorial distribution is available, but it makes the null model concrete: the values are fixed, and only their order is randomized.

# Contrasting three sequences

```{r}
#| label: sequence-contrast
set.seed(12)
clustered <- c(rep(1, 10), rep(0, 10))
alternating <- rep(c(1, 0), 10)
random_order <- sample(c(rep(1, 10), rep(0, 10)))

examples <- list(
  Clustered = clustered,
  Alternating = alternating,
  Random_order = random_order
)

results <- do.call(rbind, lapply(names(examples), function(name) {
  fit <- runs_test_binary(examples[[name]], method = "exact")
  data.frame(sequence = name, runs = fit$runs, p_value = fit$p_value)
}))
results
```

The clustered and alternating sequences are both nonrandom, but they occupy opposite tails of the runs distribution.

# Continuous observations: dichotomizing at a threshold

For quantitative observations $Y_1,\ldots,Y_n$, a common procedure is to encode

$$
X_i=
\begin{cases}
+, & Y_i>c,\\
-, & Y_i<c,
\end{cases}
$$

where $c$ is a prespecified target or the sample median.

Using the median tends to balance $n_1$ and $n_2$, which improves power. However, dichotomization discards magnitude information: a value barely above the median is treated the same as an extreme value.

```{r}
#| label: numeric-example
set.seed(42)
y <- rnorm(40)
threshold <- median(y)
symbols <- ifelse(y > threshold, "+", "-")
runs_test_binary(symbols, method = "exact")
```

## Ties at the threshold

If some observations equal $c$, decide on a rule before testing. Possibilities include:

- remove observations equal to $c$ and join the remaining sequence;
- assign ties by a prespecified scientific rule;
- use a method designed explicitly for ties.

Do not assign ties after looking for the choice that produces the smallest p-value. If many observations tie, the ordinary two-symbol runs test may be inappropriate.

# Using an R package

The `randtests` package provides `runs.test()`.

```{r}
#| eval: false
# install.packages("randtests")
library(randtests)

x_num <- ifelse(seq_raw == "U", 1, 0)
runs.test(x_num, threshold = 0.5)

# Numeric observations are commonly split at the median.
runs.test(y)
```

When comparing output, check:

- whether an exact or Normal method is used;
- whether a continuity correction is used;
- how ties are handled;
- how the two-sided p-value is defined;
- which alternative corresponds to clustering or alternation.

# Relationship to the two-sample Wald–Wolfowitz test

The phrase “runs test” is also used for a two-sample problem:

1. pool observations from two independent samples;
2. sort the pooled values;
3. label each ordered value by its sample of origin;
4. count runs of sample labels.

Too few runs suggest the two populations differ. This is related to, but conceptually distinct from, testing serial randomness in one observed sequence. In one-sample randomness problems, the original time or spatial order must be preserved.

# What the test detects

The test is sensitive to simple local ordering patterns:

- clustering of similar signs;
- excessive alternation;
- some trends, when values are coded relative to a fixed threshold;
- persistent stretches above or below a center line.

It is attractive because, conditional on $n_1$ and $n_2$, its null distribution is distribution-free.

# Limitations

A nonsignificant result does **not** prove independence or randomness. The runs test can miss:

- dependence at lags greater than one;
- periodic patterns that happen to generate an ordinary number of runs;
- changes in variance with no change in signs;
- nonlinear dependence;
- patterns lost through dichotomization.

It is also possible for a sequence to have a highly unusual ordering but a moderate number of runs.

For richer time-series diagnostics, consider autocorrelation plots, Ljung–Box tests, change-point methods, or application-specific models. Those procedures test different aspects of structure.

# Recommended reporting

A concise report should include:

- the threshold or coding rule;
- $n_1$, $n_2$, and observed $R$;
- whether the test was exact or approximate;
- whether a continuity correction was used;
- the alternative and p-value;
- a contextual conclusion.

Example:

> Coding each day as up or down produced $n_1=11$ up days, $n_2=9$ down days, and $R=10$ runs. A two-sided runs test gave a large p-value, so the data do not provide evidence that the ordering departs from randomness. This does not establish full serial independence.

# Summary

1. A run is a maximal block of identical symbols.
2. Too few runs indicate clustering; too many indicate excessive alternation.
3. Under $H_0$, all orderings with fixed $n_1$ and $n_2$ are equally likely.
4. The exact distribution is combinatorial and distribution-free.
5. For large groups,
   $$
   Z=\frac{R-\mu_R}{\sigma_R}
   \mathrel{\dot\sim}N(0,1).
   $$
6. Preserve the original order, state the coding rule, and handle ties transparently.
7. A runs test examines one kind of sequential structure; it is not a universal test of independence.

# Practice problems

1. Count the runs in
   $$
   A\,A\,B\,A\,B\,B\,B\,A\,A\,B.
   $$
   Identify $n_1$, $n_2$, and $R$.

2. For $n_1=12$ and $n_2=8$, compute $E(R)$ and $\operatorname{SD}(R)$.

3. Explain why a sequence with $R=2$ is evidence for clustering, while a sequence with the largest possible $R$ is evidence for alternation.

4. Construct two different binary sequences with the same $n_1$, $n_2$, and $R$. What does this demonstrate about the information retained by the test statistic?

5. Simulate the null distribution for $n_1=n_2=15$ and compare it with the exact PMF from `runs_pmf(15, 15)`.
